---
title: "Visualization of the model output with observed data in R"
subtitle: "Reader Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky and Karline Soetaert, Utrecht University"
date: "March 2021, update: February 2026"
output:
  pdf_document: default
  html_document: default
abstract: \noindent Here we show how to include observed data in the same graph as the output of 0D and 1D dynamic models. Such visualization can be used, for instance, to estimate values of model parameters that best fit the observed data. We illustrate different possibilities using the ``plot`` function implemented in the ``deSolve`` package but also using functions in the ``ggplot2`` package. Along the way, we will also show how to import the observed data into R.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      message = FALSE, fig.align='center')
```

# Data visualisation with a model output

It is relatively simple to visualise observed data together with model output, provided that:

* The names of data columns are the same as the names of the state variables in the model.

* The first column contains *time* for dynamic models, or the *space* variable for 1D steady-state models.

To plot the data together with the model output, all you need to do is to specify the name of the data object (matrix or data.frame) in the argument ``obs`` when calling the ``plot`` function. You can also modify the appearance of the observed data through the argument ``obspar``.

## Dynamic 0D model

First, we load the necessary package to solve 0D models.

```{r}
require(deSolve)
```

### SIR model

As an example, we consider the simple SIR model that you made in the class:

```{r}
state.SIR   <- c(S = 17500000, I = 1000, R = 0, Deceased = 0)

pars.SIR   <- c(
  b = 0.00000001,    # [1/ind/d], infection parameter 
  g = 0.07,          # [1/d],     recovery rate of infected individuals
  m = 0.007          # [1/d],     mortality rate of infected individuals
)

SIR <- function(t, state, parms) {
  with (as.list(c(state, parms)), {

    Infection  <- b*S*I
    Recovery   <- g*I
    Mortality  <- m*I

    dS          <- -Infection 
    dI          <-  Infection - Recovery - Mortality
    dR          <-  Recovery
    dDeceased   <-  Mortality    # to track number of deceased people

    return (list(c(dS, dI, dR, dDeceased)))
  }) 
}
```

We run the model for two values of the parameter ``b`` and store it in output variables ``out1`` and ``out2``.

```{r}
p1 <- p2 <- pars.SIR
p1["b"]  <- 6e-9
p2["b"]  <- 5e-9
time.seq <- seq(from=0, to=100, by=1) # time sequence, in days

out1 <- ode(y=state.SIR, times=time.seq, func=SIR, parms=p1)
out2 <- ode(y=state.SIR, times=time.seq, func=SIR, parms=p2)
```

The output of the model looks as shown in the following figure:

```{r, fig.width=6.5, fig.height=2.5}
par(mar=c(2.7,3.7,1,0.5)+0.1,mgp=c(1.8,0.7,0))
plot(out1, out2, las=1, col=1:2, lty=1, mfrow=c(1,4))
```

### Load external data

Next, we load the observed data from an external CSV file ``SIPdata.csv``.

```{r}
SIRdata <- read.csv("SIRdata.csv")
head(SIRdata)
```

We see that the file contains the number of people (fictitious) that died due to the viral infection in NL and BE as a function of time (in days). Additionally, it contains uncertainties of these values (quantified as standard errors, columns ``NL.se`` and ``BE.se``). Note that some of the values are not available (``NA``).

### Compare data with model output

To plot these data together with the model output, we first need to *rearrange* it a little. Specifically, we create two separate data-frames for the NL and BE populations, each containing both the best estimates and uncertainties. Note that we also need to rename the ``NL`` and ``BE`` columns to ``Deceased`` in order to make them compatible with the names of (one of) the state variable in the SIR model. We do not need to do this for the ``time`` column, as this is already the correct name.

```{r}
SIRdata.NL <- SIRdata[, c("time", "NL", "NL.se")]
SIRdata.BE <- SIRdata[, c("time", "BE", "BE.se")]
names(SIRdata.NL)[2:3] <- c("Deceased","se")
names(SIRdata.BE)[2:3] <- c("Deceased","se")
head(SIRdata.NL, n=3)
```

Now we are ready to compare the model output and external data. To do this, we simply need to pass the data as the value of the input parameter ``obs``. However, to include *both* datasets, we need pass them as a ``list``:

```{r, fig.width=5, fig.height=4}
par(mar=c(2.7,2.7,1,0.5)+0.1,mgp=c(1.8,0.7,0))
plot(out1, out2, lty=1,
     obs = list(SIRdata.NL, SIRdata.BE))  # include external data
```

To make the graph a little more appealing and informative, we change colors and symbols for the different datasets, add a legend, and change the line widths. We use ``arrows`` as a way to add error-bars based on the values in the ``se`` column. Check the comments in the R-code below to learn how to do this in detail.

```{r, fig.height=4, fig.width=4}
par(mar=c(2.7,2.7,1,0.5)+0.1,mgp=c(1.8,0.7,0))
plot(out1, out2, # include model outputs
     # add x and y labels
     xlab = "time (days)", ylab = "deceased (ind)",
     # add graph title  
     main = "SIR model",
     # adjust y-axis limits
     ylim = c(0,5000),
     # specify line type, line width and colors
     lty=1, lwd=2, col=c("red","green"),
     # include external data as a LIST of different datasets
     obs = list(SIRdata.BE, SIRdata.NL),
     # specify symbols and colors for the external data
     obspar = list(pch = c(16,17),         # symbols 
                   col = c("red","green")) # colors
     )

# add error-bars as arrows without arrow heads
arrows(x0 = SIRdata.BE$time, y0 = SIRdata.BE$Deceased - SIRdata.BE$se,
       x1 = SIRdata.BE$time, y1 = SIRdata.BE$Deceased + SIRdata.BE$se,
       length=0, col="red")
arrows(x0 = SIRdata.NL$time, y0 = SIRdata.NL$Deceased - SIRdata.NL$se,
       x1 = SIRdata.NL$time, y1 = SIRdata.NL$Deceased + SIRdata.NL$se,
       length=0, col="green")

# add a legend
legend("topleft", # placement
       legend = c(paste(p1["b"], "(BE)"), # text
                  paste(p2["b"], "(NL)")),
       col = c("red","green"), # colors
       lty=1, lwd=2, # line type and width
       title="b-value (country):") # legend title
```
### Approach using ggplot2

``ggplot2`` is an R-packages that offers much more flexibility to plot fancy graphs. Here, we demonstrate how a graph similar to the one above can be plotted using this package. Depending on the perspective, it may look more or less laborious. 

First, we bind the two outputs ``out1`` and ``out2`` into one data frame. We add the value of the parameter ``b`` as the last column, which will help us distinguish between the NL and BE populations.

```{r}
out <- rbind(             # bind the rows of out1 and out2
  cbind(out1, b=p1["b"]), # add a column called 'b'
  cbind(out2, b=p2["b"])  # add a column called 'b'
  )
out <- data.frame(out)
head(out, n=3)
```

At this moment, the column ``b`` contains the values of the parameter ``b``. However, when plotting the data, we will want to use this parameter to distinguish between the NL and BE populations. Thus, we set this column as a ``factor``. Additionally, we change the ``levels`` of this parameter such that the first dataset will be considered as first, and the second one as second. This is necessary because the ``ggplot`` function sorts the levels in an ascending order, which is what we do *not* want now.

```{r}
out$b <- factor(out$b)
# this would be the default order of Levels (ascending):
head(out$b)
# but we actually want the Levels to be in the same order as the values
# of the parameter b
out$b <- factor(out$b, levels=c(p1["b"], p2["b"]))
# this is the correct order now
head(out$b)
```

Next, we bind the two outputs ``BE`` and ``NL`` datasets created above into one data frame. We add a column ``Country`` to distinguish between the BE and NL populations.

```{r}
SIRdata <- rbind(
  cbind(SIRdata.BE, Country="BE"),
  cbind(SIRdata.NL, Country="NL")
)
head(SIRdata, n = 3)
tail(SIRdata, n = 3)
```

Now, we are ready to plot everything using the ``ggplot2`` functions. We start with the smallest code, which yields the default appearance. Note the ``+`` at the end of the separate commands. This is how plotting with ``ggplot2`` works; essentially, you are adding more and more objects to a plot.

```{r, fig.height=4, fig.width=5}
library(ggplot2)
ggplot() + 
  # include the model data as lines
  geom_line(data = out,
            aes(x = time, y = Deceased, color = b)) +
  # include the external data as points
  geom_point(data = SIRdata,
             aes(x = time, y = Deceased, color = Country),
             size=2)
```

Next, we add some extra commands to make the graph look more appealing and informative. Check the comments in the code to learn more. If you want to learn even more, best is to search the internet for tutorials and help on the different ``ggplot2`` functions. Be careful, however, that you do NOT fall into a rabbit hole during this process!

```{r, fig.height=4, fig.width=5}
ggplot() + 
  # include the model data as lines
  geom_line(data = out,
    aes(x = time, y = Deceased, color = b)) +
  # include the external data as points
  geom_point(data = SIRdata,
    aes(x = time, y = Deceased, color = Country),
    size = 2) +
  # include error bars based on the se column
  geom_errorbar(data = SIRdata,
    aes(x = time, ymin = Deceased-se, ymax = Deceased+se, 
        color = Country),
    width=0.2) +
  # add labels and a title
  labs(color = "b-values:", x = "time (days)", title = "SIR model") +
  # restyle the graph
  theme_bw() +
  theme(legend.position = "right",
        panel.grid = element_blank()) +
  # set colors manually
  scale_color_manual(values = rep(c("red", "green"),2))
```

To illustrate few more capabilities of ``ggplot2``, we also plot the data as bars with error-bars.

```{r, fig.height=3, fig.width=6}
# see more examples here:
# https://ggplot2.tidyverse.org/reference/position_dodge.html
ggplot(data = SIRdata,
       aes(x = as.factor(time), y=Deceased, fill=Country)) +
  geom_col(position = "dodge") + 
  geom_errorbar(aes(ymin=Deceased-se, ymax=Deceased+se, 
                    color=Country), 
                width=0.2,
                position = position_dodge(width = 0.9)) +
  xlab("time (days)") + 
  theme_bw() + 
  theme(legend.position = "right", 
        panel.grid = element_blank())
```


## Steady-state 1D model

First, we load the necessary packages to solve 1D models.

```{r}
require(marelac)
require(ReacTran)
require(rootSolve)
```

### Oxygen consumption model

As an example, we consider a very simple model that describes first-order consumption of oxygen in the sediment.

```{r}
N      <- 100           # number of boxes
L      <- 0.1           # depth of sediment [m] 
por    <- 0.8           # porosity

Grid   <- setup.grid.1D(dx.1=0.0001, L=L, N=N)
DiffO2 <- diffcoeff()$O2*86400   # molecular diffusion, m2/d
tort   <- 1-log(por^2)           # tortuosity

pars.O2 <- c(
  rO2cons = 1.5,             # rate constant, [/d]
  por     = por,             # porosity, [-]
  O2BW    = 0.3,             # BW concentration of O2, [mol/m3]
  DO2     = DiffO2/tort      # Sediment diffusion coeff, [m2/d]
)

names     <- "O2"             # name of state variable
nspec     <- 1                # number of species
state.ini <- rep(0, times=N)  # initial condition

O2func <- function(t, state, parms){
  with (as.list(parms), {
    
    O2 <- state
    O2.tran <- tran.1D(C=O2, C.up = O2BW, 
                       D = DO2, VF = por, dx = Grid)
    O2cons  <- rO2cons*O2
    
    dO2 <- O2.tran$dC - O2cons
    return(list(dO2))
  })
}
```

We run the model with different values of the parameter ``rO2cons``. 

```{r}
p1 <- p2 <- p3 <- pars.O2
p1["rO2cons"] <- pars.O2["rO2cons"]/2
p2["rO2cons"] <- pars.O2["rO2cons"]/5
p3["rO2cons"] <- pars.O2["rO2cons"]/10

std1 <- steady.1D(y=state.ini, func=O2func, parms=p1, 
      nspec=nspec, dimens=N, names=names, positive=TRUE)
std2 <- steady.1D(y=state.ini, func=O2func, parms=p2, 
      nspec=nspec, dimens=N, names=names, positive=TRUE)
std3 <- steady.1D(y=state.ini, func=O2func, parms=p3, 
      nspec=nspec, dimens=N, names=names, positive=TRUE)
```


### Load external data

We load the observed data from an external CSV file ``O2data.csv``:

```{r}
O2dat.ext <- read.csv("O2data.csv")
head(O2dat.ext, n=3)
```

We must modify the names of the data columns to match the names of the model state variable (which is ``O2``):

```{r}
# see the names of the state variables
colnames(std1$y)
# modify column names to match the state variables
colnames(O2dat.ext) <- c("x", "O2")
head(O2dat.ext, n=3)
```

### Compare data with model output

Finally, we plot the model results together with the external data. We include the external data by setting ``O2dat.ext`` as the value of the input parameter ``obs``. We additionally modify the shape and color of the data points by specifying the value for the input parameter ``obspar``.

```{r, fig.height=3.5, fig.width=4}
par(mar=c(2.7,2.7,1,0.5)+0.1,mgp=c(1.8,0.7,0))
plot(std1, std2, std3, xyswap=TRUE, grid=Grid$x.mid, 
     # include external data
     obs = O2dat.ext,
     # specify symbols and colors for the external data
     obspar = list(pch=16, col="blue"),
     # specify line type and width for the model output
     lty=1, lwd=2, 
     # specify labels
     ylab="depth (m)", xlab="O2 (mM)")

# add a legend
legend("bottomright", col=1:3, lty=1, lwd=2, 
       title="rO2cons [/d]", 
       legend=c(p1["rO2cons"], p2["rO2cons"], p3["rO2cons"]))
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Soetaert Karline (2009).  rootSolve: Nonlinear root finding, equilibrium and steady-state analysis of ordinary differential equations.  R-package version 1.6. https://CRAN.R-project.org/package=rootSolve
  
Soetaert, Karline and Meysman, Filip (2012). Reactive transport in aquatic ecosystems: Rapid model
  prototyping in the open source software R Environmental Modelling & Software, 32, 49-60.

Soetaert, Karline and Thomas Petzoldt (2020). marelac: Tools for Aquatic Sciences. R package version
  2.1.10. https://CRAN.R-project.org/package=marelac
