---
title: "Visualization of dynamic output from a 1D reaction transport model"
subtitle: "Reader Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky and Karline Soetaert, Utrecht University"
abstract: "\\noindent Here we show how to use functions from R-packages ``deSolve``, ``plot3D`` and ``magick`` to visualise 2D data sets in R. Such data sets are generated, for example, in dynamic runs of 1D reaction-transport models, where the model output is a function of space and time. We illustrate the possibilities using data generated by a model describing salinity distribution in an estuary.<br><br>"
date: "Updates: April 2021, April 2026"
output:
  html_document:
    toc: true
  pdf_document: 
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The salinity model

The "toy model" used in this reader describes salinity as a function of distance in an estuary. We assume that water in the estuary is mixed due to tidal dispersion, and that the water flow (advection rate, ``v``) changes periodically over an annual cycle (e.g., due to a seasonal cycle in precipitation upstream).

First, we load the necessary packages and define the model parameters and model function. 

```{r, message = FALSE, warning = FALSE}
require(deSolve)    
require(ReacTran)
require(rootSolve)

Length  <- 100000                                # m
N       <- 500                                   # - number of boxes
Grid    <- setup.grid.1D(L = Length, N = N)

Sal.ini  <- rep(0, times = N)  # initial condition of salinity
SVnames  <- c("Salinity")      # name of the state variable  

# model parameters
pars <- c(                                      
  riverSal = 0,       # river salinity 
  seaSal   = 35,      # seawater salinity
  v.mean   = 400,     # mean advection velocity [m/d]
  v.amp    = 300,     # amplitude of velocity change [m/d]
  Ddisp    = 30e6     # dispersion coefficient [m2/d]
)

# model function
Estuary1D <- function(t, state, parms) {  
 with (as.list(parms),{

    v       <- v.mean + v.amp*cos(2*pi*t/365)  # velocity at time t
    
    TranSal <- tran.1D(C = state, C.up = riverSal, C.down = seaSal,  
                       D = Ddisp, v = v, dx = Grid)

    dSalinity.dt  <- TranSal$dC    # the rates of change = only transport 

    list(dSalinity.dt, 
         v = v,                    # forcing variable
         meanSal = mean(state))    # extra output, e.g., mean salinity
    })
}
```

Now, we calculate the steady-state solution (``std``) and use it as the initial condition to subsequently calculate a dynamic solution over 2 years in steps of 1 day. 

Note that:

  - By default, the steady-state solution is found assuming that $t=0$. That is, the velocity in the model function is $v=v_{mean} + v_{amp}$ when the steady-state solution is calculated using the ``steady.1D`` function. 
  
  - To set the steady state solution as the initial condition in the dynamic model, we use ``y=std$y`` in the argument of the ``ode.1D`` function.
  

```{r}
# model run: steady
std <- steady.1D(y = Sal.ini, func = Estuary1D, parms = pars, 
                 nspec = 1, dimens = N, positive = TRUE)

# model run: dynamic, output every day over 2 years
out <- ode.1D(y = std$y, times = seq(0,2*365,by=1), parms = pars, 
              func = Estuary1D, nspec = 1, dimens = N, names = SVnames)
``` 

# Default graphical display

The package ``deSolve`` contains functions that work with dynamic 1D objects (``subset``, ``image``, ``matplot.1D``, ``plot.1D``) and with 0D objects (``plot``, ``matplot.0D``), etc.  To get help on these functions, type in the console:

```{}
?plot.deSolve
```

By default, the plotting functions will determine a suitable arrangement of the graphs (by setting its argument ``mfrow``). This can be overruled by specifying your own ``mfrow`` when calling the functions. Alternatively, you can set ``mfrow = NULL``, which will not change the mfrow setting.

In the code below, it is first specified that the image should be positioned in an arrangement that will have two rows and two columns (``mfrow=c(2,2)``). Then, when calling the ``plot`` function, the argument (``mfrow=NULL``) tells R not to overrule this arrangement. In addition to a x--y plot, we also plot a histogram of velocities to illustrate how often valocities in a certain velocity interval occur throughout the modeled time domain.

```{r, fig.align='center', fig.height=7, message = FALSE, warning = FALSE}
image(out, which="Salinity", grid=Grid$x.mid, mfrow=c(2,2))
plot(out, which=c("meanSal", "v"), mfrow=NULL)
hist(out, which = "v", mfrow=NULL)
```

## Subsetting and summary

To get a summary of the output produced by ``ode.1D``, we write:

```{r}
summary(out)
```

## Timeseries at specific positions

We use the following steps to visualize how the salinity changes over time at specific locations in the estuary.

* First, we extract the 2D salinity object (time, space) using the function ``subset``. This function returns a matrix with the values arranged according to time (rows) and space (columns); the times are stored as an attribute.

* Then, we use ``matplot`` for plotting the time variation at specific locations.

**Note 1:** In the example code below, we plot the legend in *another* graph. Function ``plot.new`` just creates a new graph without doing any plotting. If you do not like the fact that the two "graphs" are equally large, you may use the function ``layout``, which allows you to specify the size of the graphs. 
In the example, we use ``layout(mat = matrix(nrow = 1, ncol = 4, data = c(1,1,1,2)))``, which makes the first graph 3 times as wide as the graph with the legend.

**Note 2:** The specific locations are defined using **indices** of the vector ``Grid$x.mid``, where ``Grid$x.mid[1]`` and ``Grid$x.mid[500]`` corresponds to the beginning (more accurately: middle of the **first** grid cell) and the end (more accurately: middle of the **last** grid cell) of the estuary, respectively. 

```{r, fig.align='center'}
Salinity <- subset(out, which = "Salinity")
time     <- attributes(Salinity)$time
select   <- c(1,100,200,300,400,500)  # indices, not real locations!

layout(mat = matrix(nrow = 1, ncol = 4, data = c(1,1,1,2)))
matplot(time, Salinity[,select], col=1:6, type ="l", lty=1, 
      las=1, ylab="-", xlab="time (d)", main="Salinity time-series")
plot.new()
legend("center", col=1:6, lty=1, legend=Grid$x.mid[select]/1000, 
  title="Distance (km)")
```

**Note 3:** If you want the define the specific locations yourself, first define a function that finds the corresponding (closest) indices, then use its output in the same way as above.

```{r, fig.align='center'}
# define function for finding indices in vector x where x[ind] most closely matches xi
find_closest <- function(x, xi)
  sapply(xi, function(val) which.min(abs(x - val)))

# define actual locations
distances <- c(0, 25, 50, 75, 100)*1e3

# find the corresponding indices
select2   <- find_closest(Grid$x.mid, distances)

# use the same steps as above, using the indices select2
Salinity <- subset(out, which = "Salinity")
time     <- attributes(Salinity)$time

layout(mat = matrix(nrow = 1, ncol = 4, data = c(1,1,1,2)))
matplot(time, Salinity[,select2], col=1:6, type ="l", lty=1, 
      las=1, ylab="-", xlab="time (d)", main="Salinity time-series")
plot.new()
legend("center", col=1:6, lty=1, legend=Grid$x.mid[select2]/1000, 
  title="Distance (km)")
```

## Spatial gradients at specific time points

We use the following steps to visualise the varation of salinity in space at specific time points.

* First, we extract the 2D salinity object (time, space) using the function ``subset``. This function returns a matrix with the values arranged according to time (rows) and space (columns); the times are stored as an attribute.

* Then, we use ``matplot``.

**Note:** Now we specify the time points as *days* rather than via indices. We use the function ``find_closest`` defined above to find the indices in the vector *time* that correspond to these specific days.

```{r, fig.align='center'}
Salinity <- subset (out, which = "Salinity")
time     <- attributes(Salinity)$time
tselect  <- c(0,100,200,300,365)        # actual days
tindex   <- find_closest(time, tselect) # corresponding (closest) indices

matplot(x=Grid$x.mid/1000, y=t(Salinity[tindex,]), type ="l", 
        col=1:length(tindex), lty=1:length(tindex), lwd=2,
        xlab="distance (km)", ylab="Salinity (-)", main="Salinity profiles")
legend("bottomright", col=1:length(tindex), lty=1:length(tindex), lwd=2,
       legend=time[tindex], title="day")
```

# Richer graphical display

The function ``image2D`` from the R-package ``plot3D`` is more flexible with respect to plotting of images. For instance, it allows you to label the color key (argument ``clab``), add ``contour`` lines, increase the plotting resolution (``resfac``), etc.

For more details, see:
```{}
?contour2D
?image2D
```

## Countour lines

To use the ``contour2D`` function from ``plot3D``, the data have to be extracted first, and the ``x`` and ``y`` variables have to be explicitly passed as input arguments:

```{r, fig.height=3.5, message=FALSE}
require(plot3D)

S <- subset(out, which = "Salinity")
times <- attributes(S)$times
contour2D(z=S, x=times, y=Grid$x.mid/1000, 
          las=1, main="Salinity", clab="(-)", xlab="time (d)", ylab="distance (km)")
```

## Images with countour lines

The ``plot3D`` function ``image2D`` can be simply passed as an alternative method to the ``deSolve`` image function:

```{r, fig.height=3.5, message = FALSE, warning = FALSE}
image(out, which="Salinity", grid=Grid$x.mid/1000, 
      ylab="km", xlab="time (d)", method="image2D", clab="-", las=1, contour=TRUE)
```

It is more flexible to first extract the salinity and then use the function ``image2D`` directly. Note that here we additionally change the colormap (argument ``col``).

```{r, fig.height=3.5}
S     <- subset(out, which = "Salinity")
times <- attributes(S)$times
image2D(x=times, y=Grid$x.mid/1000, z=S, 
        ylab="km", xlab="time (d)", main="Salinity",
        clab="-", las=1, col = terrain.colors(32),
        contour=list(levels = c(10,20), col="darkred", labcex=1.5, lwd=2))
```


# Animated display

Sometimes, visualization of the results of a dynamic model by the approaches described above may not by sufficiently clear. A useful alternative might be to show the relevant output as an animation (animated gif or mp4 video), which can be done using the R-package ``magick``. Thus, this package must be installed first, e.g., by typing ``install.packages("magick")`` in the console. 

In the code below, we display the temporal evolution of the forcing variable (``v``) along with the image and profiles of the state variable ``Salinity``. 

First, we load the R-package ``magick``.

```{r, warning=FALSE, message=FALSE}
library(magick)
```

In the second step, we define a function ``anim.disp`` that will do the necessary job. It implements the following steps:

1. The ``subset`` function is used to extract the forcing and response variables from the model output. 

    The model output will be passed to the function through the input parameter ``dynout`` (i.e., ``dynout = out``, where ``out`` has been calculated above). Similarly, the names of the forcing and response variables will be passed to the function through the input variables ``forcing`` and ``variable``, respectively. In this example, we will use ``forcing = "v"`` and ``variable = "Salinity"``. Make sure that both the forcing and response variables are part of the output generated by the model function.

2. ``time`` is determined from the attributes of the response variable.

3. Output at or until a selected time point is plotted as graphs and images and exported as PNG files in a temporary sub-folder (``tmp``).

    The selected time points will be passed to the function through the input parameter ``tselect``. In this example, we will use ``tselect = seq(0, 2*365, by=10)``.

4. PNG files are loaded back into R, joined and assembled as an animated image.

5. The animated image is then exported into a file (supported formats include ``gif`` or ``mp4``), and the folder with the temporary files is removed.

Note that unless you have specific requirements, there should be no need for you to modify the code of this function. If you do want to adopt it to your specific needs, you are free to do so. The implementation below is just an example that works for this particular situation.

```{r}
anim.disp <- function(
    dynout = NULL,              # output from a dynamic model
    fname  = "video.mp4",       # default file name of the animated output
    width  = 800, height = 480, # size of the PNG file with graphs & images
    fps    = 10,                # frames per second, must be a factor of 100
                                # (e.g., 2, 4, 5, 10, 20)
    forcing, flab, funit, flim, # details about the forcing variable
    variable, vlab, vunit,      # details about the response variable
    s, slab, sunit,             # details about the spatial domain
    tselect, tlab, tunit)       # details about the selected time points
{
  if(!is.null(dynout)){

    fv    <- subset(dynout, which = forcing)   # extract forcing variable
    C     <- subset(dynout, which = variable)  # extract response variable
    time  <- attributes(C)$times
    # sometimes, last row of C may contain NA; if that happens, set the values  
    # in the last row to that in the second last row.
    if(sum(is.na(C[nrow(C),]))>0){ C[nrow(C), ] <- C[nrow(C)-1, ] }
    
    # find indices in the vector "time" where the values are closest to those in "tselect"
    tsel  <- sapply(tselect, function(val) which.min(abs(time - val)))
    
    # only use data in the range of indices tsel
    C     <- C[1:max(tsel),]
    time  <- time[1:max(tsel)]
    C.max <- max(C)
    
    # create a temporary sub-folder where the png files will be stored
    subfolder <- "tmp"
    if(dir.exists(subfolder)) unlink(subfolder, recursive = TRUE)
    if(!dir.exists(subfolder)) dir.create(subfolder)
  
    # export plots as png images for each tsel[i]
    for(i in 1:length(tsel)) {
      
      png(filename = sprintf("%s/plot_%03d.png", subfolder,i),
          width=width, height=height)
      layout(matrix(c(1,2,3,3), 2, 2, byrow = FALSE), widths = c(1,1))
      
      # top-left: plot the forcing variable (fv) up to time=tsel[i]
      indi <- which(time<tsel[i])
      plot(time[indi], fv[indi], 
           xlim = c(0, max(time)), xlab = paste(tlab, tunit), 
           ylim = flim, ylab = paste(flab, funit), 
           main = "forcing variable", type = "l", lwd = 2)
      
      # bottom-left: display image of the response variable up to time=tsel[i]
      C1 <- C
      C1[which(time>tsel[i]),] <- NA   # NA values are not displayed in the image
      image2D(x = time, y = s, z = C1, 
              xlab = paste(tlab,tunit), ylab = paste(slab, sunit), 
              clab = paste(vlab, vunit),
              clim = c(0,C.max), main = "response variable",
              ylim = c(max(s),min(s)), las = 1)
      
      # right: plot profile of the response variable at time==tsel[i]
      plot(y = s, x = C[tsel[i],], 
           xlim = c(0, C.max), ylim = c(max(s),0), 
           ylab = paste(slab, sunit), xlab = paste(vlab, vunit),
           main = sprintf("%s = %3.1f %s", tlab, time[tsel[i]], tunit),
           type = "l", lty = 1, lwd = 2, col = 1)
      
      # save the png file
      dev.off()
    }
  
    # Read the files back into R
    file_names   <- list.files(pattern = "plot_.*\\.png", 
                               path    = subfolder, full.names = TRUE)
    img_list     <- lapply(file_names, image_read)
    
    # Join & animate
    img_joined   <- image_join(img_list)
    img_animated <- image_animate(img_joined, fps = fps)
    
    # export as animated mp4/gif and tidy up
    image_write(img_animated, path = fname)
    if (dir.exists(subfolder)) unlink(subfolder, recursive = TRUE)
    message(sprintf("%s file created.", fname))
  }
}
```

Now, we are ready to generate the animation, using the function ``anim.disp`` defined above. 

Typically, this can take several minutes, depending on the number of output time points and overall resolution of the model output. Because of this, we do **not** want to do this every time we knit the Rmd document or evaluate the R-chunk. Therefore, we define a flag ``generate_animation``, which we set to ``TRUE`` when we are certain that we want to generate the animation. (For this reader, the animation has already been created, so we will set this flag to ``FALSE`` here.) 

We also define the corresponding file name where the animation will be stored (extension ``mp4`` or ``gif`` is possible).

```{r}
generate_animation <- FALSE
anim_outfile       <- "Sal.mp4" # can also be *.gif (animated gif)
```

Next, we evaluate the ``anim.disp`` function with the appropriate values for the input parameters. Note again that this function is only evaluated if the flag is set to ``TRUE``.

In this example, we want to export the ``"Salinity"`` profiles at time points between 0 and 2 years every 10 days, using ``"v"`` as the forcing variable. The video will be exported at 10 frames per second (``fps = 10``). 

Values for the other input parameters (e.g., ``flab`` for label of the forcing variable, ``funit`` for the unit of the forcing variable, etc.) should be clear once you study and understand the code of the function ``anim.disp``.

```{r generating_animation}
if(generate_animation){
  tselect <- seq(0, 2*365, by=10) # choose output time points
  anim.disp(dynout=out, fname = anim_outfile, fps = 10, width = 800, height = 480,
            variable = "Salinity", vlab = "Salinity", vunit = "(-)",
            forcing = "v", flab = "velocity", funit = "(m/s)", flim = c(0,700),
            s = Grid$x.mid/1e3, slab = "distance", sunit = "(km)",
            tselect = tselect, tlab="time", tunit="(d)")
}
```

Once the animation has been created, it can be embedded in the HTML output using the function ``include_graphics``. Note that PDF output does not support embedded animations. The following code can be used to handle this correctly, with the result shown below.

```{r, fig.align='center'}
if (knitr::is_html_output()) {
  if(file.exists(anim_outfile)){
    knitr::include_graphics(anim_outfile)  # this will embed the animation in HTML output
  } else {
    cat(paste(anim_outfile, "does not exist yet."))
  }
} else {
  cat(paste(anim_outfile, "available externally."))
}
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical
  Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Soetaert Karline (2009).  rootSolve: Nonlinear root finding, equilibrium and steady-state analysis of ordinary  differential equations.  R-package version 1.6

Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving
  Differential Equations in R: Package deSolve. Journal of Statistical Software,
  33(9), 1--25. URL http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09
  
Soetaert, Karline and Meysman, Filip, (2012). Reactive transport in aquatic ecosystems: Rapid model prototyping in the open source software R Environmental Modelling & Software, 32, 49-60.
  
Karline Soetaert (2019). plot3D: Plotting Multi-Dimensional Data. R package
  version 1.3. https://CRAN.R-project.org/package=plot3D
  
