---
title: 'Rate expressions in biogeochemistry: detritus mineralization by bacteria'
author: "Karline Soetaert and Lubos Polerecky, Utrecht University"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
date: "March 2021"
output: 
  pdf_document:
    toc: no 
  header-includes: 
    \usepackage{float} 
    \floatplacement{figure}{H} 
---

# Modelling detritus mineralization by bacteria --- Answers

## Mass balances

Based on the conceptual diagram in Figure 1, the mass balance equations are:

$$\frac{dPOC}{dt} = FluxPOC-HydrolysisPOC+BactMortality$$
$$\frac{dHMWC}{dt} = HydrolysisPOC - HydrolysisHMWC$$
$$\frac{dLMWC}{dt} = HydrolysisHMWC - BactUptake$$
$$\frac{dBACT}{dt} = BactUptake-GrowthRespiration-BasalRespiration-BactMortality$$

## Rate expressions

For the rate expressions, it is important to realise that bacteria are doing ALL the work to dissolve POC into smaller (HMWC) and smaller (LMWC) chunks, until they eventually take up the resource. Thus they are the worker for HydrolysisPOC, HydrolysisHMWC and BactUptake: 

$$HydrolysisPOC  = rPOC \times \frac{POC} {(POC + ksPOC)} \times BACT$$
$$HydrolysisHMWC  = rHMWC \times \frac{HMWC} {(HMWC + ksHMWC)} \times BACT$$
$$BactUptake  = rup \times \frac{LMWC} {(LMWC + ksLMWC)} \times BACT$$

Not all LMWC taken up by the bacteria yields new bacterial biomass. Part of this uptake is used to generate the energy required for biomass synthesis, which is done by respiration. This is called "activity respiration", and it is represented by the rate *GrowthRespiration*: 
$$GrowthRespiration = pResp \times BactUptake$$
In addition, there is basal respiration (also called "maintenance respiration"), which provides energy for the basic body functions but does not lead to growth of the bacterial biomass:
$$BasalRespiration = rBas \times BACT$$

The parameter that defines the bacterial mortality has units of $(mol~C~m^{-3})^{-1}~d^{-1}$, while the bacterial mortality should have units of $mol~C~m^{-3}~d^{-1}$. This means that we need to multiply the mortality parameter with a concentration squared to obtain the correct units for the flux. Thus, the rate expression for the mortality is:\footnote{The description of the parameter as a "quadratic mortality ct" also gives this idea away.}
$$BactMortality = rClos \times BACT^2$$

Modelers often use a quadratic mortality for small organisms that grow very fast and whose predator also grows very fast, so that the predator has high concentration if the prey has high concentration and vice versa.

## R-implementation

Here is the R-code for the model:

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
require(deSolve)  # package with solution methods

# state variables, units = mmolC/m3
state <- c(POC = 100, HMWC = 5, LMWC = 0.15, BACT = 5)

# parameters
parms <- c(
  rPOC = 0.75,    # [d-1]
  rHMWC = 0.5,    # [d-1]
  rup = 2,        # [d-1]
  ksPOC = 100,    # [mmolC.m-3]
  ksHMWC = 5,     # [mmolC.m-3]
  ksLMWC = 0.5,   # [mmolC.m-3]
  rbas = 0.1,     # [d-1]
  ploss = 0.5,    # [-]
  rclos = 0.05,   # [(mmolC.m-3)-1.d-1]
  fluxPOC = 0.5   # [mmolC.m-3.d-1]
)

# Model function
DetBact <- function(t, state, params) {
  with (as.list(c(state, params)), {

   # Rate expressions - all in units of [mmolC/m3/day]
    HydrolysisPOC = rPOC * POC / (POC + ksPOC) * BACT
    HydrolysisHMWC = rHMWC * HMWC / (HMWC + ksHMWC) * BACT
    UptakeLMWC = rup * LMWC/(LMWC + ksLMWC) * BACT
    BasalResp = rbas * BACT
    ActivityResp = UptakeLMWC * ploss
    MortBact = rclos * BACT^2

   # Mass balances [molC/m3/day]
    dPOC.dt  <-  fluxPOC - HydrolysisPOC + MortBact
    dHMWC.dt <-  HydrolysisPOC - HydrolysisHMWC 
    dLMWC.dt <-  HydrolysisHMWC - UptakeLMWC
    dBACT.dt <-  UptakeLMWC - BasalResp - ActivityResp - MortBact

    list(c(dPOC.dt, dHMWC.dt, dLMWC.dt, dBACT.dt),
         totalC = POC + HMWC + LMWC + BACT)
  })

}
```

And the model solution:

```{r, fig.height = 6, fig.width = 8}
# output times
outtimes <- seq(from = 0, to = 360, length.out = 100)  # run the model for 360 days

# solve this model, using the ode function from deSolve
out <- ode(y = state, parms = parms, func = DetBact, times = outtimes)  # solution

# visualise output
plot(out, xlab="time (d)", ylab="mmolC/m3")
```

