---
title: "Reaction-transport in porous media: a simple model of phosphorus cycling in marine sediments"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Karline Soetaert and Lubos Polerecky, Utrecht University"
date: "March 2021"
output:
  pdf_document:
    toc: no
  html_document:
    toc: yes
---

```{r child = 'Pdiagenesis_Q.Rmd'}
```

\newpage
```{r child = 'Pdiagenesis_A.Rmd'}
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Soetaert Karline (2009).  rootSolve: Nonlinear root finding, equilibrium and steady-state analysis of ordinary differential equations.  R-package version 1.6. https://CRAN.R-project.org/package=rootSolve
  
Soetaert, Karline and Meysman, Filip (2012). Reactive transport in aquatic ecosystems: Rapid model
  prototyping in the open source software R Environmental Modelling & Software, 32, 49-60.

Soetaert, Karline and Thomas Petzoldt (2020). marelac: Tools for Aquatic Sciences. R package version
  2.1.10. https://CRAN.R-project.org/package=marelac
