---
title: 'Bears and Salmon: a Food Web Model of a Lake ecosystem'
subtitle: "Exercises Accompanying the Course Environmental Modelling"
author: "Karline Soetaert and Dries Bonte, Ghent University"
date: "July 2021"
output: 
  pdf_document:
    default
---

# Answers

The equations to create are relatively simple. 

The only "complication" is the specification of *Migration*; the parameter specifies the total mass in kg C entering the area per day; we need to divide by the area ($m^2$) to arrive at a change in $kg~C~m^{-2}~d^{-1}$. 

Feeding by the bear and by the scavengers are regular ecological interactions (*BearFeeding*, *ScavFeeding*), where the bear or the scavengers are the active compartment. Thus the grazers (BEAR and SCAVENGER) set the maximal rate, while the food (SALMON, CARCASS) determines the rate limiting term, modelled by a Monod function.

The model is carbon-based. Carbon is used both as a *substrate* for the production of new biomass, but also to deliver *energy*, via respiration. 
There are two respiration terms:

* the basal respiration, which maintains the organism's basal metabolism. This is represented as a fixed fraction of the biomass per day. For simplicity, this term is merged here with the other loss rates of the organisms (e.g., due to mortality) (rates *SalmonLoss*, *BearLoss*, *ScavengerLoss*).
* activity respiration, which delivers the energy that is required for the production of new biomass from the food. This is represented as a fixed fraction of the assimilated food ($assimilation = ingestion - faecesproduction$).



$$BearFeeding = rBearFeeding \times  \frac{SALMON} {(SALMON + ksSalmon)} \times BEAR$$
$$Migration = migrate/area$$
$$SalmonLoss = rSalmonLoss \times SALMON$$
$$CarcassProd= pLossToCarcass \times BearFeeding$$
$$BearLoss  = rBearLoss \times BEAR$$
$$BearLossIngest = pBearLossIngest \times (BearFeeding-CarcassProd)$$
$$ScavFeeding = rScavFeeding \times \frac{CARCASS}{ (CARCASS + ksCarcass)} \times SCAVENGER$$
$$ScavLossIngest = pScavLossIngest \times ScavFeeding$$

$$ScavLoss    = rScavLoss \times SCAVENGER$$
$$CarcassDecay       = rCarcassDecay \times CARCASS$$

# Model implementation in R

Here is the R-code specifying the model:

```{r, eval = TRUE, warning=FALSE, message=FALSE}
require(deSolve)  # package with solution methods

# state variables, units = kgC/m2
state <- c(BEAR = 0.01, SALMON = 0, SCAVENGER = 0.005, CARCASS = 0.001)

# parameters
parms <- c(
  area             = 1000000,  # [m2]
  migrate          = 10000,    # [kg C/day] Amount of salmon entering the river system
  rSalmonLoss      = 0.05,     # [/day] Spawning, death, respiration or salmon rate constant 

  rBearFeeding     = 0.02,     # [/day] Bear feeding rate constant
  ksSalmon         = 0.01,     # [kg C/m2] Half saturation coeff for feeding of bear
  rBearLoss        = 0.01/365, # [/day] Death rate constant (hunting and other losses)
  pLossToCarcass   = 0.5,      # [-] Loss fraction of killed salmon to carcass
  pBearLossIngest  = 0.4,      # [-] Part lost after ingestion (activity resp + faeces production)

  rCarcassDecay    = 0.001,    # [/day] Decay rate constant of carcasses

  rScavFeeding     = 0.02,     # [/day] Scavenger feeding rate constant
  ksCarcass        = 0.002,    # [kg C/m2] Half saturation coeff for feeding of scavenger
  pScavLossIngest  = 0.7,      # [-] Part lost after ingestion (activity resp + faeces production)
  rScavLoss        = 1/365     # [/day] Scaverger Loss rate constant (mortality and respiration)
)

# Model function
RiverRun <- function(t, state, params) {
  with (as.list(c(state, params)), {

   # Rate expressions - all in units of [kgC/m2/day]

   # Salmon dynamics  
    Migration        <- migrate/area
    SalmonLoss       <- rSalmonLoss * SALMON

   # Bear dynamics  
    BearFeeding     <- rBearFeeding * SALMON / (SALMON + ksSalmon) * BEAR
    CarcassProd     <- pLossToCarcass * BearFeeding
    BearLossIngest  <- pBearLossIngest * (BearFeeding-CarcassProd)
    BearLoss        <- rBearLoss * BEAR

   # Scavenger dynamics      
    ScavFeeding     <- rScavFeeding * CARCASS/(CARCASS + ksCarcass) * SCAVENGER
    ScavLossIngest  <- pScavLossIngest *ScavFeeding
    ScavLoss        <- rScavLoss * SCAVENGER

   # Carcasses
    CarcassDecay    <- rCarcassDecay * CARCASS

   # Mass balances [gC/m2/day]
    dBEAR      <- BearFeeding - CarcassProd - BearLossIngest - BearLoss
    dSALMON    <- Migration    - BearFeeding - SalmonLoss
    dCARCASS   <- CarcassProd - CarcassDecay - ScavFeeding  
    dSCAVENGER <- ScavFeeding - ScavLossIngest - ScavLoss
    list(c(dBEAR, dSALMON, dSCAVENGER, dCARCASS))
  })

}
```

The model is solved twice with different parameter values, and the output plotted.

```{r, fig.height = 10, fig.width = 10}
# output time
outtimes <- seq(from = 0, to = 60, length.out = 100)  # run for 2 months

# solve this model using the ode function from package deSolve
out <- ode(y = state, parms = parms, func = RiverRun, times = outtimes)

# Change the parameter values: bears are now more efficient to capture salmon
parms2 <- parms
parms2["rBearFeeding"] <- 0.04 # instead of 0.02
out2 <- ode(y = state, parms = parms2, func = RiverRun, times = outtimes)

# visualise it (black = out, red = out2)
plot(out, out2)

legend("topleft", c("rBearFeeding = 0.02", "rBearFeeding = 0.04"), col = 1:2, lty = 1:2)
```

