---
title: "RTM course - boundary conditions"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on formulating boundary conditions for transport models"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

### Video timeline
* 0:00 Introduction
* 0:25 Spatial and temporal domains
* 1:10 Boundary conditions for steady state solutions
* 3:18 Boundary conditions for transient solutions

![](https://www.youtube.com/watch?v=ec3B41FXsjw&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=13)

You can also read the following section(s) of the textbook:

* Chapter 3.4.6. One-Dimensional Diffusion in Porous Media (Sediments)

(You can download each chapter of the textbook via the e-book University library for free!)

## Reaction-transport in an estuary

The following differential equation describes a reaction-transport model for organic carbon (concentration $C$) in an estuary of length $L$:

$$
\frac{\partial C}{\partial t} = \frac{\partial}{\partial x}\left(D\cdot \frac{ \partial C}{\partial x}\right) -\frac{\partial}{\partial x}(v \cdot C)  - r \cdot C
$$

```{r GTE, echo=FALSE}
question("Which is a good boundary condition for $C$ at the upstream boundary ($x=0$)?", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed.")),
 answer(sprintf("An imposed flux value."), correct = TRUE),
 answer(sprintf("An imposed concentration value."), correct = TRUE)
)  
```

```{r GTE2, echo=FALSE}
question("Which is a good boundary condition for $C$ at the downstream boundary ($x=L$)?", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed.")),
 answer(sprintf("An imposed flux value."), correct = TRUE),
 answer(sprintf("An imposed concentration value."), correct = TRUE)
)  
```

## Reaction-transport in a river

The following differential equation describes a reaction-transport model for organic carbon (concentration $C$) in a river of length $L$:

$$
\frac{\partial C}{\partial t} =  -\frac{\partial}{\partial x}(v \cdot C) - r \cdot C
$$

```{r GTE3, echo=FALSE}
question("Which is a good boundary condition for $C$ at the upstream boundary ($x=0$)? '", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed.")),
 answer(sprintf("An imposed flux value."), correct = TRUE),
 answer(sprintf("A zero-gradient boundary condition.")),
 answer(sprintf("An imposed concentration value."), correct = TRUE)
)  
```

```{r GTE4, echo=FALSE}
question("Which is a good boundary condition for $C$ at the downstream boundary ($x=L$)?", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed."), correct = TRUE),
 answer(sprintf("An imposed flux value.")),
 answer(sprintf("An imposed concentration value."))
)  
```

## Early diagenesis

Consider a two-component model in a porous media with porosity $\phi$. The two components are:

* dissolved $SO_4^{2-}$, which is a solute whose porewater concentration is in $mol~S~m^{-3}_L$ ($S$ refers to sulphur, $m^{-3}_L$ refers to the volume of the liquid phase).
* particulate organic matter, POM, which is a solid whose concentration is in $mol~C~m^{-3}_S$ ($C$ refers to carbon, $m^{-3}_S$ refers to the volume of the solid phase). 

The two components are involved in the process called "sulphate reduction", whose stoichiometry is as follows:

$$
POM + 0.5\cdot SO_4^{2-} \rightarrow CO_2 + 0.5\cdot S^{2-} + \cdots
$$
The rate of sulphate reduction, $R_{sr}$ is expressed in $mol~C~m^{-3}_S~d^{-1}$.

```{r earlydia1, echo = FALSE}
question("Which is a good boundary condition for $POM$ at the sediment-water interface ($x=0$)? ", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed.")),
 answer(sprintf("An imposed flux value."), correct = TRUE),
 answer(sprintf("A zero-gradient boundary condition.")),
 answer(sprintf("An imposed concentration value."))
)  
```

```{r earlydia2, echo = FALSE}
question(sprintf("Which is a good boundary condition for $SO_4^{2-}$ at the sediment-water interface ($x=0$)? "), type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed.")),
 answer(sprintf("An imposed flux value.")),
 answer(sprintf("A zero-gradient boundary condition.")),
 answer(sprintf("An imposed concentration value."), correct = TRUE)
)  
```

```{r earlydia3, echo = FALSE}
question(sprintf("Which is a good boundary condition for $POM$ deep down in the sediment ($x\\rightarrow \\infty$)?"), type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed.")),
 answer(sprintf("An imposed flux value.")),
 answer(sprintf("A zero-gradient boundary condition."), correct = TRUE),
 answer(sprintf("An imposed concentration value."))
)  
```

```{r earlydia4, echo = FALSE}
msg <- "The zero-gradient boundary condition is usually chosen for deep concentrations of solutes in diagenetic models. However, sometimes concentrations are imposed, e.g., if there has been a historic build-up of solutes at depth."

question(sprintf("Which is a good boundary condition for $SO_4^{2-}$ deep down in the sediment ($x\\rightarrow \\infty$)?"), type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed.")),
 answer(sprintf("An imposed flux value.")),
 answer(sprintf("A zero-gradient boundary condition."), correct = TRUE),
 answer(sprintf("An imposed concentration value."), correct = TRUE),
 correct=msg,
 incorrect=msg)  

```

## Textbook and Slides

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>