---
title: "Evaporation and precipitation of water"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky, Utrecht University"
date: "January 2022"
abstract: \noindent Consider a water body surrounded by air. If the relative humidity of the air is below 100%, water will evaporate. In contrast, if the relative air humidity exceeds 100%, which occurs, for instance, when air with a given water vapor content cools down, water will precipitate from the gas phase to form a liquid. In both cases, the process will continue until an equilibrium is reached, i.e., until the relative humidity reaches 100%. In this reader, we will show how these processes can be modeled in R.
output: 
  pdf_document:
    toc: no
  html_document:
    toc: yes
    df_print: paged
  header-includes: 
    \usepackage{float} 
    \floatplacement{figure}{ht}
---

```{r child = 'evaporationH2O_Q.Rmd'}
```

\newpage
```{r child = 'evaporationH2O_A.Rmd'}
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for
  Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential Equations in R:
  Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL
  http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09

Carrier W. H. (1918). “The Temperature of Evaporation,” Trans. Am. Soc. Heat. Vent. Eng., vol. 24, p. 25.