---
title: "Evaporation and precipitation of water --- Part I"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky, Utrecht University"
date: "January 2022"
output: pdf_document
abstract: \noindent Consider a water body surrounded by air. If the relative humidity of the air is below 100%, water will evaporate. In contrast, if the relative air humidity exceeds 100%, which occurs, for instance, when air with a given water vapor content cools down, water will precipitate from the gas phase to form a liquid. In both cases, the process will continue until an equilibrium is reached, i.e., until the relative humidity reaches 100%. In this reader, we will show how these processes can be modeled in R.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Theoretical background

Consider a closed container filled with water and air (Figure 1). The volume and amount of water in the liquid phase are $V_{liq}$ ($m^3$) and $N_{liq}$ ($mol$), respectively. The gas phase has a volume $V_{gas}$ and contains a mixture of air (amount $N_{air}$) and water vapor (amount $N_{vap}$). The liquid water can evaporate to form water vapor, and vice versa, water vapor can precipitate to form liquid water. These processes occur across the liquid-gas boundary (area $A_{liq}$). 

```{r}
#![Gas-tight container filled with liquid water and moist air.](liquid_vapor.pdf){width=30%}
```

In the first approximation, precipitation is assumed to be described by first-order kinetics, i.e., the rate is linearly proportional to $N_{vap}$ (more specifically to the concentration of water vapor molecules, $N_{vap}/V_{gas}$), whereas evaporation is assumed to follow zero-order kinetics, i.e., the rate is independent of $N_{vap}$ (as long as the liquid water is present). Additionally, because the exchange of water molecules occurs at the interface between the liquid and gas phase, both rates are assumed to be proportional to $A_{liq}$. Thus, the change over time in the number of water vapor molecules above the liquid is described by the following differential equation:
\begin{equation}
\label{eq1}
\frac{dN_{vap}}{dt} = k_f\cdot A_{liq} 
  - k_b\cdot A_{liq} \cdot \frac{N_{vap}}{V_{gas}},
\end{equation}
where the rate constants $k_f$ (in $mol~m^{-2}~s^{-1}$) and $k_b$ (in $m~s^{-1}$) correspond to the forward (evaporation) and backward (precipitation) reaction, respectively. Considering that the system is closed, the total amount of water molecules must be conserved. Thus, the amount of molecules of liquid water changes over time at the same rate as the amount of vapor molecules but with the opposite sign:

\begin{equation}
\label{eq2}
\frac{dN_{liq}}{dt} = - \frac{dN_{vap}}{dt}.
\end{equation}

When the time derivative $dN_{vap}/dt$ is equal to zero, the system reaches an equilibrium. Equation 1 implies that the corresponding equilibrium concentration of vapor molecules is equal to the ratio between the rate constants for the forward and backward reaction:\footnote{In this document, values of variables in a thermodynamic equilibrium are marked with an asterisk ($*$).} 
\begin{equation}
\label{eq3}
\frac{dN_{liq}}{dt} = 0 \quad\rightarrow\quad
\left( \frac{N_{vap}}{V_{gas}}\right)^* = \frac{k_f}{k_b}.
\end{equation}
Using this equilibrium concentration, Eq. 1 can be rewritten in the following form:
\begin{equation}
\label{eq4}
\frac{dN_{liq}}{dt} = - \frac{dN_{vap}}{dt} = k_b\cdot A_{liq} \cdot
\left[ \left(\frac{N_{vap}}{V_{gas}}\right)^* - 
\left(\frac{N_{vap}}{V_{gas}}\right) \right].
\end{equation}
Thus, the process of evaporation and precipitation can be thought of as being driven by a chemical disequilibrium, i.e., the net rate of exchange of water molecules between the liquid and gas phase is proportional to the difference between the equilibrium and instantaneous concentration of vapor molecules above the liquid.

## Reformulation of the equations using vapor pressure

If water vapor behaves as an ideal gas, the concentration of water vapor molecules is related to the vapor pressure, $P_{vap}$, through the ideal gas law,
\begin{equation}
\label{eq5}
\frac{N_{vap}}{V_{gas}} = \frac{P_{vap}}{R\cdot T},
\end{equation}
where $R=8.314~J~K^{-1}~mol^{-1}$ is the gas constant, and $T$ is the vapor temperature (in $K$). Using this relationship, Eq. 4 can be rewritten in the form
\begin{equation}
\label{eq6}
\frac{dN_{vap}}{dt} = \frac{k_b}{R\cdot T} \cdot A_{liq}\cdot \left( P_{vap}^* - P_{vap} \right),
\end{equation}
where $P_{vap}^*$ denotes the equilibrium vapor pressure.\footnote{$P_{vap}^*$ is related to the equilibrium concentration $(N_{vap}/V_{gas})^*$ through the ideal gas law (Eq. 5).} This form is equivalent to the empirical relationship developed by Carrier\footnote{Carrier W. H., “The Temperature of Evaporation,” Trans. Am. Soc. Heat. Vent. Eng., vol. 24, p. 25, 1918.} to describe water evaporation and precipitation: 
\begin{equation}
\label{eq7}
\frac{dN_{vap}}{dt} = \frac{c_1 + c_2\cdot v}{\Delta\bar{H}_{vap}} \cdot A_{liq}\cdot \left( P_{vap}^* - P_{vap} \right),
\end{equation}
where $\Delta \bar{H}_{vap}$ is the molar enthalpy of vaporization of water ($\approx 40.66~kJ~mol^{-1}$), $v$ is the wind velocity (in $m~s^{-1}$), and $c_1 = 0.0888~m~s^{-1}$ and $c_2=0.0783$ are empirical constants. 

By comparing Eqs. 6 and 7, we see that the precipitation rate constant $k_b$ in Eq. 1 can be calculated from the empirical constants $c_1$ and $c_2$ according to
\begin{equation}
\label{eq8}
k_b = \frac{R\cdot T}{\Delta \bar{H}_{vap}}\cdot (c_1+c_2\cdot v).
\end{equation}
Additionally, by combining this result with Eqs. 3 and 5, the evaporation rate constant $k_f$ in Eq. 1 can be calculated from $k_b$, temperature, and the equilibrium vapor pressure according to 
\begin{equation}
\label{eq9}
k_f = k_b \cdot \frac{P_{vap}^*}{R\cdot T}.
\end{equation}

## Equilibrium vapor pressure

The equilibrium vapor pressure depends on the temperature of the system comprising liquid water and water vapor and can be calculated if the temperature is known. For example, we know that water boils at a temperature of $100~^\circ C$ when the external pressure is $1~atm$. Thus, the equilibrium vapor pressure at a temperature of $T_{boil} = 373.15~K$ is $P_{vap}^*(T_{boil}) = 1~atm = 101325~Pa$. At any other temperature, $T$, the equilibrium vapor pressure is calculated using the Clausius-Clapeyron equation:
\begin{equation}
\label{eq10}
\ln\left[\frac{P_{vap}^*(T)}{P_{vap}^*(T_{boil})}\right] =
-\frac{\Delta \bar{H}_{vap}}{R}\cdot \left(\frac{1}{T} - \frac{1}{T_{boil}}\right).
\end{equation}
This result follows from equilibrium thermodynamics and assumes that the molar enthalpy of vaporisation of water, $\Delta \bar{H}_{vap}$, is temperature-independent.

## Molar fraction and partial pressure

When dealing with mixtures, properties of individual components in the mixture are described by *partial molar* quantities. One of the most basic partial molar quantities is *molar fraction*, typically denoted by $x$. For a component $A$, this quantity is defined as the ratio between the amount of component $A$ and the total amount of the mixture (both amounts are in moles):
\begin{equation}
\label{eq11}
x_A = N_A / N_{tot}\quad (mol~mol^{-1}).
\end{equation}

When dealing with gas mixtures, such as atmospheric air, individual components are typically characterized by another quantity: partial pressure. If a mixture behaves as an ideal gas, partial pressures are additive. Thus, partial pressure of an individual component is equal to the product of the total pressure, $P_{tot}$, and the molar fraction of that component: $P_A=x_A\cdot P_{tot}$. Because moist air at typical conditions behaves as an ideal gas, partial pressure of water vapor is calculated as
\begin{equation}
\label{eq12}
P_{vap} = x_{vap}\cdot P_{air},
\end{equation}
where $x_{vap}$ is the molar fraction of water vapor in air and $P_{air}$ is the total air pressure. For example, if the total air pressure is $1~atm$ and 1 in every 1000 molecules of air is water, the molar fraction of water vapor is $x_{vap}=0.001$ and the corresponding partial pressure of water vapor is $P_{vap}= 1~matm$.

## Relative air humidity

Rather than using a molar fraction, the water content of air is very often expressed in terms of a *relative humidity*, $\phi$. This quantity is defined as the ratio between the vapor pressure and the equilibrium vapor pressure. Thus, if the relative humidity of air is known, the vapor pressure is calculated according to
\begin{equation}
\label{eq13}
P_{vap} = \phi \cdot P_{vap}^*.
\end{equation}

Because the molar fraction of water vapor in air and the relative humidity of air are relative quantities, they are often expressed in %. This can cause confusion. We emphasize here that these two quantities are fundamentally different: while the molar fraction of water vapor in air is the ratio between the vapor pressure and the *total* air pressure (Eq. 12), the relative humidity of air is the ratio between the vapor pressure and the *equilibrium* vapor pressure (Eq. 13).

# Task 1

Calculate the equilibrium vapor pressure, $P_{vap}^*$, for temperatures of $10~^\circ C$, $20~^\circ C$, and $50~^\circ C$. What are the corresponding vapor pressures, $P_{vap}$, and molar fractions, $x_{vap}$, if the total air pressure is $0.95~atm$ and the relative air humidity is 80 %?

## Solution 1

First, we implement the Clausius-Clapeyron equation (Eq. 10) to calculate the equilibrium vapor pressure, then use Eq. 13 to calculate the vapor pressure, and finally use Eq. 12 to calculate the molar fraction.
```{r, echo=TRUE, eval=TRUE}
# implement Clausius-Clapeyron equation
Peq_vapor <- function(TK=300)   # input temperature in K!
  101325 * exp( -40.66e3/8.314 * (1/TK - 1/373.15) )

# conditions
Pair  <- 0.95*101325            # air pressure, in Pa
humid <- 0.8                    # relative air humidity
TK    <- c(10, 20, 50) + 273.15 # temperature, in K
Peq   <- Peq_vapor(TK)          # equilibrium vapor pressure, in Pa
Pvap  <- humid*Peq              # vapor pressure, in Pa
xvap  <- Pvap/Pair              # molar fraction of water vapor

# display results as a data.frame
data.frame(T_K = TK, Peq_Pascal = Peq, xvap_percent = xvap*100)
```

The above results show that although the relative air humidity is constant, the molar fraction of water vapor steeply increases with temperature from about 1 % to about 11 %. This increase is caused by the steep increase in the equilibrium vapor pressure with temperature.


# Task 2

Consider a closed container filled with liquid water and moist air (Figure 1). Initially, the volume of the liquid water is $V_{liq} = 1~L$, the total volume of the container ($V_{liq}+V_{gas}$) is $1~m^3$, and the relative air humidity is 50 %. The system is kept at a constant temperature of $20~^\circ C$ for $0.5~hr$. Then, the system is cooled and kept at a constant temperature of $10~^\circ C$ for another $0.5~hr$. Through the entire process the system is kept at a constant pressure of $1~atm$.

* Write a model that calculates the amount of water vapor above the liquid water as a function of time. Express this amount in moles, as a molar fraction, and as a relative air humidity. 

* Evaluate the gross and net amounts of water transferred between the liquid and gas phase during the first and second half-hour intervals.

## Assumptions

* Area of the liquid water surface is constant ($A_{liq} = 1~m^2$). 

* Air is stagnant. Thus, you can assume the wind velocity of $v=0$ (see Eq. 7 and 8).

* Dissolution in water of air components such as $N_2$ or $O_2$ can be ignored.

* The air and water vapor behave as an ideal gas. Since the total pressure of the gas phase and the temperature of the entire system are kept constant, the evaporation and precipitation processes as well as the isobaric cooling phase will lead to a volume change for the liquid water, the gas phase (air + water vapor), as well as the total volume (liquid + gas phase). Except for the change in volume of the liquid water during cooling, these changes cannot be neglected. You can assume that the walls of the container are flexible to allow for such changes to occur without the need to perform an extra work.

* To evaporate $N$ moles of water at a constant pressure and temperature, a certain amount of heat is required ($q=N\cdot \Delta\bar{H}_{vap}$). You can assume that this heat is supplied externally such that the pressure and temperature of the system *can* remain constant during the modeled time interval. Similarly, you can assume that the heat produced during precipitation is transferred to the surrounding. The cooling from $20~^\circ C$ to $10~^\circ C$ is also done by transferring heat to the surrounding.

* The liquid and gas phases are well mixed (i.e., homogeneous). Thus, you can ignore transport processes and base your model entirely on mass balances (Eqs. 1 and 2) and the ideal gas law (Eq. 5).

# Solution 2

## Model implementation in R

First, we define model parameters and empirical constants.

```{r}
# model parameters
Aliq     <- 1           # liquid water area (m2) 
Ptot     <- 101325      # total pressure (Pa)
v        <- 0           # wind speed (m h-1)
TK1      <- 293.15      # temperature during first period (K)
TK2      <- 283.15      # temperature during second period (K)

# empirical constants or parameters
Rgas     <- 8.314       # gas constant (J mol-1 K-1)
dHvap    <- 40.66e3     # molar enthalpy of vaporisation of water (J mol-1)
c1       <- 0.0888*3600 # empirical constant in Eq. 7 (m h-1)
c2       <- 0.0783      # empirical constant in Eq. 7 (-)
rho      <- 1e3         # water density (kg m-3)
MW       <- 18e-3       # molar weight of water (kg mol-1)
```

Then, we define auxillary functions for calculating the amounts of water molecules from the volume of the water vapor and liquid, respectively, and the rate constants from the temperature. We use the ``Peq_vapor()`` function defined above.

```{r}
Nwater_vapor <- function(humid, TK, V){ # calculate Nvap from volume, T, and humidity
  humid*Peq_vapor(TK)*V/(TK*Rgas)       # based on the ideal gas law (Eq. 5) and Eq. 13
}

Nwater_liquid <- function(V){           # calculate Nliq from volume
  rho*V/MW                              # using density and molar weight
}

kbT <- function(TK, v=0){               # rate constant of water precipitation (m h-1)
  Rgas*TK/dHvap*(c1+c2*v)               # Eq. 8
}

kfT <- function(TK, v=0){               # rate constant of water evaporation (mol m-2 h-1)
  kbT(TK, v) * Peq_vapor(TK)/(Rgas*TK)  # Eq. 9
}
```

Finally, we define the model function based on the differential equations 1 and 2. 

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
WaterEvapPrecip <-function(t, state, pars) {
  with (as.list(c(state, pars)),{
    
    # rate constants at the given temperature (TK given in pars)
    kb <- kbT(TK)
    kf <- kfT(TK)
    
    # volume of the gas phase (air+vapor) at a given Nvap, Nair, TK, and Ptot
    # (Nair given in pars)
    Vgas <- Rgas*TK*(Nair+Nvap)/Ptot # based on the ideal gas law
    
    # process rates, mol h-1, Eq. 1
    evaporation   <- kf*Aliq
    precipitation <- kb*Aliq * Nvap/Vgas
    
    # time-derivatives
    dNvap.dt <-  evaporation - precipitation
    dNliq.dt <- -evaporation + precipitation
    
    # extra output quantities 
    Pvap  <- Nvap/Vgas * Rgas * TK # partial pressure of vapor, Eq. 5
    Pair  <- Nair/Vgas * Rgas * TK # partial pressure of air
    Pgas  <- Pvap + Pair           # total pressure of the gas phase
    humid <- Pvap/Peq_vapor(TK)    # relative air humidity, Eq. 13
    Vliq  <- Nliq*MW/rho           # volume of the liquid water
    Vtot  <- Vgas + Vliq           # total volume (liquid + gas)
    
    # return time-derivatives and ordinary variables as a list
    list(c(dNvap.dt, dNliq.dt),
         
         # extra output:
         # amounts of molecules
         Nwater = Nvap + Nliq,                  # water, total
         Nair   = Ptot * Vgas/(Rgas*TK) - Nvap, # air, total (to check!)
         
         # process rates, mol h-1
         R.evaporation   = evaporation,
         R.precipitation = precipitation,
         
         # rate constants
         kb = kb,             # m h-1
         kf = kf,             # mol m-2 h-1
         
         # volumes, m3
         Vgas = Vgas,         # gas phase
         Vliq = Vliq,         # liquid phase
         Vtot = Vtot,         # total (gas+liquid)
         
         # pressures, Pa
         Pvap = Pvap,  # partial pressure of vapor
         Pair = Pair,  # partial pressure of air
         Pgas = Pgas,  # total gas pressure (vapor+air)
         
         # fractions
         xvap = Nvap/(Nvap+Nair), # molar fraction of water vapor in air, Eq. 11
         #xvap = Pvap/Pext,       # molar fraction of water vapor in air, Eq. 12
         humid = humid            # relative air humidity, (-)
         )
  })
}

```

## Model application

### Initial conditions and run 1

First, we define a function that returns the values of state variables and model parameters based on the conditions defined by the problem (i.e., initial total volume, initial volume of the liquid water, initial relative air humidity, and initial temperature).

```{r}
find_state_pars <- function(Vtot, Vliq, humid, TK){
  # volume of the gas phase = Vtotal - Vliquid
  Vgas <- Vtot - Vliq

  # amounts of vapor, liquid, and air in the gas phase
  Nvap <- Nwater_vapor(humid=humid, TK=TK, V=Vgas)
  Nliq <- Nwater_liquid(V=Vliq)
  Nair <- Ptot * Vgas/(Rgas*TK) - Nvap
  
  # output: state variables and parameters
  return(list(SV   = c(Nvap = Nvap, Nliq = Nliq), 
              pars = c(TK = TK, Nair = Nair)))
}
```

Now, we use the above function to generate the initial conditions and model parameters and solve the model over the first half-hour.

```{r, message=FALSE}
# initial conditions
ini   <- find_state_pars(Vtot=1, Vliq=1e-3, humid=0.5, TK=TK1)

# solve the model over the first half-hour
outtimes <- seq(from = 0, to = 0.5, length.out = 1000) # time in hr
require(deSolve)
out1 <- ode(y = ini$SV, parms = ini$pars, func = WaterEvapPrecip, times = outtimes)
```

### New initial conditions and run 2

In the second step, we use the end values of the first model run to calculate the initial conditions for the second model run. The gas was cooled from $T_1$ to $T_2$ at a constant pressure. Thus, the pressures (including the total pressure as well as the partial pressures of the components in the gas phase) do not change. However, the volume of the gas phases decreases by a factor $T_2/T_1$, as follows from the ideal gas law. Also, the equilibrium pressure decreases and, subsequently, the relative humidity increases as a result of the cooling. The change in the volume of the liquid water due to cooling is neglected (see Assumptions). 

```{r}
# calculate new initial conditions
SV   <- out1[nrow(out1),]                    # end of the previous run
ini2 <- find_state_pars(
           Vtot  = SV[["Vgas"]]*TK2/TK1 +    # isobaric cooling, gas volume changes
                   SV[["Vliq"]],             # change in liquid volume neglected
           Vliq  = SV[["Vliq"]], 
           humid = SV[["Pvap"]]/Peq_vapor(TK=TK2), 
           TK    = TK2)

# solve the model
out2 <- ode(y = ini2$SV, parms = ini2$pars, func = WaterEvapPrecip, times = outtimes)
```

## Results & Discussion

The first set of graphs (see below) shows that, due to net evaporation at the higher (initial) temperature, the relative air humidity increases from the initial 50 % to 100 % **within about 15 minutes**. This is associated with an increase in the molar fraction (from 0.014 to 0.028), vapor pressure (from 1400 Pa to 2800 Pa) and a transfer of about 0.6 moles of water from the liquid to the gas phase. 

The cooling of the 100 % humid air increased the relative air humidity to about 180 %, leading to net precipitation. The new equilibrium was reached, again, in about 15 minutes, but the total changes in $x_{vap}$, $P_{vap}$, and $N_{vap}$ were lower.

```{r, fig.height=4.7, fig.width=8}
plot(out1, out2,
     which=c("humid", "xvap", "Pvap", "Vliq", 
             "Nvap", "Nliq", "Nair", "Nwater"),
     xlab="time (hr)", lty=1, las=1, lwd=2, mfrow=c(2,4),
     ylab=c("-", "-", "Pa", "m3", rep("mol H2O",3)))
legend("bottomright", title="temperature (K):",
       legend=c(sprintf("%.1f",ini$pars["TK"]), 
                sprintf("%.1f",ini2$pars["TK"])), 
       lty=1, lwd=2, col=1:2, bty="n")
```

The second set of graphs (see below) shows that evaporation occurs at a rate of about $22~mol~h^{-1}$ and $12~mol~h^{-1}$ at the higher and lower temperature, respectively. During the first half-hour, precipitation increases from the initial value of $11~mol~h^{-1}$ and becomes equal to the evaporation rate after about 15 minutes. During the second half-hour, precipitation decreases and becomes equal to the evaporation rate after about 15 minutes. Rate constants remain constant during each half-hour interval because the temperature is constant, but the values differ for the different temperatures ($c_1$ and $c_2$ are assumed to be temperature-independent).

```{r, fig.height=2.5, fig.width=8}
plot(out1, out2,
     which=c("R.evaporation", "R.precipitation", "kf","kb"),
     xlab="time (hr)", lty=1, las=1, lwd=2, mfrow=c(1,4),
     ylab=c("mol h-1", "mol h-1", "mol m-2 h-1", "m h-1"))
legend("right", title="temperature (K):",
       legend=c(sprintf("%.1f",ini$pars["TK"]), 
                sprintf("%.1f",ini2$pars["TK"])), 
       lty=1, lwd=2, col=1:2, bty="n")
```

The third set of graphs (see below)  shows that, due to net evaporation at the higher (initial) temperature, the volume of the gas phase (air + water vapor) increased by about $14.4~L$ and the volume of the liquid water decreased by about $11~mL$. Thus, the overall volume of the system increased by about $14.4~L$. Because the total pressure was constant, these changes were associated with an increase in the vapor pressure but a decrease in the air (i.e., air components other than the water vapor) pressure. The trends were opposite at the lower temperature during the second half-hour. The isobaric cooling in between had a more dramatic effect on the gas and total volumes than then evaporation/precipitation.

```{r, fig.height=5, fig.width=8}
plot(out1, out2,
     which=c("Vgas", "Vliq", "Vtot", "Pvap", "Pair", "Pgas"),
     xlab="time (hr)", lty=1, las=1, lwd=2, mfrow=c(2,3),
     ylab=c(rep("m3",3), rep("Pa",3)))
```

### Water budget

The evaporation rate of $22~mol~h^{-1}$ implies that about $11~mol$ of water evaporated during the initial half-hour, which is considerably larger than the net increase of $0.6~mol$ in the number of water molecules transferred from the liquid phase to the gas phase. 

```{r, message=FALSE}
require(oce)
evap <- integrateTrapezoid(outtimes, out1[,"R.evaporation"])
prec <- integrateTrapezoid(outtimes, out1[,"R.precipitation"])
dN   <- evap - prec
dN2  <- diff(out1[c(1,nrow(out1)),"Nvap"])

# water budget in moles
(budget_mol <- c(evap=evap, prec=prec, dN=dN, dN2=dN2))

# water budget in liters
(budget_liter <- budget_mol*MW/rho * 1e3)
```