---
title: "${}^{18}$O exchange between two water reservoirs"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky, Utrecht University"
date: "January 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem definition

Consider an experimental setup where an open container is enclosed in a gas-tight container filled with air. Both containers are partially filled with water (Figure 1). Initially, water in the reservoir 1 is isotopically enriched in ${}^{18}O_2$ (e.g., by the addition of ${}^{18}O_2$-labeled $H_2O$), water in the reservoir 2 has a natural abundance of ${}^{18}O_2$, and there is no water vapor above the liquid. Due to evaporation and precipitation, there will be an isotope exchange between the two liquid water reservoirs. Your task is to model the dynamics of this exchange. 

![Experimental setup with two liquid water reservoirs enclosed in a gas-tight container.](water_reservoirs.pdf){width=50%}

## Assumptions

* Temperature of the system is constant.
* In the first container, the water height is $h_1$ and the area of the air-water interface is $A_1$. 
* In the second container, the water height is $h_2$ and the area of the air-water interface is $A_2$. 
* The volume of the gas-tight container is $V_3$ and the air pressure is $P_3$.
* Water molecules can be transferred between the liquid and gas phase by evaporation and precipitation. The rate of these processes is governed by equations described below.
* Changes in $V_3$ associated with changes in $h_1$ and $h_2$ due to evaporation and precipitation can be neglected.
* All reservoirs are well mixed, so water transport within the reservoirs does not need to be considered.
* Water vapor behaves as an ideal gas, i.e., its volume, partial pressure, and temperature are related through the ideal gas law ($P\cdot V = n\cdot R\cdot T$).

# Theoretical background

## Evaporation and precipitation of water

In the first instance, we consider a simplified system consisting of two water reservoirs enclosed in a gas-tight container: liquid water (volume $V_{liq}$, amount $N_{liq}$), and water vapor (volume $V_{vap}$, amount $N_{vap}$). Water in the liquid phase can evaporate to form water vapor, and vice versa, water vapor can precipitate to form liquid water. In the first approximation, precipitation is assumed to be described by first-order kinetics, i.e., the rate is linearly proportional to $N_{vap}$ (more specifically to the concentration of water vapor molecules $N_{vap}/V_{vap}$), whereas evaporation is assumed to follow zero-order kinetics, i.e., the rate is independent of $N_{vap}$ (as long as the liquid water is present). Additionally, since both processes occur at the interface between the liquid and gas phase, their rates are assumed to be proportional to the area of this interface, $A_{liq}$. Thus, the change over time in the number of water vapor molecules above the liquid is described by the following differential equation:
\begin{equation}
\label{eq1}
\frac{dN_{vap}}{dt} = k_f\cdot A_{liq} 
  - k_b\cdot A_{liq} \cdot \frac{N_{vap}}{V_{vap}},
\end{equation}
where the rate constants $k_f$ (in $mol~m^{-2}~s^{-1}$) and $k_b$ (in $m~s^{-1}$) correspond to the forward (evaporation) and backward (precipitation) reaction, respectively. In a thermodynamic equilibrium, the time derivative $dN_{vap}/dt$ is equal to zero, which implies that the equilibrium concentration of water vapor molecules is equal to the ratio between the rate constants for the forward and backward reaction:\footnote{In this document we use an asterisk ($*$) to denote a thermodynamic equilibrium.} 
$$
\left( \frac{N_{vap}}{V_{vap}}\right)^* = \frac{k_f}{k_b}.
$$

Assuming that water vapor behaves as an ideal gas, we have 
$$
P_{vap}\cdot V_{vap} = N_{vap}\cdot R\cdot T,
$$
where $P_{vap}$ is the vapor pressure (in $Pa$), $R$ is the gas constant ($8.314~J~K^{-1}~mol^{-1}$), and $T$ is the temperature (in $K$). Consequently, the concentration of vapor molecules, $N_{vap}/V_{vap}$, is proportional to the vapor pressure, $P_{vap}$, and Eq. \ref{eq1} can be rewritten in the form
\begin{equation}
\label{eq2}
\frac{dN_{vap}}{dt} = k_b'\cdot A_{liq} \cdot \left( P_{vap}^* - P_{vap} \right).
\end{equation}
In this equation, $k_b' = k_b/(R\cdot T)$ is a renormalised precipitation rate constant and $P_{vap}^*$ denotes to the vapor pressure in a thermodynamic equilibrium. A direct comparison of Eqs. 1 and 2 yields that $P_{vap}^*$ is proportional to the equilibrium concentration of the water vapor molecules and thus to the ratio between the rate constants for the forward and backward reaction:

\begin{equation}
\label{eq3}
P_{vap}^* = R\cdot T\cdot \left( \frac{N_{vap}}{V_{vap}}\right)^*
 = R\cdot T\cdot \frac{k_f}{k_b}.
\end{equation}

Equation 2 implies that, in the first approximation, the dynamic of the partitioning of water between the liquid and gas phase is driven by a thermodynamic disequilibrium: the net rate of change in the number of water molecules in the gas phase is proportional to the difference between the equilibrium and instantaneous vapor pressure above the liquid, $P_{vap}^* - P_{vap}$. This is consistent with our intuition: there is net evaporation when $P_{vap}$ is below $P_{vap}^*$, net precipitation when $P_{vap}$ exceeds $P_{vap}^*$, and no net exchange when $P_{vap}$ and $P_{vap}^*$ are equal. 

The equilibrium vapor pressure depends on temperature according to the Clausius-Clapeyron equation:\footnote{See notes on Physical Chemistry.}
\begin{equation}
\label{eq4}
\ln\left[\frac{P_{vap}^*(T)}{P_{vap}^*(T_0)}\right] =
-\frac{\Delta H_{vap}}{R}\cdot \left(\frac{1}{T} - \frac{1}{T_0}\right),
\end{equation}
where $\Delta H_{vap}$ is the enthalpy of vaporisation of water ($\approx 40.66~kJ~mol^{-1}$), $P_{vap}^*(T_0)=1~bar \approx 101325~Pa$, and $T_0 = 373.15~K$ (all values correspond to the standard boiling point of water). 

When considering air, which is a mixture of gases such as $N_2$, $O_2$, and water vapor, the partial pressure of water vapor is typically expressed as a *fraction* of the equilibrium pressure,
\begin{equation}
\label{eq5}
P_{vap}(T) = h\cdot P_{vap}^*(T),
\end{equation}
where the fraction, $h$, is called the relative humidity. 

## Isotope fractionation during evaporation and precipitation

Now, we consider isotope effects on the kinetics of water evaporation and precipitation and rewrite the mass balance Eq. 1 separately for $^{18}O$ and $^{16}O$ water molecules. We assume that

* the rate of evaporation of $^{18}O$-water is proportional to the probability that the water molecule in the liquid phase contains $^{18}O$. This probability is equal to the ${}^{18}O$ atom fraction of the liquid water, which is denoted as $x_{liq}$ and calculated as $$x_{liq} = \frac{{}^{18}N_{liq}}{{}^{18}N_{liq} + {}^{16}N_{liq}}.$$

* the rate of precipitation of $^{18}O$-water is proportional to the probability that the water molecule in the gas phase contains $^{18}O$. This probability is equal to the ${}^{18}O$ atom fraction of the water vapor, which is denoted as $x_{vap}$ and calculated as $$x_{vap} = \frac{{}^{18}N_{vap} }{{}^{18}N_{vap} + {}^{16}N_{vap}}.$$

* the rate of evaporation of $^{16}O$-water is proportional to the probability that the water molecule in the liquid phase contains $^{16}O$. This probability is equal to $1-x_{liq}$.

* the rate of precipitation of $^{16}O$-water is proportional to the probability that the water molecule in the gas phase contains $^{16}O$. This probability is equal to $1-x_{vap}$.

* the rate constants of the forward and backward reactions are different.

Based on these assumptions, the exchange of $^{18}O$ and $^{16}O$ water molecules between the liquid water and water vapor is described by the following differential equations:

\begin{equation}
\label{eq6}
\frac{{}^{18}N_{vap}}{dt} = {}^{18}k_f \cdot x_{liq} \cdot A_{liq} - 
{}^{18}k_b\cdot x_{vap} \cdot A_{liq} \cdot \frac{N_{vap}}{V_{vap}}, 
\end{equation}

\begin{equation}
\label{eq7}
\frac{{}^{16}N_{vap}}{dt} = {}^{16}k_f \cdot (1-x_{liq}) \cdot A_{liq} - {}^{16}k_b\cdot (1-x_{vap}) \cdot A_{liq} \cdot \frac{N_{vap}}{V_{vap}}.
\end{equation}

## Equilibrium fractionation

Now we show how the equilibrium isotope fractionation between the liquid water and water vapor arises from the above formulation of mass balances. In a thermodynamic equilibrium, the time derivatives are equal to zero, which yields the following relationships:

\begin{equation}
\label{eq8}
\frac{{}^{18}N_{vap}}{dt} = 0 \quad\rightarrow\quad {}^{18}k_f \cdot x_{liq} = {}^{18}k_b\cdot x_{vap}  \cdot \frac{N_{vap}}{V_{vap}},
\end{equation}

\begin{equation}
\label{eq9}
\frac{{}^{16}N_{vap}}{dt} = 0 \quad\rightarrow\quad {}^{16}k_f \cdot (1-x_{liq}) = {}^{16}k_b\cdot (1-x_{vap}) \cdot \frac{N_{vap}}{V_{vap}}.
\end{equation}

By dividing Eq. 8 with Eq. 9, we obtain
\begin{equation}
\label{eq10}
\frac{{}^{18}k_f}{{}^{16}k_f}\cdot \frac{x_{liq}}{1-x_{liq}} = \frac{{}^{18}k_b}{{}^{16}k_b} \cdot \frac{x_{vap}}{1-x_{vap}}.
\end{equation}

If we define a fractionation factor $\alpha_{liq}(vap)$ according to
\begin{equation}
\label{eq11}
\alpha_{liq} (vap) \equiv \left(\frac{{}^{18}k_b}{{}^{16}k_b}\right) \left/ \left(\frac{{}^{18}k_f}{{}^{16}k_f}\right)\right. .
\end{equation}

the relationship in Eq. 10 can be simplified as

\begin{equation}
\label{eq12}
\frac{x_{liq}}{1-x_{liq}} = \alpha_{liq}(vap) \cdot \frac{x_{vap}}{1-x_{vap}}.
\end{equation}

In isotope chemistry, isotope composition of a sample is typically expressed as the isotope ratio or atom fraction. When applied to water, the isotope ratio is calculated from the number of ${}^{18}O$ and ${}^{16}O$ water molecules,
\begin{equation}
\label{eq13}
R = \frac{{}^{18}N}{{}^{16}N},
\end{equation}
whereas the atom fraction is calculated from the number of ${}^{18}O$ and total (${}^{18}O+ {}^{16}O$) water molecules,
\begin{equation}
\label{eq14}
x = \frac{{}^{18}N}{{}^{18}N + {}^{16}N}.
\end{equation}
These expressions imply the following relationship between $x$ and $R$:
\begin{equation}
\label{eq15}
x = \frac{R}{1+R}\quad \leftrightarrow \quad R = \frac{x}{1-x}
\end{equation}

Since absolute isotope ratios or atom fractions are difficult to determine analytically, they are typically expressed as a relative difference between two samples. These differences are typically expressed in the $\delta$ notation. Specifically, the difference of sample *b* versus sample *a* is defined according to \begin{equation}
\label{eq16}
\delta_a (b) = \frac{R_b}{R_a} - 1,
\end{equation}
where $R_a$ and $R_b$ are the corresponding isotope ratios. 

We apply this definition to our specific situation and calculate the difference between the isotope composition of water vapor relative to liquid water in a thermodynamic equilibrium according to
\begin{equation}
\label{eq17}
\delta_{liq} (vap) = \frac{R_{vap}}{R_{liq}}-1.
\end{equation}

Combining equations 12, 15, and 17, we find the following relationships between the equilibrium fractionation factor, $\delta_{liq} (vap)$, and the kinetic fractionation factor, $\alpha_{liq}(vap)$:
\begin{equation}
\label{eq18}
\delta_{liq}(vap) = \frac{1-\alpha_{liq}(vap)}{\alpha_{liq}(vap)} \quad \mathrm{or} \quad \alpha_{liq}(vap)=\frac{1}{1+\delta_{liq}(vap)}.
\end{equation}
We emphasize that the parameter $\alpha_{liq}(vap)$ relates the rate constants of the forward and backward reactions (evaporation and precipitation) for the light and heavy water isotopes (Eq. 11), whereas the parameter $\delta_{liq}(vap)$ relates the difference in the ${}^{18}O/{}^{16}O$ isotope ratio between water vapor and liquid water in a thermodynamic equilibrium (Eq. 17).

## Work in progress
To do: expand the above model formulation to two water reservoirs "connected" via the gas phase (Figure 1).

# Model definition

## Parameters describing the experimental setup

```{r}
V3    <- 1      # air volume (m3)
P     <- 101325 # air pressure (Pa)
Rgas  <- 8.314  # gas constant (SI units)
TK    <- 293.15 # temperature (K)
fH2O  <- 0.01   # molar fraction of water in air at 100% humidity
kb    <- 10     # rate constant of water precipitation (m d-1)
A1    <- 1      # area of the water reservoir 1 (m2) 
                # (corresponds to 3*4 subcores, each with inner diameter 1.2 cm)
A2    <- 1      # area of the water reservoir 2 (m2)
                # square of 25x25 cm
h1    <- 1e-3   # height of the water reservoir 1 (m)
h2    <- 1e-3   # height of the water reservoir 2 (m)
xini  <- 0.002  # natural 18O atom fraction
x1ini <- 0.002  # initial 18O atom fraction in water reservoir 1
x2ini <- 0.002  # initial 18O atom fraction in water reservoir 2
x3ini <- 0.002  # initial 18O atom fraction of water vapor (=water reservoir 3)
rho   <- 1e3    # water density (kg m-3)
MW    <- 18e-3  # molar weight of water (kg mol-1)
                # rho/MW is assumed to be the same for 16O-H2O and 18O-H2O!
```

## Derived parameters

```{r}
c3eq  <- P/(Rgas*TK) * fH2O # density of H2O molecules in air at 100% humidity (mol/m3)
V1    <- A1*h1              # volume of water in reservoir 1 (m3)
V2    <- A2*h2              # volume of water in reservoir 2 (m3)
# note: the density of particles (N/V=rho/MW) is assumed to be the same for 16O-H2O and 18O-H2O!
N1ini <- rho*V1/MW          # initial amount of H2O molecules in reservoir 1 (mol)
N2ini <- rho*V2/MW          # initial amount of H2O molecules in reservoir 2 (mol)
N3eq  <- c3eq*V3            # equilibrium amount of H2O molecules in reservoir 3 (mol)
```

## Model parameters
```{r}
pars <- c(delta = 0, # delta = steady state (R3/R1-1)*1e3, in permil!
          f1    = 1, # consider reservoir 1
          f2    = 0  # consider reservoir 2
          )
```

## Initial conditions

```{r}
# function to calculate initial state based on initial humidity of the air
humidityH2Oini <- function(humidity=1){
  # all values are in moles (16O-H2O and 18O-H2O isotopes separately)
  H2Oini <- c(N1_16 = N1ini*(1-x1ini),
              N1_18 = N1ini*x1ini,
              N2_16 = N2ini*(1-x2ini),
              N2_18 = N2ini*x2ini,
              N3_16 = N3eq*(1-x3ini) * humidity,
              N3_18 = N3eq*x3ini     * humidity
             )
  return(H2Oini)
}
```

# Model definition

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
WaterExchangeModel <-function(t, state, pars) {
  # t: time, state: state variables, pars: model parameters
  with (as.list(c(state, pars)),{
    
    # convert delta (in permil!) to alpha = kb_18/kb_16
    alpha <- 1/(delta*1e-3 + 1)
    
    # calculate all rate constants based on the rate constant 
    # of water precipitation (kb), the fractionation factor (alpha), 
    # and the equilibrium concentration c3eq
    kb_16 <- kb
    kb_18 <- kb * alpha
    kf_16 <- kb * c3eq
    kf_18 <- kb * c3eq
          
    # current amounts of H2O molecules in all reservoirs
    N1 <- N1_16+N1_18 
    N2 <- N2_16+N2_18 
    N3 <- N3_16+N3_18
    # current 18O atom fractions in all reservoirs
    x1 <- N1_18/N1 
    x2 <- N2_18/N2
    x3 <- N3_18/N3
    
    # rate expressions [mol/d]
    
    # exchange between reservoir 1 (water) and 3 (water vapour)
    Evap1_16 <- kf_16 * (1-x1) * A1 * (N1>0)
    Evap1_18 <- kf_18 *    x1  * A1 * (N1>0)
    Cond1_16 <- kb_16 * (1-x3) * A1 * N3/V3
    Cond1_18 <- kb_18 *    x3  * A1 * N3/V3
    
    # exchange between reservoir 2 (water) and 3 (water vapour)
    Evap2_16 <- kf_16 * (1-x2) * A2 * (N2>0)
    Evap2_18 <- kf_18 *    x2  * A2 * (N2>0)
    Cond2_16 <- kb_16 * (1-x3) * A2 * N3/V3
    Cond2_18 <- kb_18 *    x3  * A2 * N3/V3
    
    # Time-derivatives: dN/dt = production - consumption [mol/d]
    dN1_16.dt <- (-Evap1_16 + Cond1_16)*f1
    dN1_18.dt <- (-Evap1_18 + Cond1_18)*f1
    dN2_16.dt <- (-Evap2_16 + Cond2_16)*f2
    dN2_18.dt <- (-Evap2_18 + Cond2_18)*f2
    dN3_16.dt <- ( Evap1_16 - Cond1_16)*f1 + (Evap2_16 - Cond2_16)*f2
    dN3_18.dt <- ( Evap1_18 - Cond1_18)*f1 + (Evap2_18 - Cond2_18)*f2
    
    # return time-derivatives and ordinary variables as a list
    list(c(dN1_16.dt, dN1_18.dt,
           dN2_16.dt, dN2_18.dt,
           dN3_16.dt, dN3_18.dt),
         
         # number of molecules
         Ntot_16 = N1_16+N2_16+N3_16, # total 16O-water
         Ntot_18 = N1_18+N2_18+N3_18, # total 18O-water
         N1 = N1_16 + N1_18,          # total water molecules, liquid 1
         N2 = N2_16 + N2_18,          # total water molecules, liquid 2
         N3 = N3_16 + N3_18,          # total water molecules, vapor
         Ntot = N1_16+N2_16+N3_16+N1_18+N2_18+N3_18, # all water molecules
         
         # 18O atom fractions
         x1 = N1_18/(N1_16+N1_18),
         x2 = N2_18/(N2_16+N2_18),
         x3 = N3_18/(N3_16+N3_18),
         
         # delta values (permil): delta_a_vs_b = (Ra/Rb-1)*1e3
         delta_Vap_vs_LiqA =  ( (N3_18/N3_16) / (N1_18/N1_16) - 1 ) * 1e3,
         delta_LiqB_vs_LiqA = ( (N2_18/N2_16) / (N1_18/N1_16) - 1 ) * 1e3
         )
  })
}

```

# Model solution

We calculate the dynamic solution for the initial humidity close to 0 and three values of the equilibrium fractionation factor $\delta$ (0, 50, and 100 permil):

```{r, message=FALSE}
require(deSolve)
outtimes <- seq(from = 0, to = 1, length.out = 100) # time in days
SVini <- humidityH2Oini(humidity = 1e-6)
pars1 <- pars2 <- pars
pars1["delta"] <- 50
pars2["delta"] <- 100
out0 <- ode(y = SVini, parms = pars,  func = WaterExchangeModel, times = outtimes)
out1 <- ode(y = SVini, parms = pars1, func = WaterExchangeModel, times = outtimes)
out2 <- ode(y = SVini, parms = pars2, func = WaterExchangeModel, times = outtimes)
```

We plot the number of water molecules in the model reservoirs:

```{r, fig.height=6, fig.width=8}
plot(out0, out1, out2, 
     which=c("N1",   "N2",   "N3",   "Ntot",
             "N1_16","N2_16","N3_16","Ntot_16", 
             "N1_18","N2_18","N3_18","Ntot_18"),
     xlab="time (days)", ylab="mol O2", lty=1, las=1, lwd=2, mfrow=c(3,4))
```

We plot the ${}^{18}O$ atom fractions:

```{r, fig.height=3, fig.width=8}
plot(out0, out1, out2, which=c("x1","x2","x3"),
     xlab="time (days)", las=1, lwd=2, lty=1, mfrow=c(1,3))
```

Finally, we plot the $\delta$ values:

```{r, fig.height=4, fig.width=8}
plot(out0, out1, out2, 
     which=c("delta_LiqB_vs_LiqA","delta_Vap_vs_LiqA"),
     xlab="time (days)", las=1, lwd=2, lty=1, mfrow=c(1,2))
abline(h=pars1["delta"], lty=2, col=2)
abline(h=pars2["delta"], lty=2, col=3)
legend("bottomright", title="delta (permil):",
       legend=c(sprintf("%.1f",pars["delta"]), 
                sprintf("%.1f",pars1["delta"]),
                sprintf("%.1f",pars2["delta"])), 
       lty=1, lwd=2, col=1:3, bty="n")
```