---
title: |
  | Local Equilibrium Chemistry in R --- part I:
  | One Equilibrium Reaction
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Karline Soetaert and Lubos Polerecky, Utrecht University"
date: "March 2021"
output:
  pdf_document:
    number_sections: true
  html_document: default
linestretch: 1.15
---

## Answer

The solution to this problem follows directly from the explanation provided above once we realize that the species $A$ corresponds to $NH_3$, species $B$ corresponds to $H^+$, species $AB$ corresponds to $NH_4^+$, the lump-sum species $A_{tot}$ corresponds to total dissolved ammonium ($NH_x = NH_3 + NH_4^+$), and $B_{tot}$ corresponds to total positive charge ($H_x^+ = H^+ + NH_4^+$). A small modification is required, though, because we do not know the actual rate constants for the forward and backward reaction (this is typically the case for this type of problems). Thus, we will use the given value of $K_{eq}$ as the model parameter instead of $k_f$ and $k_b$.

First, we define the model function in analogy to the ``EqModel_num`` function. We can leave the ``solveB`` function untouched and reuse it.

```{r}
AmmoniaDegassing <- function(t, state, parms) {
  with (as.list(c(state, parms)), {
    
    # calculate H from NHx and Hx numerically, using solveB
    H   <- solveB(Keq = Keq, Atot = NHx, Btot = Hx) 
    # calculate NH3 and NH4+ from NHx and H (eq. 8 and 9)
    NH3 <- Keq/(Keq+H) * NHx
    NH4 <- NHx - NH3
    
    # mass balance equations for NHx and Hx (eq. 6)
    dNHx <- -lambda * NH3
    dHx  <- 0
    
    return(list(c(dNHx, dHx), 
                H = H, NH3 = NH3, NH4 = NH4, # NH4 corresponds to [NH4+]
                pH = -log10(H),     # return also pH
                Q = NH3*H/NH4 ))    # return also Q
  })
}
```

Next, we define the model parameters and initial conditions. Here, we need to keep in mind that $pH=-log([H^+])$, where the proton concentration is in $mol~L^{-1}$! Thus, all state variables, as well as the equilibrium constant $K_{eq}$, must be in this unit, too. Because the rate constant is in the unit of $d^{-1}$, we are going to solve the model on the time scale of days.

```{r}
pars      <- c (Keq    = 5.38e-10,  # mol/L
                lambda = 1)         # 1/d

# Initial concentration of total dissolved ammonia (given)
NHx.ini   <- 1e-3 # mol/L
# Initial proton concentration, calculated from pH (given)
H.ini     <- 10^(-8) # 10^(-pH), mol/L
# Initial total positive charge concentration, calculated from eq. 7
Hx.ini    <- H.ini + H.ini/(pars[["Keq"]] + H.ini) * NHx.ini

# initial conditions, lump-sum state variables NHx and Hx
state.ini <- c(NHx = NHx.ini, Hx = Hx.ini)
```

Now we are ready to run the model and plot the results as a function of time. We calculate the results for 20 days.

```{r, fig.height=5, fig.width=10}
require(deSolve)
times <- seq(from=0, to=20, length.out=100) # days
out   <- ode(y=state.ini, times=times, func=AmmoniaDegassing, parms=pars)
plot(out, mfrow=c(2,4), xlab="time (d)")
```
We see that the most dramatic changes occur within the first 5 days, where the $pH$ decreases from 8 until about 6. This shift towards a lower $pH$ (higher $[H^+]$) causes the speciation of total ammonium to shift towards lower $[NH_3]$. After 5 days, the $NH_3$ concentration is so low that the rate of its removal is close to zero (because the rate is proportional to $[NH_3]$), leading to a much slower $pH$ dynamics.

An interesting outcome of this model is that the *total* ammonia concentration in the lake does *not* decrease appreciably (from $1~mmol~L^{-1}$ to about $0.95~mmol~L^{-1}$). This means that degassing of ammonia from a $pH$-unbuffered system is *not* going to remove the ammonium from the lake water. The lake would stop smelling bad, but it would not stop being polluted. Note, however, that this result is derived based on an unrealistic assumption that ammonia degassing is the only process that affects the $pH$ of the lake water. To overcome this problem, we need to enhance the model with other relevant processes, namely water dissociation and buffering by dissolved inorganic carbon, which will be done in the next exercises of the equilibrium chemistry series.
