---
title: "Farmer's dilemma---optimisation of crop cultures in the presence of weeds and in response to eutrophication"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Karline Soetaert and Lubos Polerecky, Utrecht University"
date: "March 2021"
output:
  pdf_document:
    toc: no
---

# Crop and weed model including economics --- Answers

To estimate the net financial gain for the farmer, we need to take into account all costs and profits associated with the cultivation and sale of the plants.

We do this by adding an extra state variable: the integrated cost. Its dynamics are quite simple:

The *Cost* increases with the daily expenditures of culturing (labor), which is either 1 euro per $m^2$ and per week ($1~EUR~m^{-2}~wk^{-1}$) if there is no continuous fertilization, or $1.025~EUR~m^{-2}~wk^{-1}$ if continuous fertilization *is* performed. Additionally, the costs are augmented with the cost of the fertiliser. 

We write for the balance equations:
$$\frac{dCost}{dt} = laborCost + costP\times Padded$$
Here Cost is in $EUR~m^{-2}$, and the rates are in $EUR~m^{-2}~d^{-1}$.

The initial cost of planting is $Cost_{ini}=0.5~EUR~m^{-2}$. 

In case the fertilisation is done before planting, we need to also include the labor cost and depreciation of material used ($5~EUR~m^{-2}$) in addition to the cost of the fertiliser. 

## R-implementation --- economic part

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
require(deSolve)  # package with solution methods

# state variables, units = [molP/m2] - except for COST: [Euro/m2]
state.ini <- c(WEED = 0.001, CROP = 0.005, P1 = 0.1, P2 = 0.1, COST = 0.5, Plost = 0)

# parameters
pars <- c(
  Paddition   = 0.9/90,    # [molP/m2/d]  Rate of P supply
  percolation = 0.05,      # [/d]         Rain rate
  ksCrop      = 2e-3,      # [molP/m2]    Monod coefficient for P uptake
  ksWeed      = 0.5e-3,    # [molP/m2]    Monod coefficient 
  ktot        = 0.300,     # [molP/m2]    Carrying capacity - space limitation
  rGcrop      = 1*0.125,   # [/d]         Growth rate, 
  rGweed      = 1*0.1,
  rMcrop      = 0.0,       # [/d]         Loss rate (e.g. mortality), 
  rMweed      = 0.0,
  N           = 25,        # [ind/m2]     Density of crop plants
  P2WW        = 62000,     # [gWW/molP]   31/0.25/0.002
  laborCost   = 1.025/7,   # [euro/m2/d]  daily cost of maintenance
  PCost       = 3/1000*31, # [euro/molP]  cost of the fertiliser, 3 eur/kg
  priceCROP   = 1/1000     # [euro per g] price for plants > 300g, 1 eur/kg of wet weight
)

# Model function
Crops <- function(t, state, parms) {
  with (as.list(c(state, parms)), {

	# variables needed for rate expressions
    TotBiom    <- WEED + CROP       # total plant biomass
    NutWeed    <- P1 + P2           # nutrients accessible to weed
    NutCrop    <- P1                # nutrients accessible to crops
  	partP1     <- P1/NutWeed        # part of P for weed from first layer
    
    # Rate expressions - all rates in units of [molP/m2/day]
    WeedGrowth <- rGweed * WEED*(1-TotBiom/ktot) *NutWeed/(NutWeed+ksWeed)
    CropGrowth <- rGcrop * CROP*(1-TotBiom/ktot) *NutCrop/(NutCrop+ksCrop) 
    
    WeedLoss   <- rMweed * WEED     # death and other loss terms
    CropLoss   <- rMcrop * CROP

    Percolate  <- P1 * percolation  # transfer of P from layer 1 to layer 2
    Ploss      <- P2 * percolation  # transfer of P from layer 2 to deep soil

    # Mass balances [molP/m2/day]
    dWEED      <- WeedGrowth - WeedLoss
    dCROP      <- CropGrowth - CropLoss
    dP1        <- Paddition - Percolate - CropGrowth - WeedGrowth * partP1
    dP2        <- Percolate - Ploss - WeedGrowth * (1 - partP1)
	  dPlost     <- Ploss + WeedLoss + CropLoss

	# Individual weight of a crop plant
    Weight     <- CROP/N*P2WW         # gram wet weight per plant

    # Economic part
    dailyCost  <- laborCost + PCost*Paddition

    
    Price      <- CROP*P2WW * (Weight > 300)*priceCROP  # total price per m2 of plants
    Profit     <- Price - COST
    
    dCOST      <- dailyCost
   
    list(c(dWEED, dCROP, dP1, dP2, dCOST, dPlost), 
      TotBiom = TotBiom, Weight = Weight,
	  TotP = WEED + CROP + P1 + P2 + Plost,
      SpaceLimFact = (1-TotBiom/ktot),
      NutLimFactCrop = NutCrop/(NutCrop+ksCrop),
      NutLimFactWeed = NutWeed/(NutWeed+ksWeed), 
	    Price = Price, Profit = Profit)
  })

}
```

The model is solved thrice over 90 days. 

* Run 1 implements continuous P addition.
* Run 2 implements one-time P fertilization.
* Run 3 implements continuous P addition at a *quarter* of the fertilization rate compared to run 1.

The 3rd run is used to assess the environmental and economic impact of reduced fertilization rate.

```{r, fig.height = 10.5, fig.width = 10}
# output time
outtimes <- seq(from = 0, to = 90, length.out = 100)  # run for 3 months

# Continuous P-addition
state.ini["COST"] <- 0.5        # price of planting & fertilizer cost
out <- ode(y = state.ini, parms = pars, func = Crops, times = outtimes)

# Initial P fertilisation
# Change the parameter values and initial conditions
pars2              <- pars
pars2["Paddition"] <- 0.0       # no continuous fertilization
pars2["laborCost"] <- 1./7      # slightly cheaper if no continuous fertilization

state2.ini         <- state.ini
state2.ini["P1"]   <- 1         # higher value for P1
state2.ini["COST"] <- 0.5 + 5 +           # price of planting, initial 
                      0.9 * pars["PCost"] # fertilisation (labor) & fertilizer cost
                                          # (0.9 mol P m-2 is added)
out2 <- ode(y = state2.ini, parms = pars2, func = Crops, times = outtimes)

# Continuous P-addition at a quarter of the original rate
state3.ini         <- state.ini
pars3              <- pars
pars3["Paddition"] <- pars3["Paddition"]*0.25
out3 <- ode(y = state.ini, parms = pars3, func = Crops, times = outtimes)

# visualise results
plot(out, out2, out3, lty = 1, lwd = 2, mfrow=c(4,4))
abline(h = 0)   # above the line: true profit; below the line: loss.
plot.new()
legend("topleft", c("continuous",  "initial", "continuous-quarter"), 
       col = 1:3, lty = 1, lwd = 2)
```

Based on the discussion of the previous results, one can discuss independently the economic and environmental implications of these model results --- the graphs speak for themselves.
