---
title: "Farmer's dilemma---optimisation of crop cultures in the presence of weeds and in response to eutrophication"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Karline Soetaert and Lubos Polerecky, Utrecht University"
date: "March 2021"
output:
  pdf_document:
    toc: no
---

# Crops and weeds model --- Answers

## R-implementation --- biological part

In addition to the amounts of P in the soil compartments, crops and weeds, we also introduce a state variable *Plost* to be able to trace the amount of P lost to the bottom soil due to percolation. Furthermore, we export the factors $(1-TotBiom/ktot)$, $NutCrop/(NutCrop+ksCrop)$ and $NutWeed/(NutWeed+ksWeed)$ to be able to trace what limits the plant growth. These two pieces of information are relevant when assessing the environmental impact of the agriculture and interpreting the P dynamics predicted by the model.

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
require(deSolve)  # package with solution methods

# state variables, units = molP/m2
state.ini <- c(WEED = 0.001, CROP = 0.005, P1 = 0.1, P2 = 0.1, Plost = 0)

# parameters
pars <- c(
  Paddition   = 0.9/90,  # [molP/m2/d]  Rate of P supply
  percolation = 0.05,    # [/d]         Rain rate dilution parameter
  ksCrop      = 2e-3,    # [molP/m2]    Monod coefficient for P uptake
  ksWeed      = 0.5e-3,  # [molP/m2]    Monod coefficient 
  ktot        = 0.300,   # [molP/m2]    Carrying capacity - space limitation
  rGcrop      = 0.125,   # [/d]         Growth rate, 
  rGweed      = 0.1,
  rMcrop      = 0.0,     # [/d]         Loss rate (e.g. mortality), 
  rMweed      = 0.0,
  N           = 25,      # [ind/m2]     Density of crop plants
  P2WW        = 62000    # [gWW/molP]   31/0.25/0.002
)

# Model function
Crops <- function(t, state, parms) {
  with (as.list(c(state, parms)), {

	# variables needed for rate expressions
    TotBiom    <- WEED + CROP       # total plant biomass
    NutWeed    <- P1 + P2           # nutrients accessible to weed
    NutCrop    <- P1                # nutrients accessible to crops
    partP1     <- P1/NutWeed        # part of P for weed from first layer

    # Rate expressions - all rates in units of [molP/m2/day]
    WeedGrowth <- rGweed * WEED*(1-TotBiom/ktot) *NutWeed/(NutWeed+ksWeed)
    CropGrowth <- rGcrop * CROP*(1-TotBiom/ktot) *NutCrop/(NutCrop+ksCrop) 
    
    WeedLoss   <- rMweed * WEED     # death and other loss terms
    CropLoss   <- rMcrop * CROP

    Percolate  <- P1 * percolation  # transfer of P from layer 1 to layer 2
    Ploss      <- P2 * percolation  # transfer of P from layer 2 to deep soil

    # Mass balances [molP/m2/day]
    dWEED      <- WeedGrowth - WeedLoss
    dCROP      <- CropGrowth - CropLoss
    dP1        <- Paddition - Percolate - CropGrowth - WeedGrowth * partP1
    dP2        <- Percolate - Ploss - WeedGrowth * (1 - partP1)
    dPlost     <- Ploss + WeedLoss + CropLoss
   
    # Individual weight of a crop plant
    Weight     <- CROP/N*P2WW         # wet weight, gram per plant
    
    list(c(dWEED, dCROP, dP1, dP2, dPlost), 
      TotBiom = TotBiom, Weight = Weight,
      TotP = WEED + CROP + P1 + P2 + Plost,       # output the total P, mass balance check
      SpaceLimFact = (1-TotBiom/ktot),            # output the "space-limitation" factor
      NutLimFactCrop = NutCrop/(NutCrop+ksCrop),  # output the nutrient limitation factor
      NutLimFactWeed = NutWeed/(NutWeed+ksWeed))  # for both crop and weed
  })

}
```

The model is solved thrice over 90 days. 

* Run 1 implements continuous P addition (with weeds).
* Run 2 implements one-time P fertilization (with weeds).
* Run 3 implements continuous P addition (without weeds).

To compare the models in a fair way, we assume that the one-time and continuos addition of P are done in such a way that the *total* amount of P *added* to the soil is the *same* after 90 days.

```{r, fig.height = 8.5, fig.width = 10}
# output time
outtimes <- seq(from = 0, to = 90, length.out = 100)  # run for 3 months

# Continuous P-addition, with weeds
out <- ode(y = state.ini, parms = pars, func = Crops, times = outtimes)

# One-time P ferilisation, with weeds
pars2              <- pars
pars2["Paddition"] <- 0.0       # No P-addition
state2.ini         <- state.ini
state2.ini["P1"]   <- 1         # Higher value for P1 at start
out2 <- ode(y = state2.ini, parms = pars2, func = Crops, times = outtimes)

# Continuous P-addition, without weeds
state3.ini         <- state.ini
state3.ini["WEED"] <- 0
out3 <- ode(y = state3.ini, parms = pars, func = Crops, times = outtimes)

# visualise it 
plot(out, out2, out3, lty = 1, lwd = 2, mfrow=c(3,4))

legend("bottomleft", c("continuous+weed",  "one-time", "continuous, no weed"), 
       col = 1:3, lty = 1, lwd = 2)
```

We see that during the initial 42 days, crops and weeds grow practically at the same rate. This is because P in the soil is *not* limiting their growth. Rather, their growth is limited by space, as the space limitation factor $(1-TotBiom/ktot)$ gradually decreases from 1 to about 0.2. 

The situation changes rather abruptly after about 42 days. In the 2nd scenario, the crop becomes limited by P due to the removal by percolation, whereas the continuous fertilization ensures that this does not happen in the 1st scenario. As a result, the crop continues to grow, reaching about 25% higher stock per $m^2$ after 90 days in the first compared to the second scenario. This is, however, *not* the case for the *weed*, which continues to grow rapidly even after 42 days in the 2nd scenario. This is, clearly, due to the fact that its roots reach deeper into the soil and thus have more access to P. Although the continuous growth of weeds has a negative impact on the growth of crops due to space limitation, it is only marginal. Without weeds, the crop weight after 90 days would be $743~g/plant$ without weeds in comparison to $680~g/plant$ with weeds (i.e., about 10% higher yield), as indicated by the results of the scenario 3 (green line).

Another notable result of this simulation is that *most* of the fertilizer is *lost* in this agriculture. Only about 0.27 out of 1.1 $mol~P~m^{-2}$ (about 25% yield on fertilizer) is converted into profitable biomass (crop) in the continuous fertilization scenario. This fraction decreases to about 0.22 out of 1.1 $mol~P~m^{-2}$ (20% yield) in the one-time fertilization scenario. These results illustrate the potential evironmental impact of this agriculture.
