---
title: "RTM course - transport fluxes"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on formulating mixing and advective fluxes for transport models"
date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first.

### Video timeline
* 0:00 Introduction
* 0:28 Advective transport
* 2:26 Molecular diffusion
* 4:34 Dispersion
* 7:04 Bioturbation
* 8:00 Diffusion coefficient = distance * velocity
* 8:50 Comparison between diffusion, dispersion and bioturbation
* 9:50 Conclusion - total transport flux

![](https://www.youtube.com/watch?v=zAj4m6tGDhQ&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=11)

You can also read the following section(s) of the textbook:

* Chapter 3.4. Transport in a One-Dimensional Model

(You can download each chapter of the textbook via the e-book University library for free!)

## Transport fluxes

Consider the following expression for the transport flux:

$$
J =  v \cdot C
$$

```{r adv, echo=FALSE}
msg <- sprintf("The transport flux $J$ is in $mol~m^{-2}s^{-1}$, the advection velocity $v$ is in $m~s^{-1}$, the concentration $C$ is in $mol~m^{-3}$.")

question("Which of the following statements is true?", type = "learnr_checkbox",
 answer(sprintf("This expression describes the flux due to advective transport"), correct = TRUE),
 answer(sprintf("This expression describes the flux due to dispersion")), 
 answer(sprintf("This expression describes the flux due to diffusion")), 
 answer(sprintf("This expression describes the flux due to bioturbation")), 

 answer(sprintf("$v$ is the flow rate of the medium (e.g., water), in $m^3~s^{-1}$")),
 answer(sprintf("$v$ is the bulk velocity of the medium (e.g., water), in $m~s^{-1}$"), correct = TRUE),
 answer(sprintf("$v$ is the dispersion or diffusion coefficient of the medium (e.g., water), in $m^2~s^{-1}$")),
 answer(sprintf("$v$ is the bulk volume of the medium (e.g., water), in $m^{3}$")),
 answer(sprintf("$C$ is the concentration of a substance in the medium, in $mol~m^{-3}$"), correct=TRUE),
 answer(sprintf("$C$ is the volume of the medium, in $m^3$")),
 answer(sprintf("$C$ is the concentration gradient of a substance in the medium, in $mol~m^{-4}$")),

 incorrect=msg,
 correct=msg)
```

Consider the following expression for the transport flux:

$$
J =  -D \cdot \frac{\partial C}{\partial x}
$$

```{r difflux, echo=FALSE}
msg <- sprintf("The flux $J$ is in $mol~m^{-2}~s^{-1}$, the diffusion, dispersion or bioturbation mixing coefficient $D$ is in $m^2~s^{-1}$, the concentration $C$ is in $mol~m^{-3}$, and the concentration gradient $\\frac{\\partial C}{\\partial x}$ is in $mol~m^{-4}$.")

question("Which of the following statements is true?", type = "learnr_checkbox",
 answer(sprintf("This expression describes the flux due to advective transport")),
 answer(sprintf("This expression describes the flux due to dispersion"), correct = TRUE), 
 answer(sprintf("This expression describes the flux due to diffusion"), correct = TRUE), 
 answer(sprintf("This expression describes the flux due to bioturbation"), correct = TRUE), 

 answer(sprintf("$D$ is the flow rate of the medium (e.g., water), in $m^3~s^{-1}$")),
 answer(sprintf("$D$ is the bulk velocity of the medium (e.g., water), in $m~s^{-1}$")),
 answer(sprintf("$D$ is the dispersion coefficient in the medium (e.g., water), in $m^2~s^{-1}$"), correct = TRUE),
 answer(sprintf("$D$ is the diffusion coefficient in the medium (e.g., water), in $m^2~s^{-1}$"), correct = TRUE),
 answer(sprintf("$D$ is the bulk volume of the medium (e.g., water), in $m^{3}$")),
 answer(sprintf("$C$ is the concentration of a substance in the medium, in $mol~m^{-3}$"), correct=TRUE),
 answer(sprintf("$C$ is the volume of the medium, in $m^3$")),
 answer(sprintf("$C$ is the concentration gradient in the medium, in $mol~m^{-4}$")),
 answer(sprintf("$\\frac{\\partial C}{\\partial x}$ is the concentration gradient in the medium, in $mol~m^{-2}$")),
 answer(sprintf("$\\frac{\\partial C}{\\partial x}$ is the concentration gradient in the medium, in $mol~m^{-4}$"), correct=TRUE),

 incorrect=msg,
 correct=msg)
```

## Qualitative statements

```{r trantruefalse, echo=FALSE}
question("Which of the following statements is true?", type = "learnr_checkbox",
 answer(sprintf("Molecular diffusion always decreases the concentration (hence the negative sign)")),
 answer(sprintf("In molecular diffusion transport goes against the concentration gradient (hence the negative sign)")),
 answer(sprintf("In molecular diffusion net transport goes against the concentration gradient (hence the negative sign)"), correct=TRUE),
 answer("Advective transport is due to a *directional* movement of particles, whereas diffusive transport is due to a *random* movement of particles.", correct = TRUE),  
 answer("Advective transport is due to a *random* movement of particles, whereas diffusive transport is due to a *directional* movement of particles.")
)
```

Complete the sentence to make a correct statement:

```{r tranqualitative, echo=FALSE}
question("In molecular diffusion, the diffusion coefficient can be written as a [. . .] of a characteristic distance that a particle travels between collisions with other particles and the characteristic velocity of the particle.", type = "learnr_checkbox",
 answer("sum"),
 answer("product", correct = TRUE),
 answer("ratio")
)
```

## Molecular diffusion

Complete the sentence to make a correct statement:

```{r diffusion, echo=FALSE}
question("In molecular diffusion, if the concentration of a substance is constant (i.e., not varying in space), there is", type = "learnr_checkbox",
 answer("a constant net transport of the substance."),
 answer("no transport of the substance."),
 answer("no net transport of the substance.", correct=TRUE)
)
```

```{r diffusion2, echo=FALSE}
question("In molecular diffusion, if the concentration gradient of a substance is constant (i.e., not varying in space), there is", type = "learnr_checkbox",
 answer("a constant net transport of the substance", correct=TRUE),
 answer("no transport of the substance"),
 answer("no net transport of the substance")
)
```

## Bioturbation

Complete the sentence to make a correct statement:

```{r bioturbation, echo=FALSE}
question("Bioturbation can be described as a diffusion-like process on time scales that are [. . .] intervals between translocation events of individual particles.", type = "learnr_checkbox",
 answer("much shorter than"),
 answer("of the same order as"),
 answer("much longer than", correct=TRUE)
)
```

## Scales of impact

```{r scaling, echo=FALSE}
question(sprintf("What is the relationship between the typical values of the dispersion coefficient ($D_{disp}$), the molecular diffusion coefficient ($D_{mol}$), and the bioturbation mixing coefficient ($D_{bio}$)?"), type = "learnr_checkbox",
 answer(sprintf("$D_{disp} \\gg D_{bio} \\gg D_{mol}$")),
 answer(sprintf("$D_{disp} \\ll D_{bio} \\ll D_{mol}$")),
 answer(sprintf("$D_{disp} \\ll D_{mol} \\ll D_{bio}$")),
 answer(sprintf("$D_{disp} \\gg D_{mol} \\gg D_{bio}$"), correct = TRUE)
)
```

## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

## References

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.
