---
title: "RTM course - Transport processes"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on formulating transport models (on a conceptual level)"

date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first.

### Video timeline
* 0:00 Introduction: towards spatially resolved models
* 0:53 Including transport on a conceptual level
* 3:59 Mass balance: transport-reaction equation

![](https://www.youtube.com/watch?v=4ZwKZiobm0Y&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=10)

You can also read the following section(s) of the textbook:

* Chapter 3. Spatial Components and transport

(You can download each chapter of the textbook via the e-book University library for free!)

## General equation I

$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{(A\cdot J)}}{\partial x}
$$

The differential equation above describes the variation of the concentration of a substance in time and space along one dimension. 


```{r GTE1, echo=FALSE}
question("Which of the following statements is true?", type = "learnr_checkbox",
 answer("We call this equation the one-dimensional reaction equation"),
 answer("We call this equation the one-dimensional transport equation"),
 answer("We call this equation the one-dimensional reaction-transport equation", correct=TRUE),
 answer("The reaction term is the first on the right hand side", correct=TRUE),  
 answer("The reaction term is the second on the right hand side"),  
 answer("The transport term is the first on the right hand side"),  
 answer("The transport term is the second on the right hand side", correct=TRUE)  
)
```

## General equation II

$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{(A\cdot J)}}{\partial x}
$$

```{r GTE2, echo=FALSE}
question("Which of the following statements is true?", type = "learnr_checkbox",
 answer(sprintf("C is the transport flux of the substance")),
 answer(sprintf("C is the concentration of the substance"), correct = TRUE),
 answer(sprintf("t is time in s"), correct = TRUE),
 answer(sprintf("t is space in m"), correct = FALSE),
 answer(sprintf("x is time in s"), correct = FALSE),
 answer(sprintf("x is space in m"), correct = TRUE)
)  
```  

## General equation III

$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{(A\cdot J)}}{\partial x}
$$


```{r GTE3, echo=FALSE}
question("Which of the following statements is true?", type = "learnr_checkbox",
 answer(sprintf("R denotes the concentration in $mol~m^{-3}$"), correct = FALSE),
 answer(sprintf("R denotes the net production rate in $mol~m^{-3}$"), correct = FALSE),
 answer(sprintf("R denotes the net production rate in $mol~m^{-2}~s^{-1}$"), correct = FALSE),
 answer(sprintf("R denotes the net production rate in $mol~m^{-3}~s^{-1}$"), correct = TRUE),
 answer(sprintf("R denotes the flux in x-direction, in $mol~m^{-2}~s^{-1}$"), correct = FALSE),
 answer(sprintf("J denotes the concentration in $mol~m^{-3}$"), correct = FALSE),
 answer(sprintf("J denotes the net production rate in $mol~m^{-3}$"), correct = FALSE),
 answer(sprintf("J denotes the net production rate in $mol~m^{-2}~s^{-1}$"), correct = FALSE),
 answer(sprintf("J denotes the net production rate in $mol~m^{-3}~s^{-1}$"), correct = FALSE),
 answer(sprintf("J denotes the transport flux in x-direction, in $mol~m^{-2}~s^{-1}$"), correct = TRUE),
 answer(sprintf("A is the length, in $m$, through which the substance is transported"), correct = FALSE),
 answer(sprintf("A is the surface area, in $m^2$, through which the substance is transported"), correct = TRUE),
 answer(sprintf("A is the volume, in $m^3$, through which the substance is transported"), correct = FALSE)
)  
```  

## Transport flux

$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{(A\cdot J)}}{\partial x}
$$

```{r transport, echo=FALSE}
msg <- sprintf("Each of the terms in the differential equation above has a SI unit of $mol~m^{-3}~s^{-1}$, as the SI unit of the concentration, C, is $mol~m^{-3}$, and the SI unit of time, $t$, is $s$ (second). The flux J in the second term on the right therefore has to have the unit of $mol~m^{-2}~s^{-1}$, because the units of $A$ cancel out and the unit of $x$ is $m$ (meter).")

question("Which of the following statements is true?", type = "learnr_checkbox",
 answer(sprintf("The transport flux $J$ describes the volume of a substance transported per unit time.")),
 answer(sprintf("The transport flux $J$ describes the amount of a substance transported per unit time.")),
 answer(sprintf("The transport flux $J$ describes the amount of a substance transported per unit area per unit time."), correct=TRUE),
 answer(sprintf("The transport flux $J$ describes the amount of a substance transported per unit volume per unit time.")),
 answer(sprintf("The transport flux $J$ describes the amount of a substance transported per unit length per unit time.")),
 answer(sprintf("The SI unit of the transport flux is $s^{-1}$.")),  
 answer(sprintf("The SI unit of the transport flux is $mol~m^{-2}~s^{-1}$."), correct=TRUE),
 answer(sprintf("The SI unit of the transport flux is $mol~m^{-3}~s^{-1}$.")),
 answer(sprintf("The SI unit of the transport flux is $mol~m^{-1}~s^{-1}$.")),  

 incorrect=msg,
 correct=msg)
```

## Simplified equation

Starting from the general equation

$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{(A\cdot J)}}{\partial x},
$$

we most often use the simplified version:

$$
\frac{\partial{C}}{\partial t} = R - \frac{\partial{J}}{\partial x}.
$$
```{r transport2, echo=FALSE}
msg <- sprintf("Mathematically, if $A$ is constant (i.e., it is *not* a function of $x$), it can be taken out of the derivative of the general equation. This yields the second - simplified - equation.")

question("When does the above equation apply?", type = "learnr_checkbox",
 answer(sprintf("When the transport flux $J$ is constant.")),
 answer(sprintf("When the area $A$ is zero.")),
 answer(sprintf("The area $A$ is constant."), correct=TRUE),
 answer(sprintf("The area $A$ is 1.")),

 incorrect=msg,
 correct=msg)
```

## Textbook and Slides

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>