---
title: |
  | Local Equilibrium Chemistry in R --- part II:
  | Three Equilibrium Reactions
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky and Karline Soetaert, Utrecht University"
date: "February 2021"
output:
  pdf_document:
    number_sections: false
  html_document: default
linestretch: 1.15
---

# Answer

## Task 1: Mass balance equations of the carbonate system

Assuming that the forward and backward reactions are elementary, we can use the first-order kinetics to describe the rate for each reaction. Considering the stoichiometry of Reactions 1--3, we obtain the following mass balance equations:

$$
\begin{array}{rclr}
\displaystyle
\frac{d[H_2CO_3]}{dt} & = & -R_1 + \color{red}{R_C} & \qquad (1a)\\[5mm]
\displaystyle
\frac{d[HCO_3^-]}{dt} & = & R_1 - R_2 & \qquad (1b) \\[5mm]
\displaystyle
\frac{d[CO_3^{2-}]}{dt} & = & R_2  & \qquad (1c)\\[5mm]
\displaystyle
\frac{d[OH^-]}{dt} & = &  R_3  & \qquad (1d)\\[5mm]
\displaystyle
\frac{d[H^+]}{dt} & = & R_1 + R_2 + R_3 & \qquad (1e)\\[5mm]
\displaystyle
\frac{d[H_2O]}{dt} & = & -R_3  & \qquad (1f)\\[5mm]
\end{array}
$$

where the individual *net* rates ($R_1$, $R_2$ and $R_3$) are defined as the difference between the forward and backward reaction rates:

$$
\begin{array}{rcll}
R_1 & = & k_{1f} \cdot [H_2CO_3] - k_{1b} \cdot [HCO_3^-] \cdot [H^+] 
& \hspace{2cm} (2a) \\[5mm]
R_2 & = & k_{2f} \cdot [HCO_3^{-}] - k_{2b} \cdot [CO_3^{2-}] \cdot [H^+] 
& \hspace{2cm} (2b) \\[5mm]
R_3 & = & k_{3f} \cdot [H_2O] - k_{3b} \cdot [OH^{-}] \cdot [H^+] 
& \hspace{2cm} (2c)
\end{array}
$$

Note that the rate of the slow process, $R_C$, only affects $H_2CO_3$.

## Task 2: Equilibrium conditions

In the absence of the slow process (i.e., $R_C=0$), the carbonate species and $OH^-$ and $H^+$ will reach an equilibrium as a result of the reversible reactions 1--3. In this equilibrium, the time derivative of each species in the system is zero. Based on equations 1a and 2a, we obtain $R_1=0$, which yields the relationship

$$
K_{1} = \frac{ [HCO_3^-] \cdot [H^+] }{ [H_2CO_3] }. \qquad\hspace{2cm} (3a)
$$
Similarly, based on equation 1c and 2b, we obtain $R_2=0$, which yields the relationship
$$
K_{2}  = \frac{ [CO_3^{2-}] \cdot [H^+] }{[HCO_3^-]} .
\qquad\hspace{2cm} (3b)
$$
Finally, the equilibrium condition implies $R_3=0$, which combined with Eq. 2c yields
$$
K_w'  = \frac{ [OH^-] \cdot [H^+] }{[H_2O]}. 
\qquad\hspace{2cm} (3c)
$$
Here, we defined the dissociation constants according to $K_1 = k_{1f}/k_{1b}$, $K_2 = k_{2f}/k_{2b}$, and $K_w' = k_{3f}/k_{3b}$. If relationships 3a--3c are valid, then the time derivatives in equations 1b, 1d and 1e are zero as well. This means that these latter three equations do *not* provide any *extra* information for the equilibrium concentrations.

**Note:** In literature, the *ionic product of water*, $K_w = [OH^-]\cdot [H^+] \approx 10^{-14}~mol^2~L^{-2}$ is more commonly reported in the context of water self-ionization. This value should, however, *not* be confused with the equilibrium constant $K_w'$. The correct value of $K_w'$ is calculated by dividing $K_w$ with the concentration of water, $[H_2O]\approx 55.3~mol~L^{-1}$, as shown in Eq. 3c (see table in Task 2).

## Task 3: Lump-sum species

By summing equations 1a, 1b and 1c, and additionally considering the proposed definition of the lump-sum variable $[DIC]$, we obtain 
$$
\frac{d[DIC]}{dt} = R_C. \qquad\hspace{2cm} (4a)
$$
Similarly, by summing equation 1b with equation 1d and equation 1c multiplied with a factor of 2, then subtracting equation 1e, and additionally considering the proposed definition of the lump-sum variable $[ALK]$, we obtain
$$
\frac{d[ALK]}{dt} = 0. \qquad\hspace{2cm} (4b)
$$
Finally, by summing equation 1d with equation 1f and additionally considering the proposed definition of the lump-sum variable $[H_xO]$, we obtain
$$
\frac{d[H_xO]}{dt} = 0. \qquad\hspace{2cm} (4c)
$$

### Key take-home message

These equations show that under the assumption of a local chemical equilibrium, $DIC$ is *added* at a rate of the *slow process* ($R_C$), while $ALK$ and $H_xO$ are *conserved* (not changing over time). Notably, $DIC$, $ALK$ and $H_xO$ are *not* affected by the fast reversible reactions!

## Task 4: From lump-sum species to original species

By solving for $[H_2CO_3]$ and $[CO_3^{2-}]$ in equations 3a and 3b, and by substituting the result to the definition of $[DIC]$, we obtain 

$$
[DIC] = [HCO_3^-] + [H^+]\cdot \frac{[HCO_3^-]}{K_{1}} + K_{2}\cdot \frac{ [HCO_3^-]}{[H^+]}
$$

When we express each term on the right-hand side using the common denominator ($K_1\cdot [H^+]$), we obtain

$$
[DIC] = \frac{K_1\cdot[H^+]\cdot[HCO_3^-]}{K_1\cdot[H^+]} + \frac{ [H^+]^2\cdot[HCO_3^-]}{K_1 \cdot[H^+]} + \frac{ K_1 \cdot K_{2} \cdot[HCO_3^-]}{K_1 \cdot [H^+]}
$$

This equation can easily be solved for $[HCO_3^-]$:

$$
\boxed{~[HCO_3^-]= \frac{K_{1}\cdot [H^+]}{[H^+]\cdot [H^+] + K_{1}\cdot [H^+] +  K_{1}\cdot K_{2}}\cdot [DIC]~}
\qquad\hspace{2cm} (5a)
$$

Combining this result with equation 3a, we obtain 

$$
\boxed{~[H_2CO_3] = \frac{[H^+]\cdot [H^+]}{[H^+]\cdot [H^+] + K_{1}\cdot [H^+] +  K_{1}\cdot K_{2}}\cdot [DIC]~}
\qquad\hspace{2cm} (5b)
$$

Similarly, combining the result with equation 3b, we obtain

$$
\boxed{~[CO_3^{2-}] = \frac{K_1\cdot K_2}{[H^+]\cdot [H^+] + K_{1}\cdot [H^+] +  K_{1}\cdot K_{2}}\cdot [DIC]~}
\qquad\hspace{2cm} (5c)
$$

Regarding the concentrations of $[OH^-]$ and $[H_2O]$, they can be calculated from the relationship 3c. First, by substituting $[H_2O]$ based on the definition of the lump-sum $H_xO$, we obtain
$$
K_w' = \frac{[OH^-]\cdot [H^+]}{[H_xO] - [OH^-]},
$$
which upon rearrangement yields the expression for $[OH^-]$ as a function of the lump-sum $[H_x0]$ and $[H^+]$:
$$
\boxed{~[OH^-] = \frac{K_w'}{K_w' + [H^+]}\cdot [H_xO]~} \qquad\hspace{2cm} (5d)
$$
Finally, by combining this last result with the definition of $H_xO$, we obtain
$$
[H_2O] = [H_xO] - [OH^-] = \left( 1-\frac{K_w'}{K_w' + [H^+]} \right)
\cdot [H_xO],
$$
which yields the expression for $[H_2O]$ as a function of the lump-sum $[H_x0]$ and $[H^+]$:
$$
\boxed{~[H_2O] = \frac{[H^+]}{K_w' + [H^+]}\cdot [H_xO].~} \qquad\hspace{2cm} (5e)
$$

Once the concentrations $[HCO_3^-]$, $[CO_3^{2-}]$ and $[OH^-]$ are calculated from equations 5a--5d based on the values of $[DIC]$, $[H_xO]$ and $[H^+]$, the alkalinity is calculated from the definition:

$$
[ALK] = [HCO_3^-] + 2\cdot [CO_3^{2-}] + [OH^-] - [H^+]
\qquad\hspace{2cm} (5f)
$$

### Key take-home message

Because $[HCO_3^-]$ and $[CO_3^{2-}]$ depend on $[DIC]$ and $[H^+]$ (Eqs. $5a$ and $5c$), and $[OH^-]$ depends on $[H_xO]$ and $[H^+]$ (Eq. $5d$), **equation 5f provides an intimate relationship between $[ALK]$, $[DIC]$, $[H_xO]$ and $[H^+]$** when the carbonate and water species are in equilibrium. Using this relationship, one can calculate 

1. $[ALK]$ if the values of $[DIC]$, $[H_xO]$ and $[H^+]$ are known, or

2. $[DIC]$ if the values of $[H^+]$, $[H_xO]$ and $[ALK]$ are known, or

3. $[H^+]$ if the values of $[DIC]$, $[ALK]$ and $[H_xO]$ are known.

Note that equation $5f$ is highly non-linear; thus these calculations need to be done *numerically*. This contrasts to part I of this reader series, where it was possible to also find an analytical solution. Although this is no longer possible when the number of fast acid-base reactions (e.g., Reactions 1--3) increases, we can proceed thanks to numerical solvers (see next section).

Another important point to note here is the statement 3: $[H^+]$ *can be calculated* if the values of $[DIC]$, $[ALK]$ and $[H_xO]$ are known. This statement is the basis for the *general* approach in $pH$ modeling: one needs to first *identify the appropriate lump-sum species* (e.g., $DIC$, $ALK$ and $H_xO$) and then formulate the *effect of each slow process* on these lump-sum species. For example, in this particular exercise, the effects are: (i) the addition or removal of $DIC$ (Eq. 4a), (ii) conservation of $ALK$ (Eq. 4b), and (iii) conservation of $H_xO$ (Eq. 4c). Then, by solving the differential equations for the lump-sum species numerically, one can quantify the corresponding evolution of $pH$ based on the statement 3, i.e., by numerically solving for $[H^+]$ from Eq. 5f. This approach will be illustrated in the following section. 

## Task 5: Application to atmospheric CO$_2$ exchange 

### Implementation in R

First, we define a function that calculates alkalinity from the known concentrations of $DIC$ and $H_xO$ and $pH$ (equation 5f).

```{r}
solveALK <- function(K1 = 4.44e-07, K2 = 4.67e-11, Kw.prime = 1.79e-16, 
                     HxO = 55.3, DIC, pH){
  
  H      <- 10^(-pH)
  HCO3   <- K1*H  / (H^2 + K1*H + K1*K2) * DIC    # eq 5a
  CO3    <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC    # eq 5c
  OH     <- Kw.prime / (Kw.prime + H) * HxO       # eq 5d
  ALK <- HCO3 + 2*CO3 + OH - H                    # eq 5f
  return(ALK)
}
```

Next, we define a function that calculates $[H^+]$ from the known concentrations of $DIC$, $ALK$ and $H_xO$ (equation 5f). This is done by finding the root of a function defined as the left-hand-side minus the right-hand-side of equation 5f, using a numerical solver implemented in ``uniroot``.

```{r}
solveH <- function(K1 = 4.44e-07, K2 = 4.67e-11, Kw.prime = 1.79e-16,
                   HxO = 55.3, DIC, ALK){
  
  # function whose root has to be sought
  rootFun <- function(H) {

    HCO3    <- K1*H  / (H^2 + K1*H + K1*K2) * DIC    # eq 5a
    CO3     <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC    # eq 5c
    OH      <- Kw.prime / (Kw.prime + H) * HxO       # eq 5d
    ALK.est <- HCO3 + 2*CO3 + OH - H                 # eq 5f
    return(ALK.est - ALK)
  }
  
  # uniroot will find the root    
  h <- uniroot(f = rootFun, lower = 0, upper = 1e-3, tol = 1e-30)
  return(h$root)
}
```

It is always good to test the functions. For pure water with zero $[DIC]$ and $[ALK]$, we expect to get $pH$ of about 7 (i.e., neutral $pH$). This follows from Eqs. 3c and 5f, because at these conditions, we have $[OH^-]=[H^+]$ and $[OH^-]\cdot [H^+] = K_w'\cdot [H_2O]$, which yields $[H^+]=\sqrt{K_w'\cdot [H_xO]}\approx 10^{-7}$ and so $pH=-log_{10}([H^+])=7$. The result is as expected:

```{r}
(pH.neutral  <- -log10(solveH(DIC = 0, ALK = 0)))
```

In contrast, for $[DIC] = 2~mmol~L^{-1}$ and $[ALK] = 2~mmol~L^{-1}$, we expect to get $pH$ of around $8$. Furthermore, when we substitute that $pH$ value to calculate back $[ALK]$, we expect to get the original value. Indeed, the results are as expected:

```{r}
(pH.new  <- -log10(solveH(DIC = 2e-3, ALK = 2e-3)))
(ALK     <-  solveALK(DIC = 2e-3, pH = pH.new))
```

Next, we implement the model function based on the local equilibrium assumption. Note that the equilibrium constants (dissociation constants) are in $mol~L^{-1}$, so all species must be in this unit, too!

```{r}
CO2dissol <- function(t, state, parms) {
  with (as.list(c(state, parms)), {

    # mass balance equations
    dDIC <- RC
    dALK <- 0
    dHxO <- 0

    # calculate the relevant species for the output
    H    <- solveH(K1 = K1, K2 = K2, Kw.prime = Kw.prime, 
                   HxO = HxO, DIC = DIC, ALK = ALK)
    HCO3 <- K1*H  / (H^2 + K1*H + K1*K2) * DIC   # eq 5a
    CO3  <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC   # eq 5c
    OH   <- Kw.prime / (Kw.prime + H) * HxO      # eq 5d

    return(list(c(dDIC, dALK, dHxO), 
                pH = -log10(H), 
                H = H, OH = OH, H2O = HxO - OH,
                H2CO3 = DIC - HCO3 - CO3,
                HCO3 = HCO3,
                CO3  = CO3))
  })
}
```

We run the model for 10 days, using the initial conditions given in the task.

```{r, fig.height=6, fig.width=10, message=FALSE}
require(deSolve)

pars    <- c(RC = 0.1e-3,          # [mol/L/day]
             K1 = 4.44e-07,        # [mol/L]
             K2 = 4.67e-11,        # [mol/L]
             Kw.prime = 1.79e-16)  # [mol/L]

pH.ini  <- 8
DIC.ini <- 2e-3
HxO.ini <- 55.3
ALK.ini <- solveALK(HxO = HxO.ini, DIC = DIC.ini, pH = pH.ini)

yini    <- c(DIC = DIC.ini, ALK = ALK.ini, HxO = HxO.ini)
times   <- seq(from=0, to=10, length.out=100)

out     <- ode(y=yini, times=times, func=CO2dissol, parms=pars)

plot(out, mfrow = c(2,5), las = 1)
```

We see that after 10 days, the addition of $CO_2$ at a rate or `r pars["RC"]` $mol~L^{-1}~d^{-1}$ leads to water acidification, i.e., a decrease in $pH$ from 8 to about 6.6.
