---
title: |
  | Local Equilibrium Chemistry in R --- part II:
  | Two Equilibrium Reactions
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky and Karline Soetaert, Utrecht University"
date: "February 2021"
output:
  pdf_document:
    number_sections: false
  html_document: default
---

# Answers

## Task 1: Mass balance equations of the carbonate system

Assuming that the forward and backward reactions are elementary, we can use the first-order kinetics to describe the rate laws for each reaction. Considering the stoichiometry of Reactions 1 and 2, we obtain the following mass balance equations:

$$
\begin{array}{rclr}
\displaystyle
\frac{d[H_2CO_3]}{dt} & = & R_1 + R_C & \qquad (1a)\\[5mm]
\displaystyle
\frac{d[HCO_3^-]}{dt} & = & -R_1 + R_2 & \qquad (1b) \\[5mm]
\displaystyle
\frac{d[CO_3^{2-}]}{dt} & = & -R_2  & \qquad (1c)\\[5mm]
\displaystyle
\frac{d[H^+]}{dt} & = & -R_1 - R_2 & \qquad (1d)
\end{array}
$$

where the individual rates are defined as the sum of rates of the forward and backward reactions:

$$
R_1 = - k_{1f} \cdot [H_2CO_3] + k_{1b} \cdot [HCO_3^-] \cdot [H^+] 
\qquad\hspace{2cm} (2a)
$$

$$
R_2 = - k_{2f} \cdot [HCO_3^{-}] + k_{2b} \cdot [CO_3^{2-}] \cdot [H^+] 
\qquad\hspace{2cm} (2b)
$$

Note that the rate $R_C$ of the slow process only affects $H_2CO_3$.

## Task 2: Equilibrium conditions

In the absence of the removal process RC, the carbonate species will reach an equilibrium as a result of the reversible reactions 1 and 2. In this equilibrium, the time derivative of each species in the system is zero. Based on equations 1a and 2a, we obtain $R_1=0$, which yields the relationship

$$
K_{1} = \frac{ [HCO_3^-] \cdot [H^+] }{ [H_2CO_3] } \qquad\hspace{2cm} (3a)
$$
Similarly, based on equation 1c and 2b, we obtain $R_2=0$, which yields the relationship
$$
K_{2}  = \frac{ [CO_3^{2-}] \cdot [H^+] }{[HCO_3^-]} 
\qquad\hspace{2cm} (3b)
$$

Here, we defined the dissociation constants $K_1$ and $K_2$ as $K_1 = k_{1f}/k_{1b}$ and $K_2 = k_{2f}/k_{2b}$. If relationships 3a and 3b are true, then the time derivatives in equations 1b and 1d are zero. This means that these two equations do *not* provide any *extra* information for the equilibrium concentrations.

## Task 3: Lump-sum species

By summing equations 1a, 1b and 1c, we obtain 
$$
\frac{d[DIC]}{dt} = R_C \qquad\hspace{2cm} (4a)
$$
Similarly, by summing equation 1b with equation 1c multiplied with a factor of 2, and finally subtracting equation 1d, we obtain
$$
\frac{d[ALK]}{dt} = 0 \qquad\hspace{2cm} (4b)
$$
These equations show that DIC is *added* at a rate of the *slow process* ($R_C$), while ALK is *conserved* (not changing in time). Notably, both DIC and ALK are *not* effected by the fast equilibration reactions.

## Task 4: From lump-sum species to original species

By solving for $[H_2CO_3]$ and $[CO_3^{2-}]$ in equations 3a and 3b, and by substituting the result to the definition of $[DIC]$, we obtain 

$$
[DIC] = [HCO_3^-] + [H^+]\cdot \frac{[HCO_3^-]}{K_{1}} + K_{2}\cdot \frac{ [HCO_3^-]}{[H^+]}
$$

When we bring each term on the right-hand side to the common denominator ($K_1\cdot [H^+]$), we obtain

$$
[DIC] = \frac{K_1\cdot[H^+]\cdot[HCO_3^-]}{K_1\cdot[H^+]} + \frac{ [H^+]^2\cdot[HCO_3^-]}{K_1 \cdot[H^+]} + \frac{ K_1 \cdot K_{2} \cdot[HCO_3^-]}{K_1 \cdot [H^+]}
$$

This equation can easily be solved for $[HCO_3^-]$:

$$
[HCO_3^-]= \frac{K_{1}\cdot [H^+]}{[H^+]\cdot [H^+] + K_{1}\cdot [H^+] +  K_{1}\cdot K_{2}}\cdot [DIC]
\qquad\hspace{2cm} (5a)
$$

Combining this result with equation 3a, we obtain 

$$
[H_2CO_3] = \frac{[H^+]\cdot [H^+]}{[H^+]\cdot [H^+] + K_{1}\cdot [H^+] +  K_{1}\cdot K_{2}}\cdot [DIC]
\qquad\hspace{2cm} (5b)
$$

Similarly, combining the result with equation 3b, we obtain

$$
[CO_3^{2-}] = \frac{K_1\cdot K_2}{[H^+]\cdot [H^+] + K_{1}\cdot [H^+] +  K_{1}\cdot K_{2}}\cdot [DIC]
\qquad\hspace{2cm} (5c)
$$

Once the concentrations $[HCO_3^-]$, $[CO_3^{2-}]$ and $[H^+]$ are known, the carbonate alkalinity is calculated based on the definition:

$$
[ALK] = [HCO_3^-] + 2\cdot [CO_3^{2-}] - [H^+]
\qquad\hspace{2cm} (5d)
$$

Because $[HCO_3^-]$ and $[CO_3^{2-}]$ depend on $[DIC]$ and $[H^+]$ (see 5a and 5c), \bf equation 5d provides an intimate relationship between $[ALK]$, $[DIC]$ and $[H^+]$ \rm when the carbonate species are in an equilibrium. Using this relationship, one can calculate 

* $[ALK]$ if the values of $[DIC]$ and $[H^+]$ are known, or

* $[H^+]$ if the values of $[DIC]$ and $[ALK]$ are known, or

* $[DIC]$ if the values of $[H^+]$ and $[ALK]$ are known.

Because equation 5d is highly non-linear, these calculations are done numerically.

## Task 5: Application to atmospheric CO$_2$ exchange 

### Implementation in R

First, we define a function that calculates carbonate alkalinity from the known concentration of DIC and pH (equation 5d).

```{r}
solveALK <- function(K1 = 1.422e-06, K2 = 1.082e-09, DIC, pH){
  
  H <- 10^{-pH}
  HCO3   <- K1*H /  (H^2 + K1*H + K1*K2) * DIC    # eq 5a
  CO3    <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC    # eq 5c
  ALK    <- HCO3 + 2*CO3 - H                      # eq 5d
  
  return(ALK)
}
```

Second, we define a function that calculates $[H^+]$ from the known carbonate alkalinity and DIC concentration (equation 5d). This is done by finding a root of a function defined as the left-hand-side minus the right-hand-side of equation 5d.

```{r}
solveH <- function(K1 = 1.422e-06, K2 = 1.082e-09, DIC, ALK){
  
  # function whose root has to be sought
  rootFun <- function(H) {
    
    HCO3    <- K1*H /  (H^2 + K1*H + K1*K2) * DIC   # eq 5a
    CO3     <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC   # eq 5c
    ALK.est <- HCO3 + 2*CO3 - H  # right-hand-side of eq. 5d
    
    return(ALK.est - ALK)
  }
  
  # uniroot will find the root    
  h <- uniroot(f = rootFun, lower = 0, upper = 1e-3, tol = 1e-30)
  return(h$root)
}
```

It is always good to test the functions

```{r}
(pH  <- -log10(solveH(DIC = 2e-3, ALK = 2.18e-3)))
(ALK <- solveALK(DIC = 2e-3, pH = 8))
```

The model function that implements the model based on the local equilibrium assumption is defined below. Note that the equilibrium constants (dissociation constants) are in $mol~L^{-1}$, so all species should be in this unit!

```{r}
CO2dissol <- function(t, state, parms) {
  with (as.list(c(state,parms)), {

    # mass balance equations
    dDIC <- RC
    dALK <- 0

    # for output
    H <- solveH(K1 = K1, K2 = K2, DIC = DIC, ALK = ALK)
    HCO3    <- K1*H /  (H^2 + K1*H + K1*K2) * DIC   # eq 5a
    CO3     <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC   # eq 5c
    
    return(list(c(dDIC, dALK), 
                pH = -log10(H), H = H, 
                H2CO3 = DIC - HCO3 - CO3,
                HCO3 = HCO3,
                CO3  = CO3))
  })
}
```

Now we run the model for 10 days, using the initial conditions given in the task.

```{r, fig.height=6, fig.width=10}
require(deSolve)

parms <- c(RC = 0.1e-3,    # [mol/L/day]
           K1 = 1.422e-06, # [mol/L]
           K2 = 1.082e-09) # [mol/L]

pH.ini  <- 8
DIC.ini <- 2e-3
ALK.ini <- solveALK(DIC = DIC.ini, pH = pH.ini)

yini  <- c(DIC = DIC.ini, ALK = ALK.ini)
times <- seq(from=0, to=10, length.out=100)

out   <- ode(y=yini, times=times, func=CO2dissol, parms=parms)

plot(out, mfrow = c(2,4), las = 1)
```

We see that the addition of $CO_2$ at the given rate leads to an acidification of the water from the initial pH of 8 to pH of about 6.3 after 10 days.
