---
title: |
  | Local Equilibrium Chemistry in R --- part II:
  | Three Equilibrium Reactions
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky and Karline Soetaert, Utrecht University"
date: "February 2021"
output:
  pdf_document:
    number_sections: true
  html_document: default
---

```{r child = 'equilibriumHCO3_Q.Rmd'}
```

```{r child = 'equilibriumHCO3_A.Rmd'}
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for
  Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential Equations in R:
  Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL
  http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09

Jean-Pierre Gattuso, Jean-Marie Epitalon, Heloise Lavigne and James Orr (2020). seacarb:
  Seawater Carbonate Chemistry. R package version 3.2.13.
  https://CRAN.R-project.org/package=seacarb
  
Andreas F. Hofmann, Karline Soetaert, Jack J. Middelburg and Filip J. R. Meysman, 2010,
  AquaEnv - An Aquatic Acid-Base Modelling Environment in R.  
  Aquatic Geochemistry DOI 10.1007/s10498-009-9084-1,
  https://CRAN.R-project.org/package=AquaEnv
  
S. J. Hawkes 1995, 
  pKw is almost never 14.0.
  Journal of Chemical Education, 72(9): 799--802,
  DOI: 10.1021/ed072p799