---
title: "RTM course - Units"
author: "Lubos Polerecky and Karline Soetaert"
description: "exercises on working with units"
date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Units

WORK IN PROGRESS

In quantitative problems, units play an important role. In this course, units will be practically 'everywhere'. For example,

- we will often need to *convert between different units* when setting up our models or defining model parameters;

- the first *quality check* of our models will very often include a *check on units*;

In our experience, *most frequent mistakes* made by students when solving quantitative problems occur *during units conversions.* To avoid such mistakes, we recommend that you learn and adopt the following approach.

1. At the beginning of your calculations, *convert all quantities to SI units*. 

   In this course, we will most often use these SI units: m (meter), s (second), mol (mole), kg (kilogram), and K (Kelvin). 

2. Be aware that you can *subtract* or *add* quantities if, and only if, they have the *same units*. 

3. On the other hand, multiplying or dividing quantities with different units is no problem. In this case, the *units* of the different quantities also need to be *multiplied* or *divided*.

## Converting units to SI units

### Prefixes

In practical life, we often use prefixes such as kilo ($10^3$), mega ($10^6$), milli ($10^{-3}$), micro ($10^{-6}$), etc. We do this mostly for convenience, e.g., if we want to avoid reporting too large or too small numbers. 

### Recommended practice

When dealing with quantities in such units, it is best to first convert those prefixes to the corresponding magnitude in the form $10^n$, where $n$ is an integer number. For example, 

- $22~mm \rightarrow 22\times 10^{-3}~m$
- $45~\mu s \rightarrow 45\times 10^{-6}~s$
- $300~\mu g \rightarrow 300\times10^{-6}~ g =$ $300 \times (10^{-9} \times 10^3)~g=$ $300 \times 10^{-9} \times (10^3~g) =$ $300\times 10^{-9} ~kg$. 

**Note**: Although, technically, $k$ (kilo) refers to $10^3$, in the last example we converted $\mu g$ to $kg$, not to $g$. This is because, oddly, *kilogram* (kg) is the SI unit of mass. 

This is important to do when dealing with *derived* SI units, such as Newton (N), Joule (J), etc., which are defined in terms of SI base units. For example, $N= kg\cdot m\cdot s^{-2}$, whereas $J=N\cdot m =  kg\cdot m^2\cdot s^{-2}$. This will be important later, when we get to the multiplication and division of units. 

### Conventional units

In other practical situations, we use non-SI units. For example, we use 'liters' ($L$) when talking about volumes, the 'tonne' ($t$) when talking about heavy objects, 'molars' ($M$) when talking about concentrations of solutes in a liquid, or 'degrees Celsius' ($^\circ C$) when talking about temperature.

We need to be aware of their relationship to the SI units:

- $1~L = 0.001~m^3 = 10^{-3}~m^3$
- $1~t = 1000~kg = 10^3~kg$
- $1~M = 1~mol~L^{-1}$
- $T~(K) = T~(^\circ C) + 273.15$

### Recommended practice

When dealing with quantities in non-SI units, it is best to first convert them to SI units. If there are prefixes, those should be converted to the corresponding magnitude in the form $10^n$, where $n$ is an integer number. For example, 

$$
\begin{array}{rcl}
220~\mu M & = & 220~\times \underbrace{10^{-6}}_{\mu}~\underbrace{mol~L^{-1}}_{M} \\  
& = & 220 \times 10^{-6}~mol~(\underbrace{10^{-3}~m^3}_L)^{-1} \\
& = & 220 \times 10^{-6}~mol~10^{3}~m^{-3} \\
& = & 220 \times (10^{-6}\times 10^3)~mol~m^{-3}\\
& = & 220\times 10^{-3}~mol~m^{-3} \\
& = & 0.22~mol~m^{-3}
\end{array}
$$

## Adding or subtracting quantities

1. Consider the following quantities:

- lengths: $L_1 = 1~m$, $L_2 = 2~m$; time intervals: $T_1 = 10~s$, and $T_2 = 20~s$.

   You *can* add lengths $L_1$ an $L_2$, which yields $L_1+L_2=3~m$. 

   You can also add time intervals $T_1$ and $T_2$, which yields $T_1 + T_2 = 30~s$. 

   However, you *cannot* add or subtract length with a time interval, for example, $L_1$ with $T_1$, or $L_1$ with $T_2$, as the quantities do not have the same unit.

2. Next, consider the following quantities:

- lengths: $L_1 = 1~m$, $L_2 = 150~mm$; time intervals: $T_1 = 1~min$, and $T_2 = 20~s$.

   As before, you should be able to add lengths $L_1$ and $L_2$. However, you can only do so if they have the same units! We use SI units. So, we have $L_1+L_2 = 1~m + 0.15~m = 1.15~m$. 

   Similarly, you can add time intervals $T_1$ and $T_2$ only if they have the same units! We use SI units. So, we have $T_1 + T_2 = 60~s + 20~s = 80~s$. 

   As before, you *cannot* add or subtract lengths and time intervals.

## Test yourself

```{r Organism_Currency, echo=FALSE}
msg <- "Carbon is used both for biomass and energy production, so this is the best model currency."

question("What would be a good model currency for such a model?", type = "learnr_radio",
 answer("C (carbon)", correct=TRUE),
 answer("N (nitrogen)"),
 answer("A mix of N and C"),
 
 incorrect=msg,
 correct=msg)
```

