---
title: "Evaporation and precipitation of water"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky, Utrecht University"
date: "January 2022"
abstract: \noindent If liquid water is surrounded by a gas phase (e.g., atmospheric air), some water molecules will transfer from the liquid phase to the gas phase by a process called evaporation, while some others will transfer back from the gas phase to the liquid phase by a process called precipitation. Generally, the rates of these two processes differ from each other, leading to either net evaporation or net precipitation. However, if the system is enclosed in a fixed volume and its temperature is kept constant, the rates become at some point equal to each other, leading to zero net transfer of water between the two phases. This state corresponds to a thermodynamic equilibrium. In this reader, we will show how to model evaporation and precipitation of water as two dynamic processes occuring at the same time.
output: 
  pdf_document:
    toc: no
  html_document:
    toc: yes
    df_print: paged
  header-includes: 
    \usepackage{float} 
    \floatplacement{figure}{ht}
---

```{r child = 'evaporationH2O_Q.Rmd'}
```

\newpage
```{r child = 'evaporationH2O_A.Rmd'}
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for
  Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential Equations in R:
  Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL
  http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09

Carrier W. H. (1918). “The Temperature of Evaporation,” Trans. Am. Soc. Heat. Vent. Eng., vol. 24, p. 25.