---
title: 'Viruses in the aquatic environment'
subtitle: "Exercises Accompanying the Course Environmental Modelling"
author: "Karline Soetaert and Dries Bonte, Ghent University"
date: "July 2021"
output: 
  pdf_document:
    default
---

# Answers

## Units

Reasonable units for the state variables are

* algal cells: $cells~ml^{-1}$
* viruses: $virions~ml^{-1}$ 

## Mass balances and rate expressions

* Healthy (uninfected) algal cells (*H*) are created by cell division, and die due to natural causes (i.e., independent of the presence of viruses). The net growth of the algal cells is described by the logistic growth model characterized by the rate constant *lambda* (units of $d^{-1}$) and the carrying capacity *k* (units of $cells~ml^{-1}$). 

* Viral infection of algal cells depends on a random encounter between free virions and algae. Thus, the process is proportional to the product of the concentration of healthy cells, $H$, and the concentration of free virions, $V$. The corresponding rate constant, *beta*, has units of $(virions~ml^{-1})^{-1}~d^{-1}$. 

* Death of infected cells (by bursting) is a first-order process characterized by a rate constant *delta* (units of $d^{-1}$). 

* When an infected algal cell bursts, it releases *n* free virions (units of $virion~cell^{-1}$). Conversely, it takes *one* virion to infect a healthy algal cell. Thus, the infection efficiency parameter is $e=1~virion~cell^{-1}$.

* Free virions are not immortal but decay at a first-order rate characterized by the rate constant *c* (units of $d^{-1}$).

Based on these considerations, the mass balance equation for the uninfected (*H*) and infected (*I*) algal cells and for the free virions are:

$$\frac{dH}{dt} = NetAlgalGrowth - Infection$$
$$\frac{dI}{dt} = Infection - Bursting$$
$$\frac{dV}{dt} = n\cdot Bursting  - VirusDeath - e\cdot Infection$$

where

$$NetAlgalGrowth = lambda \cdot H \cdot \left(1 - \frac{H}{k}\right)$$
$$Infection = beta \cdot H \cdot V$$
$$Bursting = delta\cdot I$$
$$VirusDeath = c \cdot V$$

## Parameters

The parameter values are in the following table:

| parameter | Value   | Units     | Description |
|-----------| --------|-----------|-------------|
| lambda    |	1       |$d^{-1}$   | Net growth rate constant of healthy cells  |
| delta 	  | 0.55  	|$d^{-1}$  	| Burst rate constant of infected cells    |
| beta	    | 0.00002	|$(virions~ml^{-1})^{-1}d^{-1}$| Infection rate constant |
| c	        | 5.5	    |$d^{-1}$   | Death rate constant of viruses          |
| k	        | 1e5     |$cells~ml^{-1}$ | Algal carrying capacity             |
| n  	      | 100	    |$virions~cell^{-1}$ | Burst size                        |
| e         | 1       |$virions~cell^{-1}$  | Infection efficiency     |

## R-implementation

Here is the code specifying the model:

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
require(deSolve)  # package with solution methods

# state variables, units = cells/ml (H, I) or virions/ml (V)
state <- c(H = 100, I = 150, V = 50000)

# parameters 
parms <- c(
  lambda = 1,     # [/d] net growth rate constant of healthy algal cells
  delta = 0.55,   # [/d] burst rate constant of infected cells
  beta = 0.00002, # [(virions/ml)^-1 d-1] infection rate constant
  c = 5.5,        # [/d] death rate constant of virus
  k = 1e5,        # [cells/ml] carrying capacity for algal cells
  n = 100,        # [virus/cell] burst size
  e = 1           # [virus/cell] number of viruses required to infect an algal cell
)

# Model function
VirusDynamics <- function(t, state, params) {
  with (as.list(c(state, params)), {

   # Rate expressions
    NetAlgalGrowth <- lambda*H*(1-H/k)  # [cells/ml/d]
    Infection      <- beta * H * V      # [cells/ml/d]
    Bursting       <- delta*I           # [cells/ml/d]
    VirusDeath     <- c*V               # [virions/ml/d]
 
   # Mass balances [cells/ml/d]
    dH <- NetAlgalGrowth - Infection
    dI <- Infection - Bursting
    dV <- n*Bursting - VirusDeath - e*Infection
    
    list(c(dH, dI, dV), Cells = H + I, Bursting = Bursting, 
      Infection = Infection, logV = log10(V), logC = log10(H+I))
  })

}
```

Here are plots of the model solution:

```{r, fig.height = 10, fig.width = 10}
# output time
outtimes <- seq(from = 0, to = 365, length.out = 365*10)  # run for 365 days

# solve this model, using the ode function from deSolve
out <- ode(y = state, parms = parms, func = VirusDynamics, times = outtimes)

plot(out, mfrow = c(3, 3), xlab="time (d)")
plot(out[,c(4,5)], type = "l", main = "phase-plane view", 
      xlab = "Viruses", ylab = "Algae")
```
