---
title: "Methods"
author: "Authors"
date: "4/8/2022"
output: pdf_document
---

# Methods

The Methods section describes the "how". It should explain 

* the theoretical underpinning of the model (state variables, including units, conceptual diagram, chemical equations, if relevant, mass balance equations);
* key assumptions (rate expressions, boundary conditions);
* model implementation in R (model parameters & state variables, model function).

## Equations

When writing chemical or mathematical equations, you can use \LaTeX\ commands within the Rmd file. For example:

* Displayed chemical equation (without a label):
$$CH_2O + O_2 \rightarrow CO_2 + H_2O$$

* Displayed chemical equation (with a label):
\begin{equation}
\label{eq:miner}
CH_2O + O_2 \rightarrow CO_2 + H_2O
\end{equation}

* Displayed mathematical equation (with a label):
\begin{equation}
\label{eq1}
\frac{\partial O_2}{\partial t} = Transport(O_2) + R_{aeration} - R_{miner}
\end{equation}
The label can then be used to reference Eq. \ref{eq1} and Eq. \ref{eq:miner} even if you change the order of equations later on.

## Model implementation in R

In this section you will show, and explain, how you implemented the model in R. As an example, here we include the implementation of the BOD model (available in the ``RTM_1D`` template).

```{r, warning=FALSE, message=FALSE}
require(ReacTran)

# units: time=days, space=meters, amount=moles, concentration=mol/m3

# model grid
Length <- 1000                                 # [m]
N      <- 100                                  # [-] number of boxes
Grid   <- setup.grid.1D(L = Length, N = N)     # grid of N equally-sized boxes

# initial conditions - state variables are defined in the middle of grid cells
O2      <- rep(0.1,   times = N)               # [mol/m3]
BOD     <- rep(0.001, times = N)               # [mol/m3]

# initial state of the system: a vector with all state variables (2*N)
state.ini   <- c(O2, BOD)

# names of the modeled state variables
SVnames <- c("O2", "BOD")

# model parameters
pars <- c(      
  D         = 100,   # [m2/d]      dispersion coefficient (tidal mixing)
  v         = 10,    # [m/d]       advection velocity
  kDecay    = 0.05 , # [/d]        rate constant of BOD decay (first-order process)
  K.O2      = 0.001, # [mol/m3]    half-saturation O2 concentration for BOD decay
  inputBOD  = 10,    # [mol/m2/d]  BOD input rate upstream
  BODdown   = 0.1,   # [mol/m3]    BOD concentration downstream
  O2up      = 0.25,  # [mol/m3]    O2 concentration upstream
  satO2     = 0.3,   # [mol/m3]    saturation concentration of O2 (i.e., solubility)
  kAeration = 0.1    # [/d]        rate constant for air-water O2 exchange
)

# Model function
BOD1D <- function(t, state, parms) {  # state is a long vector, at time t
  with (as.list(parms),{
    
  # The vectors of the state variables O2 and BOD are
  # "extracted" from the LONG vector state passed to the function as input.
    O2  <- state[ (0*N+1) : (1*N) ]   # first N elements for O2
    BOD <- state[ (1*N+1) : (2*N) ]   # second N elements for BOD

  # Transport - tran.1D approximates the spatial derivatives
  # note: for O2: zero-gradient boundary downstream (default)
    tranO2  <- tran.1D(C    = O2,
                       C.up = O2up,         # imposed conc upstream, 
                       D = D, v = v,        # dispersion, advection 
                       dx = Grid)           # Grid

    tranBOD <- tran.1D(C       = BOD,
                       flux.up = inputBOD,  # imposed flux upstream
                       C.down  = BODdown,   # imposed conc downstream
                       D = D, v = v,        # dispersion, advection
                       dx = Grid)           # Grid

  # rate expressions [mol/m3/d] - values in the middle of grid cells
    Decay    <- kDecay * BOD * O2/(O2+K.O2) # BOD decay, limited by O2
    Aeration <- kAeration * (satO2-O2)      # air-water exchange of O2

  # Time-derivatives: dC/dt = transport + production - consumption [mol/m3/d]
    dO2.dt    <- tranO2$dC  + Aeration - Decay 
    dBOD.dt   <- tranBOD$dC            - Decay   

  # return vector of time-derivatives and ordinary variables as a list     
    return(list(c(dO2.dt, dBOD.dt),  # time-derivatives 
                                     # (the same order as state variables!!)
  # additional output:
    
    # process rates along the domain (1D vector)
    Decay           = Decay,                 # mol/m3/d
    Aeration        = Aeration,              # mol/m3/d
    
    # mean process rates (a number)
    MeanDecay       = mean(Decay),           # mol/m3/d
    MeanAeration    = mean(Aeration),        # mol/m3/d
    
    # rates integrated along the domain (for budgetting) 
    TotalDecay      = sum(Decay*Grid$dx),    # mol/m2/d
    TotalAeration   = sum(Aeration*Grid$dx), # mol/m2/d
    
    # fluxes at domain boundaries (for budgetting)
    BODinflux = tranBOD$flux.up,     # BOD flux INTO the system upstream,     mol/m2/d
    BODefflux = tranBOD$flux.down,   # BOD flux OUT of the system downstream, mol/m2/d       
    O2influx  = tranO2$flux.up,      # O2 flux INTO the system upstream,      mol/m2/d
    O2efflux  = tranO2$flux.down))   # O2 flux OUT of the system downstream,  mol/m2/d
  })
}
```

## Modeled scenarios

In this section you can also include the *calculation* (not yet plotting) of model outputs for specific model scenarios. 

For example, here we find a steady-state solution for the default model parameters.

```{r}
std <- steady.1D(y = state.ini, parms = pars, func = BOD1D, 
          positive = TRUE, names = SVnames, nspec = length(SVnames), dimens = N,
          atol = 1e-10, rtol = 1e-10) 
```

In contrast, here we solve the model assuming no aeration.

```{r}
p2 <- pars
p2["kAeration"] <- 0
std2 <- steady.1D(y = state.ini, parms = p2, func = BOD1D, 
          positive = TRUE, names = SVnames, nspec = length(SVnames), dimens = N,
          atol = 1e-10, rtol = 1e-10)
```

We also generate the content of tables. We do not display them just yet, but we wait until the Results section.

```{r}
toselect <- c("TotalDecay", "TotalAeration", "BODinflux", "BODefflux",
              "O2influx", "O2efflux")
BUDGET <- data.frame(default     = unlist(std[toselect]), 
                     no_aeration = unlist(std2[toselect]))
```

This can go on until you clarify all methodological aspects covered in your project.