---
title: "Results"
author: "Authors"
date: "4/8/2022"
output: pdf_document
---

# Results

The Results section should summarize your key findings supported by graphs and/or tables.

## Including variables generated in Methods

When knitting this file independently of the ``methods.Rmd`` file, you need to ensure that the values of the variables generated by the R-code in the ``methods.Rmd`` file are known here. You can do this by including the following R-chunk (enclosed between the opening and closing $`\,`\,`$) somewhere at the beginning of the ``results.Rmd`` file:

```
{r, message=FALSE, echo=FALSE, eval=TRUE}
res <- knitr::knit_child('methods.Rmd', quiet = TRUE)
```

```{r, message=FALSE, echo=FALSE, eval=FALSE}
res <- knitr::knit_child('methods.Rmd', quiet = TRUE)
```

Once you have finished editing the ``results.Rmd`` file and are ready to knit the "master" file of your project, you can omit knitting of the ``methods.Rmd`` child here by using ``eval=FALSE`` in the above R-chunk:

```
{r, message=FALSE, echo=FALSE, eval=FALSE}
res <- knitr::knit_child('methods.Rmd', quiet = TRUE)
```

## Plotting graphs

It is recommended to use *one* R-chunk for plotting one figure (plot). In this way, you can individually optimize the size of the figure and the corresponding figure caption.

Note that if you specify ``fig.cap``, the figure will become a *floating* object and will have a reference number. How this floating object is placed within your final document can be prioritized in the yaml header of your Rmd file. Here, we use
```
    \usepackage{float} 
    \floatplacement{figure}{ht}
```
which means that the first preference is to place the figure in the same place where the figure is generated (``h``, meaning *here*), while the second preference is to place it at the top of the next page (``t``, meaning *top*). These are \LaTeX-specific tricks.

Note that there is no need to display the R-code that is used to generate the graphs, as this code is typically quite long but not overly informative. This is achieved by using the ``echo=FALSE`` flag in the corresponding R-chunk. 

```{r, echo=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap="A comparison of model outputs for a default scenario and for a scenario without aeration."}
plot(std, std2, grid = Grid$x.mid, 
     which=c("O2", "BOD", "Decay", "Aeration"),
     mfrow=c(2,2), xlab = "distance (m)",
     ylab =c("molO2/m3","molBOD/m3","molBOD/m3/d", "molO2/m3/d"))
legend("topright", legend=c("default","no aeration"),
       lty=1:2, col=1:2, bty = "n" )
```

## Reporting values in tables

Tables generated in the Methods section can be nicely typset using the ``kable`` function from the ``knitr`` package.

```{r}
knitr::kable(BUDGET, digits = 2, caption = "Budgets for the different model scenarios.")
```

