---
title: 'Plant Coexistence and Herbivory'
subtitle: "Exercises Accompanying the Course Environmental Modelling"
author: "Dries Bonte and Karline Soetaert, Ghent University"
date: "July 2021"
output: 
  pdf_document:
    default
---

# Answers

It is most efficient to create a model that can answer all questions at once. We do this by adding a number of extra parameters:

* rEmigration, the herbivore emigration rate constant, set to 0 if there is no emigration (default).
* rInfect, the nematode infection rate constant, set to 0 if there is no infection (default).

## R implementation

```{r, message=FALSE}

require(deSolve)  # package with solution methods

# state variables, Units: g/m2
state <- c(GRASS = 50, IRIS = 2, HERB = 0.2)

# parameters
parms <- c(
  ktot        = 200,   # [g/m2] carrying capacity of grass and unpalatable Iris
  assEff      = 0.35,  # [-] efficiency at which cattle incorporate biomass into body mass
  ks          = 50,    # [g/m2] half-saturation constant for functional response of grazing on grass
  rGraz       = 0.17,  # [/d] grazing rate constant of herbivores
  rGrowG      = 0.05,  # [/d] growth rate constant grass
  rGrowI      = 0.04,  # [/d] growth rate constant Iris
  rResp       = 0.018, # [/d] basal metabolic rate constant 
  rEmigration = 0,     # [/d] emigration rate constant
  rInfect     = 0,     # [/(g/m2)/d] nematode infection rate constant
  inhibct     = 0      # [/(g/m2)] strength of inhibition by Iris 
)

# Model function
GrassLand <- function(t, state, params) {
  with (as.list(c(state, params)), {

   # Rate expressions [g/m2/d]
    TotBiom     <- GRASS+IRIS
  
   # plants
    GrassGrowth <- rGrowG * GRASS * (1-(TotBiom/ktot))
    IrisGrowth  <- rGrowI * IRIS  * (1-(IRIS   /ktot))
    Infection   <- rInfect *IRIS*IRIS
    
   # herbivores  
    Grazing     <- rGraz * HERB * GRASS/(GRASS+ks) *exp(-inhibct*IRIS)
    Emigration  <- rEmigration * HERB*HERB  #(density dependence)
    BasalResp   <- rResp * HERB

   # Mass balances [g/m2/d]
    dGRASS     <- GrassGrowth - Grazing
    dIRIS      <- IrisGrowth  - Infection
    dHERB      <- assEff*Grazing - Emigration - BasalResp

    list( c(dGRASS, dIRIS, dHERB), 
            total = TotBiom)
  })

}
``` 

Now we solve the model for various conditions: first the default run, then a run with IRIS-dependent grazing, with nematode infection, and finally two runs with emigration. We plot all results in one figure. 

```{r, fig.height = 8, fig.width = 8}
# output time
outtimes <- seq(from = 0, to = 3650, length.out = 1000)  

# solve this model, using the ode function from deSolve
out <- ode(y = state, parms = parms, func = GrassLand, times = outtimes)

# iris-dependen grazing
parms2 <- parms
parms2["inhibct"] <- 1
out2 <- ode(y = state, parms = parms2, func = GrassLand, times = outtimes)

# Nematode infection
parms3 <- parms2
parms3["rInfect"] <- 0.001
out3 <- ode(y = state, parms = parms3, func = GrassLand, times = outtimes)

# Emigration
parms4 <- parms3
parms4["rEmigration"] <- 0.0001
out4 <- ode(y = state, parms = parms4, func = GrassLand, times = outtimes)

# Emigration + infection, but no inhibition of grazing by Iris
parms5 <- parms4
parms5["inhibct"] <- 0
out5 <- ode(y = state, parms = parms5, func = GrassLand, times = outtimes)

# plot all at once
plot(out, out2, out3, out4, out5, type = "l", lwd = 2, lty = 1, xlab="time (d)")
legend ("bottomright", legend = 1:5, title = "run", 
        col = 1:5, lwd = 2, lty =1)
```
