---
title: "RTM course - R code for a reaction-transport model in porous media"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on formulating reaction-transport models in R"
date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

This tutorial series explains the basics of reaction-transport modeling in R. In this fifth part, you will learn how reaction-transport models are *implemented in R* using the R-package ``ReacTran``. It is recommended to be fully acquainted with the material covered in the first four parts of this tutorial series (``transportProcesses``, ``transportFluxes``, ``transportPorous``, and ``transportBoundaries``) before proceeding with this last part.

The first video explains the overall *structure* of a 1D reaction-transport model.

### Video timeline
* 00:00 Introduction
* 00:32 "Skeleton" of a ReacTran R-code
* 01:25 Step 1 - Problem definition
* 02:56 Step 2 - Rate expressions and mass balances
* 04:13 Step 3 - Environmental settings
* 05:46 Step 4.1 - Implementation in R - definition of the grid
* 06:38 Step 4.2 - Definition of spatially variable model parameters
* 08:00 Step 4.3 - Definition of other parameters
* 08:26 Step 4.4 - Definition of state variables
* 09:52 Step 4.5 - Definition of the model function 
* 10:49 Step 4.6 - Transport terms (incl. boundary conditions)
* 12:48 Step 4.7 - Reaction terms
* 13:32 Step 4.8 - Time derivatives
* 15:04 Step 4.9 - Define output of the model function
* 16:06 Step 4.10 - Calculate steady state solution
* 16:44 Step 4.11 - Plot solution
* 17:07 Step 4.12 - Construct budget
* 17:35 Step 4.13 - Calculate dynamic solution
* 18:06 Conclusion

![](https://www.youtube.com/watch?v=NXAnPCZUCyE&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=14)

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>

The second video gives a brief overview of the development of a 2D reaction-transport model.

![](https://www.youtube.com/watch?v=1-FqyEHv7Jk)

## Sediment carbon dynamics

The R-code below implements simple dynamics of particulate organic carbon (POC) in a fully saturated sediment. The code is based on the following **differential equation**:

$$
\frac{\partial C}{\partial t} = \frac{1}{1-\phi_x} \cdot \frac{\partial}{\partial x}\left[(1-\phi_x) \cdot D_b \cdot \frac{ \partial C}{\partial x}\right] -
\frac{1}{1-\phi_x} \cdot \frac{ \partial }{\partial x} \left[(1-\phi_x) \cdot v \cdot C\right]
- r \cdot C
$$
where 

  * $C$ is the POC concentration in $mol~C~m^{-3}_S$ ($S$ refers to the sediment's solid phase),
  * $x$ is sediment depth in $m$, 
  * $t$ is time in $d$ (days), 
  * $\phi_x$ is the sediment porosity, thus $1-\phi_x$ is the corresponding solid volume fraction of the sediment (both are, in general, depth-dependent)
  * $D_b$ is the bioturbation mixing coefficient in $m^2~d^{-1}$, 
  * $v$ is the advective velocity in $m~d^{-1}$, equal to the sediment accretion velocity, and 
  * $r$ is the rate constant in $d^{-1}$ describing POC mineralization (POC mineralization is assumed to follow first-order kinetics).

The **boundary conditions** are as follows:

  * POC (deposition) flux at the sediment-water interface is $1~mmol~C~m^{-2}~d^{-1}$, i.e., 
  
    $(1-\phi_x) D_b \frac{ \partial C}{\partial x}|_{x=0} = 1\times 10^{-3}~mol~m^{-2}~d^{-1}$.

  * Zero-gradient deep in the sediment, i.e., 
  
    $\left.\frac{\partial C}{\partial x}\right|_{x\rightarrow\infty}= 0$.

Values of the **model parameters** are as follows (all are assumed to be depth-independent):

* $r = 0.01~d^{-1}$
* $\phi = 0.8$
* $D_b = 1\times 10^{-6}~m^2~d^{-1}$
* $v = 2\times 10^{-6}~m~d^{-1}$

Our **aim** is to find a steady-state solution of the model in a sediment column with a depth of $L=20~cm$.

To do that, we define the following code, which specifies 

  * the model *grid*,
  * the model *parameters*,
  * the model *function*, which returns the **time-derivatives**.
  
Read the code carefully to recognize and understand the corresponding parts.

```{r Cex-setup, echo = TRUE, message = FALSE}
# load package with solution methods (includes deSolve, rootSolve)
require(ReacTran)

# UNITS: space: m, time: d, concentration: mol/m3

# spatial domain: total length of 0.2m, 200 grid cells
Length    <- 0.20   # [m]
N         <- 200

# model GRID with an exponentially increasing grid size, starting from 0.05 cm
Grid      <- setup.grid.1D(L = Length, N = N, dx.1 = 0.05/100)

# model PARAMETERS
pars <- c(
 depoPOC = 0.001, # [mol/m2 BULK/d]  POC flux at SWI
 r       = 0.01,  # [/d]             POC mineralization rate constant
 por     = 0.8,   # [-]              sediment porosity (depth-independent)
 Db      = 1e-6,  # [m2/d]           sediment bioturbation (mixing) coefficient
 v       = 2e-6   # [m/d]            sediment accretion velocity
)

# model FUNCTION, returning time derivatives
Cmodel <- function (t, C, pars)   # C is a LONG vector
{
  with (as.list(pars),{ 
    
    # transport of POC, VF = solid volume fraction = 1-porosity!
    Tran.C <- tran.1D(C = C, flux.up = depoPOC, # upper boundary: flux
                      dx = Grid, VF = 1-por,    # grid, volume fraction
                      D = Db, v = v)            # mixing coef., advection velocity
 
    # reaction rate
    Miner  <- r * C                             # [mol/m3 SOLID/d] 

    # time derivative
    dC.dt  <- Tran.C$dC - Miner                 # mass balance
    
    TotalMiner <- sum(Miner*Grid$dx*(1-por))    # depth-integrated mineralization rate
  
    return(list(c(dC.dt),                       # vector of time-derivatives
          TotalMiner = TotalMiner))             # [molC/m2 BULK/d]
 })
}
```

```{r, echo = TRUE, message = FALSE}
<<Cex-setup>>
```

### TASK

In the code below, the steady-state solutions are found for two values of the parameter $r$. Add another model run for $r=0.001~d^{-1}$, and plot all three results in one graph.

```{r Cex, exercise=TRUE, fig.height=4, fig.width=4, exercise.lines = 15}
state.ini <- runif(n=N)   # initial guess of the state variables in each grid box

# run with default model parameter values
std0 <- steady.1D (y=state.ini, func=Cmodel, parms=pars, 
                   names="POC", nspec=1, dimens=N, 
                   positive=TRUE) # search for a solution with positive values! 

# run with a different value of r (0.005 d-1)
p1       <- pars
p1["r"]  <- 0.005
std1 <- steady.1D (y=state.ini, func=Cmodel, parms=p1, 
                   names="POC", nspec=1, dimens=N,
                   positive=TRUE)

###### ADD YOUR CODE HERE #####
#std2 <-

# plot results
plot(std0, std1, std2, grid=Grid$x.mid, xyswap=TRUE, lty=1, las=1,
     ylab="depth (m)", xlab="POC (molC/m3 solid)")
# add legend
legend("bottomright", col=1:3, lty=1,
       title="r", legend=c(pars["r"], p1["r"], p2["r"]))
```

## Adding DIC

DIC is produced as a result of POC mineralization. Your task now is to **add** DIC as another state variable to the model. Note that while POC is a solid substance, DIC is a solute.

  * Don't forget to convert from $m^3$ of solid ($m^3_S$) to $m^3$ of liquid ($m^3_L$), where appropriate.

  * Remember that the volume fractions for solutes and solid substances differ from each other.
  
  * Assume that the effective sediment diffusion coefficient of DIC is $5\times 10^{-5}~m^2~d^{-1}$ and the bottom water DIC concentration is $2~mol~m^{-3}$. Note that these two values have already been added to the vector of model parameters.

Expand the following code to include DIC dynamics.

```{r DICdia, exercise = TRUE, exercise.lines = 30}
require(ReacTran)

Length    <- 0.20   # sediment length [m]
N         <- 200    # number of grid cells
Grid      <- setup.grid.1D(L = Length, N = N, dx.1 = 0.05/100)

pars <- c(
 Db      = 1e-6,   # [m2/d]     sediment bioturbation coefficient (mixing)
 v       = 2e-6,   # [m/d]      sediment accretion velocity
 por     = 0.8,    # [-]        sediment porosity (constant)
 r       = 0.01,   # [/d]       POC consumption rate constant
 depoPOC = 1e-3,   # [mol/m2/d] POC flux at SWI
 bwDIC   = 2,      # [mol/m3]   NEW: bottom water concentration DIC
 diffDIC = 5e-5    # [m2/d]     NEW: effective diffusion coefficient for DIC
)

DICmodel <- function (t, state, pars)   # y is a LONG vector
{
  with (as.list(pars),{
    # extract POC from the long input vector state
    POC    <- state[1:N]         # first N values: POC
    # NEW: extract DIC from the long input vector state
    DIC    <- state[(  ):(  )]   # next  N values: DIC
    
    # POC: transport, reaction, mass balance
    Tran.POC <- tran.1D(C = POC, flux.up = depoPOC, # upper boundary: flux
                      dx=Grid, VF=1-por,            # grid, volume fraction
                      D=Db, v=v)                    # bioturbation, advection
    Miner    <- r * POC                # reaction rate, [mol/m3 SOLID/d]
    dPOC.dt  <- Tran.POC$dC - Miner    # time-derivative for POC, [mol/m3 SOLID/d]

    # DIC: transport, mass balance
    ##### FILL IN THE CORRECT PARAMETERS HERE #####
    Tran.DIC <- tran.1D(C = DIC, C.up = ,  # upper boundary: concentration
                      dx=Grid, VF= ,       # grid, volume fraction
                      D= , v= )            # diffusion, advection
    ##### SPECIFY THE CORRECT MASS BALANCE for DIC #####
    dDIC.dt <-    # time-derivative for DIC, [mol/m3 LIQUID/d]
    
    TotalMiner <- sum(Miner*Grid$dx*(1-por)) # depth-integrated rate
    return(list(c(dPOC.dt, dDIC.dt), TotalMiner = TotalMiner)) 
 })
}

state.ini <- runif(n=2*N)
std <- steady.1D (y=state.ini, func=DICmodel, parms=pars, 
                  names=c("POC","DIC"), nspec=2, dimens=N, positive = TRUE) 
plot(std, grid=Grid$x.mid, xyswap=TRUE, lty=1, las=1,
     ylab="depth (m)", xlab=c("molC/m3 SOLID", "molC/m3 LIQUID"))
```

```{r DICdia-solution}
require(ReacTran)
Length    <- 0.20   # sediment length [m]
N         <- 200    # number of grid cells
Grid      <- setup.grid.1D(L = Length, N = N, dx.1 = 0.05/100)

pars <- c(
 Db      = 1e-6,   # [m2/d]     sediment bioturbation coefficient (mixing)
 v       = 2e-6,   # [m/d]      sediment accretion velocity
 por     = 0.8,    # [-]        sediment porosity (constant)
 r       = 0.01,   # [/d]       POC consumption rate constant
 depoPOC = 1e-3,   # [mol/m2/d] POC flux at SWI
 bwDIC   = 2,      # [mol/m3]   NEW: bottom water concentration DIC
 diffDIC = 5e-5    # [m2/d]     NEW: effective diffusion coefficient for DIC
)

DICmodel <- function (t, state, pars)   # state is a LONG vector
{
  with (as.list(pars),{
    # NEW code: extract POC and DIC from the long input vector state
    POC    <- state[   1  :    N ]   # first N values: POC
    DIC    <- state[(N+1) : (2*N)]   # next  N values: DIC
    
    # POC: transport, reaction, mass balance
    Tran.POC   <- tran.1D(C = POC, flux.up = depoPOC, # upper boundary: flux
                      dx=Grid, VF=1-por,              # grid, volume fraction
                      D=Db, v=v)                      # mixing, advection
    Miner   <- r * POC                # reaction rate, [mol/m3 SOLID/d]
    dPOC.dt <- Tran.POC$dC - Miner    # time-derivative for POC, [mol/m3 SOLID/d]

    # DIC: transport, mass balance
    Tran.DIC <- tran.1D(C = DIC, C.up = bwDIC,  # upper boundary: concentration
                      dx=Grid, VF=por,          # grid, volume fraction
                      D=diffDIC, v=v)           # diffusion, advection
    dDIC.dt  <- Tran.DIC$dC + Miner*(1-por)/por # time-derivative for DIC, [mol/m3 LIQUID/d]
    
    TotalMiner <- sum(Miner*Grid$dx*(1-por)) # depth-integrated rate
    return(list(c(dPOC.dt, dDIC.dt), TotalMiner = TotalMiner)) 
 })
}

# find steady state and plot it
state.ini <- runif(n=2*N)
std <- steady.1D (y=state.ini, func=DICmodel, parms=pars, 
                  names=c("POC","DIC"), nspec=2, dimens=N, positive = TRUE) 
plot(std, grid=Grid$x.mid, xyswap=TRUE, lty=1, las=1,
     ylab="depth (m)", xlab=c("molC/m3 SOLID", "molC/m3 LIQUID"))
```

## More exercises

If you finished this tutorial but some of your answers were incorrect, you are encouraged to do the quiz again. You can do this by clicking on <span style="color:red">**Start Over**</span> at the bottom of the menu to the left.

To learn more about 1D reaction-transport modeling in R, try to solve the following exercises. Note that the level of their complexity progressively increases.

```
RTMexercise("estuaryAnoxia", output="PDF")
RTMexercise("Pdiagenesis", output="PDF")
RTMexercise("diagenesis", output="PDF")
RTMexercise("pHprofiles", output="PDF")
```

## References

R Core Team (2020). R: A language and environment for statistical computing. R
  Foundation for Statistical Computing, Vienna, Austria. URL
  https://www.R-project.org/.

Soetaert, Karline and Meysman, Filip (2012). Reactive transport in aquatic ecosystems: Rapid model prototyping in the open source software R. Environmental Modelling & Software, 32, 49-60.
  
Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.

The 2D SIR-model video created by Anna van der Kaaden. It is part of the youtube playlist available at https://www.youtube.com/playlist?list=PL80j3_tiZjWnljuE_j7pH7zNrMD6guHj1

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>
