---
title: "Modelling carbon flow during culturing of an aerobic methanotroph"
author: "Lubos Polerecky"
date: "05-04-2022"
abstract: "\\noindent Model used for the analysis of data presented in the paper *Methanotrophy by a Mycobacterium species that dominates a cave microbial ecosystem* by van Spanning et al. (2022). Data acquired by Paul Bodelier."
output:
  pdf_document:
    toc: no
  word_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Choice of data and model parameters

```{r}
dataset <- 3
d       <- c(dataset==1, dataset==2, dataset==3)
```

# All data sets and default model parameters

## Data set 1

```{r}
# experimental data, set1
t.exp    <- c(0, 22, 94, 166, 262, 358, 430, 526)
CH4.exp  <- c(22.18, 21.74, 20.08, 18.46, 14.08, 11.31, 10.20, 8.91)
CO2.exp  <- c(2.95, 4.51, 10.10, 12.10, 13.03, 14.04, 17.35, 17.95)
xCO2.exp <- c(NA, 12.90, 17.65, 21.60, 24.83, 26.26, 27.21, 28.10)*1e-2
DATA1     <- data.frame(time = t.exp, CH4 = CH4.exp, CO2 = CO2.exp, xCO2 = xCO2.exp)

# default model parameters
pars1 <- c(
  CH4.ini    = 22.18,   # all values in umol/bottle
  CO2.ini    = 2.95,
  Cdead.ini  = 1.15*8.69-2.95,
  Clive.ini  = 1,
  rMO        = 0.0274,  # CH4 removal rate (umol/bottle/h),
  kMin       = 3*3/530, # mineralization of dead biomass (/h)
  eff        = 0.45,    # assimilation efficiency
  xCH4.src   = 0.5,     #0.41,
  xCdead.src = 0.2,     #0.23,
  xCO2.ini   = 0.01,
  xClive.ini = 0.01
)
```

## Data set 2

```{r}
# experimental data, set2
t.exp    <- c(0, 22, 94, 166, 262, 358, 430, 526)
CH4.exp  <- c(23.80, 23.00, 20.28, 19.10, 14.89, 11.98, 10.73, 9.38)
CO2.exp  <- c(2.00, 4.87, 9.67, 12.20, 14.10, 16.05, 18.87, 21.71)
xCO2.exp <- c(NA, 15.95, 16.37, 19.70, 23.30, 23.86, 23.56, 23.22)*1e-2
DATA2     <- data.frame(time = t.exp, CH4 = CH4.exp, CO2 = CO2.exp, xCO2 = xCO2.exp)

# default model parameters
pars2 <- c(
  CH4.ini    = 23.80,   # all values in umol/bottle
  CO2.ini    = 2.0,
  Cdead.ini  = 6.2,
  Clive.ini  = 1,
  rMO        = 0.031,  # CH4 removal rate (umol/bottle/h),
  kMin       = 3*3/530, # mineralization of dead biomass (/h)
  eff        = 0.2,    # assimilation efficiency
  xCH4.src   = 0.26,     #0.41,
  xCdead.src = 0.27,     #0.23,
  xCO2.ini   = 0.01,
  xClive.ini = 0.01
)
```

## Data set 3

```{r}
# experimental data, set3
t.exp    <- c(0, 22, 94, 166, 262, 358, 430, 526)
CH4.exp  <- c(23.66, 23.48, 21.22, 19.77, 16.41, 14.58, 13.43, 10.87)
CO2.exp  <- c(2.96, 3.74, 11.86, 15.38, 17.68, 20.69, 23.54, 24.17)
xCO2.exp <- c(NA, 9.46, 10.40, 13.81, 16.32, 17.62, 18.24, 18.91)*1e-2
DATA3     <- data.frame(time = t.exp, CH4 = CH4.exp, CO2 = CO2.exp, xCO2 = xCO2.exp)

# default model parameters
pars3 <- c(
  CH4.ini    = 23.65,     # all values in umol/bottle
  CO2.ini    = 2.96,
  Cdead.ini  = 11.63-2.96,
  Clive.ini  = 1,         # arbitrary
  rMO        = 0.0245,    # CH4 removal rate (umol/bottle/h),
  kMin       = 3*3/530,   # mineralization of dead biomass (/h)
  eff        = 0,         # assimilation efficiency
  xCH4.src   = 0.278,
  xCdead.src = 0.13,
  xCO2.ini   = 0.01,
  xClive.ini = 0.01
)
```

# Model function

```{r}
C13model <-function(t, state, parms) { 
  with (as.list(c(state, parms)),{
 
  xCH4   <- CH4_C13/CH4
  xCO2   <- CO2_C13/CO2
  xCdead <- Cdead_C13/Cdead
  xClive <- Clive_C13/Clive

  rMin  <- kMin*Cdead # mineralization of dead Corg

  # total C balances
  dCH4.dt    <- -rMO
  dCO2.dt    <- (1-eff)*rMO + rMin
  dCdead.dt  <- -rMin
  dClive.dt  <- eff*rMO
  
  # 13C balances
  dCH4_C13.dt    <- -rMO * xCH4.src
  dCO2_C13.dt    <- (1-eff)*rMO * xCH4.src + rMin * xCdead.src
  dCdead_C13.dt  <- -rMin * xCdead.src
  dClive_C13.dt  <- eff*rMO * xCH4.src
  
  list(
    c(dCH4.dt,   dCH4_C13.dt,   dCO2.dt,   dCO2_C13.dt, 
      dCdead.dt, dCdead_C13.dt, dClive.dt, dClive_C13.dt),
    xCH4 = xCH4, xCO2 = xCO2, xCdead = xCdead, xClive = xClive
      )
  })
}

# function for calculating initial state from model parameters
state.ini <- function(parms){
  with(as.list(parms), {
  c(CH4   = CH4.ini,   CH4_C13   = CH4.ini*xCH4.src,
    CO2   = CO2.ini,   CO2_C13   = CO2.ini*xCO2.ini,
    Cdead = Cdead.ini, Cdead_C13 = Cdead.ini*xCdead.src,
    Clive = Clive.ini, Clive_C13 = Clive.ini*xClive.ini
  )})
}
```

# Selected data set: L+`r paste(which(d))`

```{r, echo=FALSE}
if (d[1]){ DATA <- DATA1; pars <- pars1 }
if (d[2]){ DATA <- DATA2; pars <- pars2 }
if (d[3]){ DATA <- DATA3; pars <- pars3 }
```

## Model parameters for sensitivity analysis

```{r, eval=d[1], echo=d[1]}
p0 <- p1 <- p2 <- p3 <- pars
p1["Cdead.ini"]  <- 0
p1["xCH4.src"]   <- 0.4
p2["eff"]        <- 0.0
p2["xCH4.src"]   <- 0.38
p3["xCH4.src"]   <- 0.4
p3["xCdead.src"] <- 0.26
```

```{r, eval=d[2], echo=d[2]}
p0 <- p1 <- p2 <- p3 <- pars
p1["Cdead.ini"]  <- 0
p1["xCH4.src"]   <- 0.28
p2["eff"]        <- 0.0
p3["xCH4.src"]   <- 0.5
```

```{r, eval=d[3], echo=d[3]}
p0 <- p1 <- p2 <- p3 <- pars
p1["Cdead.ini"] <- 0
p1["xCH4.src"]  <- 0.23
p2["eff"]       <- 0.7
p2["xCH4.src"]  <- 0.5
p3["xCH4.src"]  <- 0.5
```

## Solve the model

```{r, message=FALSE}
require(deSolve)
outtimes <- seq(from = 0, to = 530, length.out = 100)
out0 <- ode(y = state.ini(p0), parms = p0, func = C13model, times = outtimes)
out1 <- ode(y = state.ini(p1), parms = p1, func = C13model, times = outtimes)
out2 <- ode(y = state.ini(p2), parms = p2, func = C13model, times = outtimes)
out3 <- ode(y = state.ini(p3), parms = p3, func = C13model, times = outtimes)
```

## Full model results

```{r, echo=FALSE, fig.height = 7, fig.width = 8, fig.align='center'}
# display complete output (for debugging)
plot(out0, out1, out2, out3,
     which = c("CH4", "CO2", "Cdead", "Clive",
               "CH4_C13", "CO2_C13", "Cdead_C13", "Clive_C13",
               "xCH4", "xCO2", "xCdead", "xClive"),
     xlab="time (h)", lty=1, lwd=c(1,2,1,2), mfrow=c(3,4))
```

## Model results with experimental data

```{r, fig.height = 3, fig.width = 8, fig.align='center', echo=FALSE}
# display exp. data and model predictions
plot(out0, out1, out2, out3, 
     obs = DATA,
     obspar = list(pch=16, col=which(d)),
     mfrow=c(1,3), lty=c(1,2,1,2), lwd=c(1,2,1,2))
legend("bottomright", legend=0:3, lty=c(1,2,1,2), lwd=c(1,2,1,2), col=1:4)
```

# All experimental data and model results

```{r, message=FALSE, echo=FALSE}
out1 <- ode(y = state.ini(pars1), parms = pars1, func = C13model, times = outtimes)
out2 <- ode(y = state.ini(pars2), parms = pars2, func = C13model, times = outtimes)
out3 <- ode(y = state.ini(pars3), parms = pars3, func = C13model, times = outtimes)
pars1_50 <- pars1; pars1_50["xCH4.src"] <- 0.5
pars2_50 <- pars2; pars2_50["xCH4.src"] <- 0.5
pars3_50 <- pars3; pars3_50["xCH4.src"] <- 0.5
out1_50 <- ode(y = state.ini(pars1_50), parms = pars1_50, func = C13model, times = outtimes)
out2_50 <- ode(y = state.ini(pars2_50), parms = pars2_50, func = C13model, times = outtimes)
out3_50 <- ode(y = state.ini(pars3_50), parms = pars3_50, func = C13model, times = outtimes)
```

```{r, fig.height = 3, fig.width = 8, fig.align='center', echo=FALSE}
# display exp. data and model predictions
plot(out1, out2, out3, out1_50, out2_50, out3_50,
     obs = list(DATA1, DATA2, DATA3), 
     obspar = list(pch=16, col=1:3),
     mfrow=c(1,3), lty=c(1,1,1,2,2,2), col=c(1,2,3,1,2,3), lwd=1)
legend("bottomright", legend=1:3, lty=1, lwd=2, col=1:3)
```

# Reference

- Rob J. M. van Spanning, Qingtian Guan, Chrats Melkonian, James Gallant, Lubos Polerecky, Jean-François Flot, Bernd W. Brandt, Martin Braster, Paul Iturbe Espinoza, Joost W. Aerts, Marion M. Meima-Franke, Sander R. Piersma, Catalin M. Bunduc, Roy Ummels, Arnab Pain, Emily J. Fleming, Nicole N. van der Wel, Vasile D. Gherman, Serban M. Sarbu, Paul L. E. Bodelier, and Wilbert Bitter (2022) *Methanotrophy by a Mycobacterium species that dominates a cave microbial ecosystem.* **Nature Microbiology 7**, 2089–2100. https://doi.org/10.1038/s41564-022-01252-3
