---
title: "RTM course - Rate Laws for Irreversible Chemical Reactions"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on creating rate expressions for elementary chemical reactions"
date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first. Only the first part (0:00-5:21) is relevant for this tutorial.

### Video timeline
* 0:00 Elementary chemical reactions
* 5:20 Reversible reactions
* 8:07 Chemical equilibrium
* 9:30 Modelling processes occurring at vastly different time-scales
* 17:24 Application to ammonia dissociation and pH modelling

![](https://www.youtube.com/watch?v=loHVpLNQBZE&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=5)

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>

You can also read the following section(s) of the textbook:

* Chapter 2.3. Formulation of chemical reaction

(You can download each chapter of the textbook via the e-book University library for free!)

## True/False statements

```{r statements, echo=FALSE}
question("Check boxes with a TRUE statement", type = "learnr_checkbox",
 answer("Stoichiometric coefficients in an elementary chemical reaction directly determine the exponents in the expression for the reaction rate", correct=TRUE),
 answer("Stoichiometric coefficients in any chemical reaction directly determine the exponents in the expression for the reaction rate", correct=FALSE),
 answer("Monod kinetics can be derived from an enzymatic reaction", correct=TRUE) 
)
```

## Reaction rate expression

Consider the following irreversible elementary chemical reaction:
$$
2A + B \rightarrow A_2B
$$

```{r elementary, echo=FALSE}
msg <- "There are two correct answers; answer 1 assumes that the reaction rate R is expressed in mole of B per time; answer 4 assumes that R is expressed in mole of A per time - it is simplest to express the rate in terms of mole of B"

question("Which of the following expressions correctly describe the reaction rate?", type = "learnr_checkbox",
 answer(sprintf("$R=r \\times [A]^2 \\times [B]$"), correct=TRUE),
 answer(sprintf("$R=r \\times [A] \\times [B]^2$"), correct=FALSE),
 answer(sprintf("$R=-r \\times [A_2B]$"), correct=FALSE),
 answer(sprintf("$R=r \\times [A] \\times [B]^{1/2}$"), correct=TRUE),
 answer(sprintf("$R=r \\times [A]^{1/2} \\times [B]$"), correct=FALSE),

 incorrect=msg,  
 correct=msg)
```

## Units

Consider the following irreversible elementary chemical reaction:
$$
2A + B \rightarrow A_2B
$$

Concentrations of A, B and $A_2B$ are expressed in $mol~m^{-3}$, and the expression for the reaction rate is 
$$
R=r \times [A]^2 \times [B]
$$

```{r rate, echo=FALSE}
question("What is the unit of the rate *R*?", type = "learnr_checkbox",
 answer(sprintf("$mol~m^{-3}$"), correct=FALSE),
 answer(sprintf("$mol~m^{-3} s^{-1}$"), correct=TRUE),
 answer(sprintf("$mol^2~m^{-6} s^{-1}$"), correct=FALSE)
)
```


```{r ct, echo=FALSE}
question("What is the unit of the rate constant *r*?", type = "learnr_checkbox",
 answer(sprintf("$mol~m^{-3}$"), correct=FALSE),
 answer(sprintf("$mol~m^{-3} s^{-1}$"), correct=FALSE),
 answer(sprintf("$(mol~m^{-3})^{-2} s^{-1}$"), correct=TRUE)
)
```

## Mass balance equation

Consider the following irreversible elementary chemical reaction and rate expression:
$$
2A + B \rightarrow A_2B
$$
$$
R=r \times [A]^2 \times [B]
$$

```{r massbal, echo=FALSE}
question("Toggle the correct mass balance equations:", type = "learnr_checkbox",
 answer(sprintf("$\\frac{d[A]}{dt}=-R^2$"), correct=FALSE),
 answer(sprintf("$\\frac{d[A]}{dt}= -2\\times R$"), correct=TRUE),
 answer(sprintf("$\\frac{d[A]}{dt}= -R/2$"), correct=FALSE),
 answer(sprintf("$\\frac{d[B]}{dt}= R$"), correct=FALSE),
 answer(sprintf("$\\frac{d[B]}{dt}= -R$"), correct=TRUE),
 answer(sprintf("$\\frac{d[B]}{dt}= -R/2$"), correct=FALSE),
 answer(sprintf("$\\frac{d[A_2B]}{dt}= R$"), correct=TRUE),
 answer(sprintf("$\\frac{d[A_2B]}{dt}= R^2$"), correct=FALSE),
 answer(sprintf("$\\frac{d[A_2B]}{dt}= 2 \\times R$"), correct=FALSE),
 answer(sprintf("$\\frac{d[A_2B]}{dt}= R/2$"), correct=FALSE)
)
```

## R-implementation

We repeat the equations from the previous question here:

$$
\frac{d[A]}{dt}=-2 \times R; \qquad
\frac{d[B]}{dt}= - R; \qquad
\frac{d[A_2B]}{dt}= R \\
R = r \times [A]^2 \times [B]
$$

It is quite easy to implement this chemical model in R, as shown here.

```{r, message=FALSE}
require(deSolve)   # package with the solution methods

y.ini <- c(A=1, B=1, A2B=0)  # initial concentrations of the state variables, [mol/m3]
pars  <- c(r=0.01)           # rate constant, [(molA/m3)^{-2}/s]

# time-derivative function
elementaryReaction <- function(t, y, p){  # t=time, y=state variables, p=parameters
  with (as.list(c(y,p)),{
    # Rate expressions
    R <- r * A^2 * B         # molB/m3/s
    # Mass balance equations
    dA.dt   <- -2*R
    dB.dt   <- -  R
    dA2B.dt <-    R
    # return vector of time-derivatives and ordinary variables
    list(c(dA.dt, dB.dt, dA2B.dt),  # NOTE: same order as the definition of y.ini!
      sumA = A + 2*A2B,    # lump-sum of A, molA/m3
      sumB = B +   A2B     # lump-sum of B, molB/m3
    )
  })
}
```

The model is run twice, each time with a different value of the rate constant $r$:
```{r}
# model solution
times <- seq(from=0, to=3600, length.out=100)  # time in seconds, 100 equally spaced values
out  <- ode(y=y.ini, times=times, parms=pars, func=elementaryReaction)

pars2 <- c(r=0.05) # 5-fold faster reaction
out2  <- ode(y=y.ini, times=times, parms=pars2, func=elementaryReaction)

plot(out, out2, lty=1, col=1:2)
```

**TASK:** Run the model with a rate that is twice as slow as the default rate. Plot all three model results in one graph.

```{r run2, exercise=TRUE, exercise.cap = "Three model runs", exercise.lines = 10}
# model solution
times <- seq(from=0, to=3600, length.out=100)
out   <- ode(y=y.ini, times=times, parms=pars, func=elementaryReaction)

pars2 <- c(r=0.05)
out2  <- ode(y=y.ini, times=times, parms=pars2, func=elementaryReaction)

out3  <-     ### RUN WITH r=0.005
plot(out, out2, out3, lty=1, col=1:3)
```

## More exercises

To learn more about modelling of chemical reactions --- and specifically of irreversible elementary chemical reactions --- try to solve the *ozone* model available in the R-package RTM. To display this exercise, type the following commands in the R-console:

```
require(RTM)
RTMexercise("ozone")
```

## References

R Core Team (2020). R: A language and environment for statistical computing. R
  Foundation for Statistical Computing, Vienna, Austria. https://www.R-project.org/.

Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential
  Equations in R: Package deSolve. Journal of Statistical Software, 33(9), 1--25.
  DOI: 10.18637/jss.v033.i09. http://www.jstatsoft.org/v33/i09/

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.
