---
title: 'The RTM package'
subtitle: "a Learning Environment for Reaction Transport Modelling"
author: "Lubos Polerecky, Dries Bonte and Karline Soetaert"
date: "July 2021"
output: pdf_document
abstract: \noindent Since many years we have been teaching courses on environmental and reaction-transport modelling to master students at Ghent and Utrecht University. Our aim is to teach students how to create models from scratch and apply them to advance their understanding of natural systems, including those perturbed by anthropogenic activities. Typically, students attending our classes have different backgrounds and highly variable levels of skills in quantitative and computational methods, which makes it difficult to find a teaching style that suits everyone. To overcome this issue, we revised our didactic approach and now teach in the form of a "flipped class", where students prepare themselves ahead of the class using material available on-line, and then, during the class, put the learned concepts into practice by solving problems and exercises with a direct feedback from us. To facilitate this teaching approach, we created educational videos, tutorials, and exercises that guide the students in learning and exploring the topics covered in the courses. Here, we present the R-package RTM, which merges this teaching material together with our modelling and teaching experiences under the common platform of the programming language R.
vignette: >
  %\VignetteIndexEntry{RTM: learning environment for reaction transport modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{echo = FALSE, include = FALSE}
output: rmarkdown::html_vignette
  \usepackage[T1]{fontenc}
```

```{r, include = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The courses Environmental Modelling and Reactive Transport Modelling in the Hydrosphere have been taught at Ghent and Utrecht University, respectively, since many years. Students attending these courses learn how to make mechanistic models that start from first principles of physics, chemistry, biology, and ecology, are mathematically formulated as differential equations, and are solved numerically by a computer. The models aim to advance their understanding of natural systems by allowing prediction of the systems' behaviour under various scenarios (e.g., response of the global carbon cycle to fossil fuel burning).

The aim of the courses is to teach students how to make models from scratch. This differs from many modelling courses where students merely learn how to apply existing tools. More specifically, students learn to

  - create conceptual models that describe environmental scenarios or problems,
  - translate conceptual models into mechanistic models and, ultimately, mathematical equations that describe the variation of the state variables characterising the system in time and space,
  - implement these mathematical equations in a computer program and solve them,
  - interpret the results in terms of the original environmental scenario or problem, and
  - apply the models to realistic scenarios.

We use R (R core team 2020) as the programming language for implementing and solving these models. This is mainly because there is an "ecosystem" of R-packages that provide methods for solving the type of differential equations that we deal with in the courses, including *rootSolve*, *deSolve*, *ReacTran* (Soetaert, 2009; Soetaert et al., 2010; Soetaert and Meysman, 2012). These packages have been implemented so that they hide much of the mathematical details behind the methods, allowing students to focus on the original problem rather than being concerned with finding a numerical approach to solve it. Additionally, they provide auxiliary functions that facilitate tabulation and visualisation of model results, fitting of models to data, etc.

Our students typically follow a master program in Earth or Life Sciences, and are interested in disciplines such as oceanography, biology, biogeochemistry, hydrology, or physics. Some have a good understanding of the processes occurring in the natural environment, but have never written a computer program before; others have ample experience in programming, mathematics or computer science, but little knowledge about the processes in nature. Due to this variation in background, knowledge and skills, a big challenge is to find a teaching style that can stimulate efficient and effective learning for all students. 

# Our way of teaching: "flipped class"

In the "traditional" learning environment, the instructor explains the context, theory and examples during lectures, and encourages students to practice the taught material by themselves at home (e.g., by solving problems and exercises given as homework assignments). In our "flipped class" approach, this process is reversed. Students obtain the theoretical lectures in advance and prepare themselves before the class. Then, during the class, they practice the learned concepts by solving problems and exercises with a direct feedback from the instructor.

One of the advantages of the flipped class is that students can work more focused and obtain a more thorough understanding of the material. Typically, the level of understanding of students in the modelling class is highly diverse. Thus, in the traditional way of teaching, the brighter students often get bored as they feel that the material is being taught too slowly, while less bright students become demotivated or frustrated as they cannot follow. In the flipped class approach, students can take the time they need to prepare themselves for the lessons individually. Subsequently, while in the classroom, they can directly interact with tutors and therefore explore the topics to the depth determined by their own ability, motivation and ambition. This provides more room for tutoring "a la tete du client".

The modelling course taught by Karline Soetaert and Dries Bonte at Ghent University has been in the form of a flipped class since the academic year 2019--2020, while the course taught by Lubos Polerecky and Karline Soetaert at Utrecht University first implemented this approach in the academic year 2020--2021. This way of teaching proved to be a positive experience:
while in previous years students would complain that they do not get enough practice in class, there is now ample time for practical exercises. Also the class is much more dynamic, with dramatically improved students participation and involvement.

# Implementation of the flipped classroom

To prepare for each lesson, students are offered the choice to either read one or more chapters from the textbook (Soetaert and Herman, 2009), or watch a short video that explains the theory. Then they are encouraged to test the understanding of the theory by answering a number of short (often multiple-choice) questions. After this "priming", students go to the classroom, where they submerge themselves, under the supervision of tutors, in solving more substantial exercises.

## Videos that elaborate on the theory

The theory behind the material taught in the modelling courses is explained in a series of educational videos. The videos, which typically take around 15 minutes or less, were created as follows:

1. Didactically sound slide presentations were created.
2. The narrative for the material on the slides was written down.
3. The narrative was recorded and combined with the slides.
   
Steps 1 and 2 were created by the tutors, while step 3 was partially outsourced and partially edited by the tutors.

Creation of the educational videos offers several longer-term advantages. For example, they were made available to the public as a YouTube playlist (\url{https://www.youtube.com/playlist?list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn}), and so can serve as an online course. Also, if students wish to refresh their knowledge of the subject, they can re-watch the videos at any time and in any order. Importantly, they can watch the videos at their own pace (e.g., if some students find the videos too slow, multi-tasking or the fast-forward knob is an option).

## Quiz-like questions

Each video is accompanied by a set of quiz-like questions and small exercises. These were implemented in the form of *tutorials* provided by the R-package *learnr* (Schloerke et al., 2020). They mostly consist of multiple-choice questions, sprinkled here and there with a more substantial exercise that needs an R-code for it to be solved.

The quizzes were designed to allow the students to test their basic understanding of the subject explained in the video. Before we introduced these quizzes, we often noted that some students overestimate their grasp of the material and are overly confident, while others are baffled by the complexity of the material and may loose confidence that they will eventually manage to understand it. The quizzes allow students to make this assessment by, and for, themselves. They are not graded, but the students are encouraged to use them for their own benefit, as they facilitate better preparation for the more substantial exercises in the classroom and --- ultimately --- for the exam. The quizzes also improve students’ quantitative and problem-solving skills by giving them hints and guidance as to how to tackle specific questions.

## Classroom exercises

While in the classroom, students work in small groups (2--4 students per group) on practical modelling exercises and applications. The focus is that the exercises address current environmental issues, the approaches required to solve the exercises are versatile, and that students design, develop and fully implement their own models from scratch (e.g., starting from a conceptual diagram and ending with an R-code that generates quantitative predictions).

At the time of writing this vignette, the exercises address, among others, the following environmental issues:

* *The CO$_2$ problem*. Atmospheric CO$_2$ concentrations are described by an Earth-system box model that comprises carbon fluxes among the major Earth's carbon reservoirs. The model is used to estimate the impact of anthropogenic activities (e.g., burning of fossil fuels, deforestation) on the atmospheric CO$_2$ concentrations and to explore the effects of certain mitigation strategies (e.g., ocean fertilisation).
* *The ozone problem*. Ozone is a green-house gas and, at high concentrations, harmful to humans and animals. It also plays an important protective role by absorbing harmful UV radiation. The ozone model is used to test how anthropogenic emissions of NO (nitric oxide) due to combustion or burning of fuels alter the natural ozone dynamics in the lower atmosphere. It is a good example of a "typical" chemistry model.
* *Ocean acidification*. This is a very complex problem as it deals with equilibrium chemistry. The topic is introduced by three exercises of increasing complexity. A more extensive model dealing with ocean acidification is not part of the mandatory course material, but is provided as auxiliary material that students may consult if they are interested in the topic (see below).
* *The COVID pandemic*.  An epidemiological model (so-called SIR model) is developed to describe the spread of the SARS-CoV2 (corona) virus within a population. In addition to infection and recovery, the model includes hospitalisations, mortality, vaccination, and virus mutations. The students practice how to use information from epidemiologists (who are not modelers) to develop and apply models.
* *Agricultural eutrophication*. This model describes how fertilization with nutrients affects the competition between crops and weeds, and explores the impact of different fertilization strategies on agricultural run-off. In addition to ecological interactions, the model also contains a (simplistic) economic component that lets students explore the monetary aspects of farming.
* *Dead zones in a river*. This one-dimensional reaction-transport model describes the distribution of dissolved oxygen concentrations in a river subjected to excessive nutrient input (ammonia). The model is used to estimate the maximal nutrient inputs that can be allowed to keep the minimal oxygen concentration above a certain critical level.

Additional exercise topics include

* Introduction to R for modellers
* Making of conceptual diagrams and formulating of mass balance equations
* Dissolution of silica particles (e.g., diatom frustules)
* Bacteria-mediated mineralisation of detritus
* Early diagenesis in marine sediments (a simplified P cycle, and the coupled cycles of C, N, O, and S)
* Food web modelling of a lake or river ecosystem
* Competition between plants in grasslands
* Competition between floating plants and algae in shallow lakes
* Virus dynamics in the ocean
* Bio-economic modelling of a Scallop aquaculture

New material can be added in the future, thus creating a repository from which modules can be selected based on students' interests that vary from year to year.

## Auxiliary material

CONTINUE EDITING HERE

This consists of a number of small documents that go into somewhat more detail for various topics. It is not compulsory material, so the students get to see this after the exam, and before they start the projects (see below).

These "readers" are meant to satisfy the curiosity of some students that have a desire to go more into depth for certain topic. 

They comprise documents dealing with:

* How to implement *events* in dynamic models developed in R
* How to use *forcing functions* based on data in models developed in R
* How to easily showing *observed data* alongside model results in R 
* How to *fit* a 1D reaction-transport model to data in R 
* How to *visualise* dynamic outputs from a 1D reaction-transport model in R
* How to include *pH dynamics* in a 1D reaction-transport model in R
* How to mathematically investigate the response of systems to a *perturbation from equilibrium*
* How to make *interactive applications* in R
* The basics of the *numerical methods* used for reaction-transport modelling in R

## Longer projects

After having done the written exam, the students work on a specific environmental topic in small groups of 3-4 people. In contrast to the exercises made in class, they do this with less guidance from the teachers. In addition, they also need to provide a report, where they outline the scientific rationale of their work, describe their model, and document the results, much like they would do while writing a scientific paper. 

Note that the topics of these projects are NOT part of the RTM package.

# About this package

This R-package contains the tutorials, exercises and extra material that the students have to work through during their courses at Utrecht University. 

Anyone that want to self-teach modelling can consult the tutorials and make the exercises. The answers are also provided, in a subdirectory under the inst package directory

## tutorials

The function *RTMtutorial* runs the tutorials. To list all possible tutorials:

```{r}
RTM:::RTMtutorial("?")
```

At the start of each tutorial is a link to the youtube video that deals with the theory.

## Exercises

R-function *RTMexercise* opens the PDF with the questions for each exercise. To list all possible exercises:

```{r}
RTM:::RTMexercise("?")
```

A number of model templates is available and can be downloaded from the Rstudio menu: "File->new File->Rmarkdown..->from template->name_of_the_template". 

## Accessory material

R-function *RTMreader* opens the PDF with the accessory (non-mandatory) material. To list all possible documents:
\scriptsize
```{r}
RTM:::RTMreader("?")
```
\normalsize
# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for
  Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Barret Schloerke, JJ Allaire and Barbara Borges (2020). learnr: Interactive Tutorials for R.
  R package version 0.10.1. https://CRAN.R-project.org/package=learnr
  
Soetaert Karline (2009).  rootSolve: Nonlinear root finding, equilibrium and steady-state analysis of ordinary differential equations.  R-package version 1.6
  
Soetaert Karline, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential
  Equations in R: Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL
  http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09
  
Soetaert, Karline and Meysman, Filip (2012). Reactive transport in aquatic ecosystems: Rapid model
  prototyping in the open source software R Environmental Modelling & Software, 32, 49-60.
  