---
title: "Evaporation and precipitation of water"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky, Utrecht University"
date: "January 2022"
output: pdf_document
---

# Task 1 --- solution

First, we implement the Clausius-Clapeyron equation (Eq. 10) to calculate the equilibrium vapor pressure, then use Eq. 13 to calculate the vapor pressure, and finally use Eq. 12 to calculate the molar fraction.
```{r, echo=TRUE, eval=TRUE}
# implement Clausius-Clapeyron equation
Peq_vapor <- function(TK=300)   # input temperature in K!
  101325 * exp( -40.66e3/8.314 * (1/TK - 1/373.15) )

# conditions
Pair  <- 0.95*101325            # air pressure, in Pa
humid <- 0.8                    # relative air humidity
TK    <- c(10, 20, 50) + 273.15 # temperature, in K
Peq   <- Peq_vapor(TK)          # equilibrium vapor pressure, in Pa
Pvap  <- humid*Peq              # vapor pressure, in Pa
xvap  <- Pvap/Pair              # molar fraction of water vapor

# display results as a data.frame
data.frame(T_K = TK, Peq_Pascal = Peq, xvap_percent = xvap*100)
```

The above results show that although the relative air humidity is constant, the molar fraction of water vapor steeply increases with temperature from about 1 % to about 11 %. This increase is caused by the steep increase in the equilibrium vapor pressure with temperature.

# Task 2 --- solution

## Model implementation in R

First, we define model parameters and empirical constants.

```{r}
# model parameters
Aliq     <- 1           # liquid water area (m2) 
Ptot     <- 101325      # total pressure (Pa)
wind_v   <- 0           # wind speed (m h-1)
TK1      <- 293.15      # temperature during first period (K)
TK2      <- 283.15      # temperature during second period (K)

# empirical constants or parameters
Rgas     <- 8.314       # gas constant (J mol-1 K-1)
dHvap    <- 40.66e3     # molar enthalpy of vaporisation of water (J mol-1)
c1       <- 0.0888*3600 # empirical constant in Eq. 7 (m h-1)
c2       <- 0.0783      # empirical constant in Eq. 7 (-)
rho      <- 1e3         # water density (kg m-3)
MW       <- 18e-3       # molar weight of water (kg mol-1)
```

Then, we define auxillary functions for calculating the amounts of water molecules from the volume of the water vapor and liquid, respectively, and the rate constants from the temperature. We use the ``Peq_vapor()`` function defined above.

```{r}
Nwater_vapor <- function(humid, TK, V){ # calculate Nvap from volume, T, and humidity
  humid*Peq_vapor(TK)*V/(TK*Rgas)       # based on the ideal gas law (Eq. 5) and Eq. 13
}

Nwater_liquid <- function(V){           # calculate Nliq from volume
  rho*V/MW                              # using density and molar weight
}

kbT <- function(TK, v=0){               # rate constant of water precipitation (m h-1)
  Rgas*TK/dHvap*(c1+c2*v)               # Eq. 8
}

kfT <- function(TK, v=0){               # rate constant of water evaporation (mol m-2 h-1)
  kbT(TK, v) * Peq_vapor(TK)/(Rgas*TK)  # Eq. 9
}
```

Finally, we define the model function based on the differential equations 1 and 2. 

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
WaterEvapPrecip <-function(t, state, pars) {
  with (as.list(c(state, pars)),{
    
    # rate constants at the given temperature (TK given in pars)
    kb <- kbT(TK, v)
    kf <- kfT(TK, v)
    
    # volume of the gas phase (air+vapor) at a given Nvap, Nair, TK, and Ptot
    # (Nair given in pars)
    Vgas <- Rgas*TK*(Nair+Nvap)/Ptot # based on the ideal gas law
    
    # process rates, mol h-1, Eq. 1
    evaporation   <- kf*Aliq * (Nliq/(Nliq+1e-5)) # ensure that evaporation only occurs
                                                  # if liquid water is present
    precipitation <- kb*Aliq * Nvap/Vgas
    
    # time-derivatives
    dNvap.dt <-  evaporation - precipitation
    dNliq.dt <- -evaporation + precipitation
    
    # extra output quantities 
    Pvap  <- Nvap/Vgas * Rgas * TK # partial pressure of vapor, Eq. 5
    Pair  <- Nair/Vgas * Rgas * TK # partial pressure of air
    Pgas  <- Pvap + Pair           # total pressure of the gas phase
    humid <- Pvap/Peq_vapor(TK)    # relative air humidity, Eq. 13
    Vliq  <- Nliq*MW/rho           # volume of the liquid water
    Vtot  <- Vgas + Vliq           # total volume (liquid + gas)
    
    # return time-derivatives and ordinary variables as a list
    return(list(c(dNvap.dt, dNliq.dt),
         
         # extra output:
         # amounts of molecules
         Nwater = Nvap + Nliq,                  # water, total
         Nair   = Ptot * Vgas/(Rgas*TK) - Nvap, # air, total (to check!)
         
         # process rates, mol h-1
         R.evaporation   = evaporation,
         R.precipitation = precipitation,
         
         # rate constants
         kb = kb,             # m h-1
         kf = kf,             # mol m-2 h-1
         
         # volumes, m3
         Vgas = Vgas,         # gas phase
         Vliq = Vliq,         # liquid phase
         Vtot = Vtot,         # total (gas+liquid)
         
         # pressures, Pa
         Pvap = Pvap,  # partial pressure of vapor
         Pair = Pair,  # partial pressure of air
         Pgas = Pgas,  # total gas pressure (vapor+air)
         
         # fractions
         xvap = Nvap/(Nvap+Nair), # molar fraction of water vapor in air, Eq. 11
         humid = humid            # relative air humidity, (-)
         ))
  })
}

```

## Model application

### Initial conditions and run 1

First, we define a function that returns the values of state variables (i.e., initial amounts of water vapor and liquid water molecules) and model parameters (i.e., temperature and amount of air molecules) based on the conditions defined by the problem (i.e., initial total volume, initial volume of the liquid water, initial relative air humidity, and initial temperature).

```{r}
find_state_pars <- function(Vtot, Vliq, humid, TK, v){
  Vgas <- Vtot - Vliq  # volume of the gas phase = Vtotal - Vliquid

  # amounts of vapor, liquid, and air in the gas phase
  Nvap <- Nwater_vapor(humid=humid, TK=TK, V=Vgas)
  Nliq <- Nwater_liquid(V=Vliq)
  Nair <- Ptot * Vgas/(Rgas*TK) - Nvap
  
  # output: state variables and parameters
  return( list(SV   = c(Nvap = Nvap, Nliq = Nliq), 
               pars = c(TK   = TK,   Nair = Nair, v = v)) )
}
```

Now, we use the above function to generate the initial conditions and model parameters and solve the model over the first half-hour.

```{r, message=FALSE}
# initial conditions
Vtot.ini <- 1; Vliq.ini <- 1e-3; humid.ini <- 0
ini   <- find_state_pars(Vtot=Vtot.ini, Vliq=Vliq.ini, humid=humid.ini, TK=TK1, v=wind_v)

# solve the model over the first half-hour
outtimes <- seq(from = 0, to = 0.5, length.out = 400) # time in hr
require(deSolve)
out1 <- ode(y = ini$SV, parms = ini$pars, func = WaterEvapPrecip, times = outtimes)
```

### New initial conditions and run 2

In the second step, we use the end values of the first model run to calculate the initial conditions for the second model run. The gas was cooled from $T_1$ to $T_2$ at a constant pressure. Thus, the pressures do not change (this is true for the total pressure as well as for the partial pressures of the components in the gas phase). However, the volume of the gas phases decreases by a factor $T_2/T_1$, as follows from the ideal gas law. Also, the equilibrium pressure decreases and, subsequently, the relative humidity increases as a result of the cooling. The change in the volume of the liquid water due to cooling is neglected (see Assumptions). 

```{r}
# calculate new initial conditions
SV.end <- out1[nrow(out1),]                    # end of the previous run
ini2   <- find_state_pars(
           Vtot  = SV.end[["Vgas"]]*TK2/TK1 +  # gas volume changes (isobaric cooling)
                   SV.end[["Vliq"]],           # change in liquid volume neglected
           Vliq  = SV.end[["Vliq"]], 
           humid = SV.end[["Pvap"]]/Peq_vapor(TK=TK2), 
           TK    = TK2,
           v     = wind_v)

# solve the model
out2 <- ode(y = ini2$SV, parms = ini2$pars, func = WaterEvapPrecip, times = outtimes)
out2[,"time"] <- out2[,"time"] + max(outtimes) # modify time (for a better display)
```

## Results & Discussion

The first set of graphs (see below) shows that, due to net evaporation at the higher (initial) temperature, the relative air humidity increases from the initial 0 % to 100 % **within about 15 minutes**. This is associated with an increase in the molar fraction (from 0 to 0.028), vapor pressure (from 0 Pa to 2800 Pa) and a transfer of about 1.2 moles of water from the liquid to the gas phase. 

The cooling of the 100 % humid air after $30~min$ increases the relative air humidity to about 180 %, leading to net precipitation. The new equilibrium is reached, again, in about 15 minutes, but the total changes in $x_{vap}$, $P_{vap}$, and $N_{vap}$ are smaller.

```{r, fig.height=4.4, fig.width=8}
plot(out1, out2,
     which=c("humid", "xvap", "Pvap", "Pair", 
             "Nvap", "Nliq", "Nair", "Nwater"),
     xlab="time (hr)", lty=1, las=1, lwd=2, mfrow=c(2,4),
     ylab=c("-", "-", "Pa", "Pa", rep("mol H2O",3)))
legend("bottomright", title="temperature (K):",
       legend=c(sprintf("%.1f",ini$pars["TK"]), 
                sprintf("%.1f",ini2$pars["TK"])), 
       lty=1, lwd=2, col=1:2, bty="n")
```

The second set of graphs (see below) shows that evaporation occurs at a rate of about ``r sprintf("%.1f", out1[1,"R.evaporation"])`` $mol~h^{-1}$ and ``r sprintf("%.1f", out2[1,"R.evaporation"])`` $mol~h^{-1}$ at the higher and lower temperature, respectively. During the first and second half-hour period, the precipitation rate increases and decreases towards the evaporation rate, respectively, reaching the level in about 15 minutes. Rate constants remain constant during each half-hour interval because the temperature is constant, but the values differ for the different temperatures (see Eqs. 8 and 9, where $c_1$ and $c_2$ are assumed to be temperature-independent).

```{r, fig.height=2.5, fig.width=8}
plot(out1, out2,
     which=c("R.evaporation", "R.precipitation", "kf","kb"),
     xlab="time (hr)", lty=1, las=1, lwd=2, mfrow=c(1,4),
     ylab=c("mol h-1", "mol h-1", "mol m-2 h-1", "m h-1"))
legend("right", title="temperature (K):",
       legend=c(sprintf("%.1f",ini$pars["TK"]), 
                sprintf("%.1f",ini2$pars["TK"])), 
       lty=1, lwd=2, col=1:2, bty="n")
```

The third set of graphs (see below)  shows that, due to net evaporation at the higher (initial) temperature, the volume of the gas phase (air + water vapor) increases by about ``r sprintf("%.2f", 1e3*diff(out1[c(1,nrow(out1)),"Vgas"]))`` $L$ and the volume of the liquid water decreases by ``r sprintf("%.1f", 1e6*diff(out1[c(nrow(out1),1),"Vliq"]))`` $mL$. Thus, the overall volume of the system increases by about ``r sprintf("%.2f", 1e3*diff(out1[c(1,nrow(out1)),"Vtot"]))`` $L$. Because the total pressure is constant, these changes are associated with an increase in the vapor pressure but a decrease in the air pressure (i.e., pressure of air components other than the water vapor; see Pvap and Pair in the first set of graphs). The trends are opposite at the lower temperature during the second half-hour. The isobaric cooling in between the two half-hour intervals has a similar effect on the gas and total volumes as the evaporation/precipitation, but no effect on the liquid volume (as assumed).

```{r, fig.height=2.5, fig.width=8}
plot(out1, out2,  which=c("Vgas", "Vliq", "Vtot"),
     xlab="time (hr)", lty=1, las=1, lwd=2, mfrow=c(1,3), ylab="m3")
```

## Water budgets

Finally, we construct the water budgets by integrating\footnote{We use the ``integrateTrapezoid`` function from the ``oce`` package to perform the integration.} the evaporation and precipitation rates over time, which yields *gross* amounts of evaporated and precipitated water, respectively. We also calculate the net amount of evaporated water as the difference in the gross amounts, which should be equal to the difference in the final and initial amounts of water vapor (it is). We also convert the amount of liquid water (in moles) to the volume of liquid water (in $mL$), which is easier to imagine.

```{r, message=FALSE}
require(oce)
evap <- integrateTrapezoid(outtimes, out1[,"R.evaporation"])   # gross amount evaporated
prec <- integrateTrapezoid(outtimes, out1[,"R.precipitation"]) # gross amount precipitated
dN   <- diff(out1[c(1,nrow(out1)),"Nvap"])                     # difference final - initial

# water budget in moles
(budget_mol <- c(evap=evap, prec=prec, net=evap-prec, net2=dN))

# water budget in milliliters
(budget_ml <- budget_mol*MW/rho * 1e6)
```

The above results show that, when going from ``r humid.ini`` % to ``100`` % humid air at ``r TK1-273.15`` $^\circ C$, about ``r sprintf("%.2f", budget_mol["evap"])`` and ``r sprintf("%.2f", budget_mol["prec"])`` moles of water is evaporated and precipitated, respectively, whereas the net increase in the amount of water vapor is only ``r sprintf("%.2f mol", budget_mol["net"])``. These amounts correspond to ``r sprintf("%.0f mL", budget_ml["evap"])`` of water evaporated, ``r sprintf("%.0f mL", budget_ml["prec"])`` of water precipitated, and ``r sprintf("%.0f mL", budget_ml["net"])`` of net-evaporated water. Thus, the **gross** exchange of water between the liquid and gas phase is about **10-fold larger** than the **net** increase in water vapor. Moreover, for an initial setup consisting of ``r Vliq.ini*1e3`` L of liquid water surrounded by ``r (Vtot.ini-Vliq.ini)*1e3`` L of air with a relative humidity of ``r humid.ini*100`` % (surface area of ``r Aliq`` $m^2$), about 
``r sprintf("%.0f", budget_ml["evap"] / (Vliq.ini*1e6) * 1e2)`` %
of water evaporates, 
``r sprintf("%.0f", budget_ml["prec"] / (Vliq.ini*1e6) * 1e2)`` %
of water precipitates, and
``r sprintf("%.0f", budget_ml["net"] / (Vliq.ini*1e6) * 1e2)`` %
of the liquid water eventually ends up as water vapor within about 15 minutes. 

For completeness, we also construct a similar water budget for the second half-hour.

```{r, message=FALSE}
require(oce)
evap <- integrateTrapezoid(outtimes, out2[,"R.evaporation"])   # gross amount evaporated
prec <- integrateTrapezoid(outtimes, out2[,"R.precipitation"]) # gross amount precipitated
dN   <- diff(out2[c(1,nrow(out2)),"Nvap"])                     # difference final - initial

# water budget in milliliters
(budget_ml2 <- c(evap=evap, prec=prec, net=evap-prec, net2=dN)*MW/rho * 1e6)
```
Thus, the decrease in temperature from ``r TK1-273.15`` $^\circ C$ to ``r TK2-273.15`` $^\circ C$ resulted in about ``r sprintf("%.0f mL", budget_ml2["prec"])`` of water precipitated, ``r sprintf("%.0f mL", budget_ml2["evap"])`` of water evaporated, and ``r sprintf("%.0f mL", -budget_ml2["net"])`` of water net-precipitated. 

# Conclusion

The model results illustrate that much more water evaporation and precipitation occurs than "meets the eye"! The exchange occurs on the time-scale of minutes, consistent with our daily experience.
