---
title: 'Chemistry: Ozone in the troposphere'
author: "Karline Soetaert and Lubos Polerecky, Utrecht University"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
date: "February 2021"
output:
  pdf_document: default
  html_document: default
---

# Answers

## Rate expressions

The reactions to be considered are:

$$Reaction~1:\qquad NO_2 + hv \rightarrow NO+O,$$
$$Reaction~2:\qquad O+O_2 \rightarrow O_3,$$
$$Reaction~3:\qquad NO+O_3 \rightarrow O_2 + NO_2,$$
where $O_2$ is assumed to be constant and thus not modeled explicitly. 

Based on the reactions and the input flux of $NO$ ($\sigma$), we write mass balance equations for the *total* mass of $O$, $NO$, $NO_2$, and $O_3$ in the lower atmosphere as follows:

$$\frac{d[O]}{dt}  = R_1 - R_2,$$
$$\frac{d[NO]}{dt}   = R_1 - R_3 + \sigma,$$
$$\frac{d[NO_2]}{dt}  = - R_1 + R_3,$$
$$\frac{d[O_3]}{dt}  = R_2 - R_3,$$
where $R_1$, $R_2$ and $R_3$ denote the rates of the reactions 1, 2 and 3, respectively. Here, we denote the *total mass* in the compartment $x$ as $[x]$. This notation should not be confused with the notation often used for concentrations.

To derive the rate expressions, we assume that each reaction is an elementary rection and thus its rate is described by the first-order kinetics. Thus, for $R_1$, $R_2$ and $R_3$ we write:

$$R_1 = k_1\cdot [NO_2],$$
$$R_2 = k_{2a}\cdot [O]\cdot [O_2],$$
$$R_3 = k_3\cdot [NO]\cdot [O_3].$$
The rate expression for $R_2$ can be simplified if we assume the total mass of oxygen $[O_2]$ to be constant. Thus, we obtain:
$$R_2 = k_2\cdot [O],$$
where $k_2 = k_{2a}*[O_2]$.

To implement the dependency of the photo-dissociation rate on the light intensity, we write the corresponding reaction rate constant  as
$$k_1 = k_{1a}+k_{1b}\cdot radiation.$$
 

## Model implementation in R

This model is implemented and solved in R. We impose variable light intensity that changes over a day.

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
require(deSolve)  # package with solution methods

# state variables, units = moles
state <- c(O = 0, NO = 1.3e8, NO2 = 5e11, O3 = 8e11)  # initial conditions

# parameters: rate coefficients
parms <- c(
  k3     = 1e-11,     # [/(mol/d)] 
  k2     = 1e10,      # [/d]
  k1a    = 1e-30,     # [/d] note: k1 = k1a + k1b*radiation
  k1b    = 1,         # [/(microEinst/m2/s)/d]
  sigma  = 1e11,      # [mol/d]           NO emission rate
  maxrad = 1200       # [microEinst/m2/s] maximal radiation
)

# Model function
Ozone <- function(t, state, params) {
  with (as.list(c(state, params)), {

  radiation <- max(0, sin(t*2*pi))*maxrad    # radiation at time t (if time in days)

  # Rate expressions
  k1 <- k1a + k1b*radiation

  R1 <- k1*NO2
  R2 <- k2*O
  R3 <- k3*NO*O3
  
  # Mass balances [moles/day]
  dO   <-  R1 - R2
  dNO  <-  R1      - R3 + sigma
  dNO2 <- -R1      + R3
  dO3  <-       R2 - R3

  list(c(dO, dNO, dNO2, dO3), 
       radiation = radiation)
  })

}
```

The model is run for 5 days; it is solved using the ode function from package deSolve. We compare the dynamics with and without the anthropogenic emissions.

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}

outtimes <- seq(from = 0, to = 5, length.out = 300)  # run for 5 days

out <- ode(y = state, parms = parms, func = Ozone, times = outtimes, method = "vode")

parms2 <- parms
parms2["sigma"] <- 0.0
out2 <- ode(y = state, parms = parms2, func = Ozone, times = outtimes, method = "vode")

plot(out, out2, las = 1, lwd=1, lty=1) # las specifies the rotation of axis labels
plot.new()
legend("top", legend=c("with emissions", "no emissions"), lty=1, lwd=1, col=1:2)
```

The model output shows that, at the onset of the day, the masses of $O$ and $NO$ increase drastically due to the photo-dissociation reaction, which rapidly exhausts $NO_2$. As most of the $O$ produced reacts with $O_2$ at a very high rate to form $O_3$, the mass of $O$ increases only little compared to $NO$. The continuous input of NO, however, leads to a gradual build-up of $NO_2$, which then yields a gradual decrease in the ozone levels during the night. 
