---
title: "The Coupled Cycles of Carbon, Oxygen, Nitrogen and Sulfur in Marine Sediments"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky and Karline Soetaert, Utrecht University"
date: 'July 2021'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Answers 

## Rate expressions

$$\begin{array}{l}
AeroMin = rMin \cdot \frac{O_2}{O_2+k_{O_2}} \cdot Corg \\[5mm]
Nitri = rNit \cdot \frac{O_2}{O_2+k_{O_2}} \cdot NH_3 \\[5mm]
Denitri = rMin \cdot \frac{k_{O_2}}{O_2+k_{O_2}} \cdot \frac{NO_3^-}{NO_3^-+k_{NO_3}} \cdot Corg \\[5mm]
Sred = rMin \cdot \frac{k_{O_2}}{O_2+k_{O_2}} \cdot \frac{k_{NO_3}}{NO_3^-+k_{NO_3}} \cdot \frac{SO_4^{2-}}{SO_4^{2-}+k_{SO_4}} \cdot Corg \\[5mm]
Sox = rSox \cdot \frac{O_2}{O_2+k_{O_2}} \cdot H_2S \\[5mm]
Methanogen = rMin \cdot \frac{k_{O_2}}{O_2+k_{O_2}} \cdot \frac{k_{NO_3}}{NO_3^-+k_{NO_3}} \cdot \frac{k_{SO_4}}{SO_4^{2-}+k_{SO_4}} \cdot Corg
\end{array}
$$

Units of the rates:
  
  * $AeroMin$, $Denitri$, $Sred$ and $Methanogen$ are in $mol~C~m^{-3}_S~d^{-1}$.
  
  * $Nitri$ is in $mol~N~m^{-3}_L~d^{-1}$.
  
  * $Sox$ is in $mol~S~m^{-3}_L~d^{-1}$.

Note that $AeroMin+Denitri+Sred+Methanogen$ is equal to $rMIN \cdot Corg$, as follows directly from the fact that the sum of the limitation and inhibition terms for a specific substrate is 1. Thus, the *total* mineralisation rate is first-order with respect to $Corg$ and independent of the terminal electron acceptor. The limitation and inhibition terms only dictate the *contributions* of the different oxidants to the overall mineralisation process.

## Mass balance equations

The updated mass balance equations that include all processes mentioned are:\footnote{Note the factors $\phi$ or $1-\phi$, which cannot be canceled out from the spatial derivative because the porosity varies with depth.}
$$
\begin{array}{l}
\frac{\partial Corg}{\partial t} = \frac{1}{1-\phi}\cdot \frac{\partial}{\partial x}\left[(1-\phi)\cdot D_b\cdot \frac{ \partial Corg}{\partial x}\right] 
-\frac{1}{1-\phi}\cdot \frac{ \partial }{\partial x}\left[(1-\phi)\cdot v\cdot Corg\right]
- AeroMin - Denitri - Sred - Methanogen \\[4mm]
\frac{\partial O_2}{\partial t} = \frac{1}{\phi}\cdot \frac{\partial}{\partial x}\left[\phi \cdot D_{O2}\cdot \frac{ \partial O_2}{\partial x}\right]
-\frac{1}{\phi}\cdot \frac{ \partial }{\partial x}\left[\phi\cdot v\cdot O_2\right] - AeroMin \cdot f_{2L} - 2 \cdot Nitri - 2 \cdot Sox \\[4mm]
\frac{\partial NH_3}{\partial t} = \frac{1}{\phi}\cdot  \frac{\partial}{\partial x}\left[\phi \cdot D_{NH_3}\cdot  \frac{ \partial NH_3}{\partial x}\right] 
-\frac{1}{\phi}\cdot \frac{ \partial }{\partial x}\left[\phi\cdot v\cdot NH_3\right]
+ \frac{16}{106} \cdot  (AeroMin + Denitri + Sred + Methanogen) \cdot  f_{2L} - Nitri \\[4mm]
\frac{\partial NO_3^-}{\partial t} = \frac{1}{\phi} \cdot \frac{\partial}{\partial x}\left[\phi\cdot  D_{NO_3^-}\cdot  \frac{ \partial NO_3^-}{\partial x}\right] 
-\frac{1}{\phi}\cdot \frac{ \partial }{\partial x}\left[\phi\cdot v\cdot NO_3^-\right]- 0.8 \cdot Denitri + Nitri \\[4mm]
\frac{\partial SO_4^{2-}}{\partial t} = \frac{1}{\phi}\cdot \frac{\partial}{\partial x}\left[\phi\cdot D_{SO_4^{2-}}\cdot \frac{ \partial SO_4^{2-}}{\partial x}\right] 
-\frac{1}{\phi}\cdot \frac{ \partial }{\partial x}\left[\phi\cdot v\cdot SO_4^{2-}\right] 
- 0.5 \cdot Sred\cdot f_{2L} + Sox \\[4mm]
\frac{\partial H_2S}{\partial t} = \frac{1}{\phi}\cdot \frac{\partial}{\partial x}\left[\phi\cdot D_{H_2S}\cdot \frac{ \partial H_2S}{\partial x}\right] 
-\frac{1}{\phi}\cdot \frac{ \partial }{\partial x}\left[\phi\cdot v\cdot H_2S\right]
+ 0.5\cdot Sred\cdot f_{2L}- Sox \\[4mm]
\frac{\partial CO_2}{\partial t} = \frac{1}{\phi} \cdot \frac{\partial}{\partial x}\left[\phi\cdot  D_{CO_2}\cdot  \frac{ \partial CO_2}{\partial x}\right] 
-\frac{1}{\phi}\cdot \frac{ \partial }{\partial x}\left[\phi\cdot v\cdot CO_2\right]
+ (OxicMin + Denitri + Sred + 0.5\cdot Methanogen)\cdot f_{2L} \\[4mm]
\frac{\partial CH_4}{\partial t} = \frac{1}{\phi} \cdot \frac{\partial}{\partial x}\left[\phi\cdot  D_{CH_4}\cdot  \frac{ \partial CH_4}{\partial x}\right] 
-\frac{1}{\phi}\cdot \frac{ \partial }{\partial x}\left[\phi\cdot v\cdot CH_4\right]
+ 0.5\cdot Methanogen\cdot f_{2L}
\end{array}
$$
where $f_{2L}=(1-\phi)/\phi$.

# R implementation

```{r, message = FALSE}
require(marelac)
require(ReacTran)
```

### The model grid and associated properties

```{r}
# spatial domain: total length of 0.10 m, 500 boxes
Length    <- 0.1   # [m]
N         <- 500

# grid with an exponentially increasing grid size, first cell 0.05 cm (5e-4 m)
Grid      <- setup.grid.1D(L = Length, N = N, dx.1 = 0.05/100)

# function describing the variation of porosity (volume fraction of LIQUID) with depth
porFun.L  <- function(x, por.SWI, por.deep, porcoef)
  return( por.deep + (por.SWI-por.deep)*exp(-x*porcoef) )

# function describing the SOLID volume fraction (svf = 1-porosity)
porFun.S  <- function(x, por.SWI, por.deep, porcoef)
  return( 1-porFun.L(x, por.SWI, por.deep, porcoef) )

# calculate porosity and svf on the grid (mid-points and box interfaces, etc.)
porLiquid <- setup.prop.1D(func=porFun.L, grid=Grid, por.SWI=0.9, por.deep=0.7, porcoef=100)
porSolid  <- setup.prop.1D(func=porFun.S, grid=Grid, por.SWI=0.9, por.deep=0.7, porcoef=100)

# molecular diffusion coefficients (m2/d) from the marelac package
diff.O2   <- diffcoeff(S=35, t=20)$O2  * 3600*24
diff.CO2  <- diffcoeff(S=35, t=20)$CO2 * 3600*24
diff.CH4  <- diffcoeff(S=35, t=20)$CH4 * 3600*24
diff.NH3  <- diffcoeff(S=35, t=20)$NH3 * 3600*24
diff.NO3  <- diffcoeff(S=35, t=20)$NO3 * 3600*24
diff.SO4  <- diffcoeff(S=35, t=20)$SO4 * 3600*24
diff.H2S  <- diffcoeff(S=35, t=20)$H2S * 3600*24

# effective diffusion coefficients in the sediment (corrected for tortuosity)
porInt    <- porLiquid$int                # porosity at the box interfaces            
diffO2    <- diff.O2 /(1-log(porInt^2))
diffCO2   <- diff.CO2/(1-log(porInt^2))
diffCH4   <- diff.CH4/(1-log(porInt^2))
diffNO3   <- diff.NO3/(1-log(porInt^2))
diffNH3   <- diff.NH3/(1-log(porInt^2))
diffSO4   <- diff.SO4/(1-log(porInt^2))
diffH2S   <- diff.H2S/(1-log(porInt^2))
```

## Other model parameters

```{r}
parms <- c(
 Dbio     = 5e-4/365,   # [m2/d]      bioturbation mixing coefficient
 v_adv    = 5e-6,       # [m/d]       advection velocity = sediment accretion rate
 rMin     = 0.005,      # [/d]        POC mineralisation rate constant
 depoPOC  = 1e-3,       # [mol/m2/d]  POC deposition rate at SWI (sediment-water interface)
 rNit     = 10,         # [/d]        Nitrification rate constant
 rSox     = 10,         # [/d]        H2S reoxidation rate constant
 kO2      = 0.001,      # [mol/m3]    half-saturation O2 concentration
 kNO3     = 0.0001,     # [mol/m3]    half-saturation NO3 concentration  
 kSO4     = 0.1,        # [mol/m3]    half-saturation SO4 concentration  
 bwO2     = 0.300,      # [mol/m3]    O2 concentration at SWI
 bwCO2    = 2,          # [mol/m3]    CO2 concentration at SWI
 bwNO3    = 0.010,      # [mol/m3]    NO3 concentration at SWI
 bwNH3    = 0.001,      # [mol/m3]    NH3 concentration at SWI
 bwSO4    = 28,         # [mol/m3]    SO4 concentration at SWI
 bwH2S    = 0,          # [mol/m3]    H2S concentration at SWI
 bwCH4    = 0           # [mol/m3]    CH4 concentration at SWI
)
```

## Definition and initialisation of state variables

```{r}
names     <- c("POC", "CO2", "O2", "NH3", "NO3", "SO4", "H2S", "CH4")
nspec     <- length(names)
POC.ini   <- rep(0, length = N)      # initial conditions 
CO2.ini   <- rep(0, length = N)
O2.ini    <- rep(0, length = N)
NH3.ini   <- rep(0, length = N)
NO3.ini   <- rep(0, length = N)
SO4.ini   <- rep(0, length = N)
H2S.ini   <- rep(0, length = N)
CH4.ini   <- rep(0, length = N)
state.ini <- c(POC.ini, CO2.ini, O2.ini, NH3.ini, NO3.ini, SO4.ini, H2S.ini, CH4.ini)
```

## Definition of the model function

```{r}
Diamodel <- function (t, state, pars)   # state is a LONG vector
{
  with (as.list(pars),{ 
    
    # unpack state variables
    POC <- state[(0*N+1):(1*N)]    # first N elements: POC
    CO2 <- state[(1*N+1):(2*N)]    # next N elements:  CO2
    O2  <- state[(2*N+1):(3*N)]    # next N elements:  O2
    NH3 <- state[(3*N+1):(4*N)]    # next N elements:  NH3
    NO3 <- state[(4*N+1):(5*N)]    # next N elements:  NO3
    SO4 <- state[(5*N+1):(6*N)]    # next N elements:  SO4
    H2S <- state[(6*N+1):(7*N)]    # next N elements:  H2S
    CH4 <- state[(7*N+1):(8*N)]    # next N elements:  CH4

    # === transport rates ===
    # note: zero gradient by default at lower boundaries
 
    # solid substances, VF = solid volume fraction = 1-porosity!    
    tran.POC <- tran.1D(C = POC, flux.up = depoPOC,  # upper boundary: flux 
                        dx = Grid, VF = porSolid,    # grid and volume fraction (1-por)
                        D = Dbio, v = v_adv)         # mixing (bioturbation) and advection

    # dissolved substances, VF = liquid volume fraction = porosity!
    tran.CO2 <- tran.1D(C = CO2, C.up = bwCO2,       # upper boundary: concentration
                        dx = Grid, VF = porLiquid,   # grid and volume fraction (por)
                        D = diffCO2, v = v_adv)      # diffusive mixing and advection
    
    tran.O2  <- tran.1D(C = O2, C.up = bwO2,
                        dx = Grid, VF = porLiquid,
                        D = diffO2, v = v_adv)
    
    tran.NH3 <- tran.1D(C = NH3, C.up = bwNH3,       
                        dx = Grid, VF = porLiquid,   
                        D = diffNH3, v = v_adv)      
    
    tran.NO3 <- tran.1D(C = NO3, C.up = bwNO3,       
                        dx = Grid, VF = porLiquid,   
                        D = diffNO3, v = v_adv)      
    
    tran.SO4 <- tran.1D(C = SO4, C.up = bwSO4,       
                        dx = Grid, VF = porLiquid,   
                        D = diffSO4, v = v_adv)      
    
    tran.H2S <- tran.1D(C = H2S, C.up = bwH2S,
                        dx = Grid, VF = porLiquid,   
                        D = diffH2S, v = v_adv)      

    tran.CH4 <- tran.1D(C = CH4, C.up = bwCH4,
                        dx = Grid, VF = porLiquid,   
                        D = diffCH4, v = v_adv)      

    # === reaction rates ===
    
    # [mol/m3 SOLID/d] (per volume of solid)
    AeroMin      <- rMin *  O2/(O2+kO2)                                     * POC   
    Denitri      <- rMin * kO2/(O2+kO2) *  NO3/(NO3+kNO3)                   * POC   
    SO4reduction <- rMin * kO2/(O2+kO2) * kNO3/(NO3+kNO3) *  SO4/(SO4+kSO4) * POC
    Methanogen   <- rMin * kO2/(O2+kO2) * kNO3/(NO3+kNO3) * kSO4/(SO4+kSO4) * POC
    
    Mineralisation  <- AeroMin  + Denitri + SO4reduction + Methanogen
    
    # [mol/m3 LIQUID/d] (per volume of liquid) 
    Nitri   <- rNit * O2/(O2+kO2) * NH3     
    Sox     <- rSox * O2/(O2+kO2) * H2S
    
    # === mass balances : dC/dt = transport + reactions ===
    
    # solid substances [mol/m3 SOLID/d]  
    dPOC.dt   <- ( tran.POC$dC                                    # transport
                 - AeroMin - Denitri - SO4reduction - Methanogen) # reactions, 
    
    # dissolved substances  [mol/m3 LIQUID/d]
    poro      <- porLiquid$mid
    f2Liquid  <- (1-poro)/poro  # to convert from /solid to /liquid
    
    dCO2.dt   <- ( tran.CO2$dC
                 + (AeroMin + Denitri + SO4reduction + 0.5*Methanogen)*f2Liquid )
    dO2.dt    <- ( tran.O2$dC - 2*Nitri - 2*Sox - AeroMin*f2Liquid )
    dNH3.dt   <- ( tran.NH3$dC - Nitri 
                 + 16/106*(AeroMin + Denitri + SO4reduction + Methanogen)*f2Liquid )  
    dNO3.dt   <- ( tran.NO3$dC + Nitri - 4/5*Denitri*f2Liquid )
    dSO4.dt   <- ( tran.SO4$dC - 0.5*SO4reduction*f2Liquid + Sox)
    dH2S.dt   <- ( tran.H2S$dC + 0.5*SO4reduction*f2Liquid - Sox)
    dCH4.dt   <- ( tran.CH4$dC + 0.5*Methanogen*f2Liquid)

    # === depth-integrated rates: [mol/m2 BULK/d] 
    TotalAero   <- sum(AeroMin       * Grid$dx * porSolid$mid)
    TotalDenit  <- sum(Denitri       * Grid$dx * porSolid$mid)
    TotalSO4red <- sum(SO4reduction  * Grid$dx * porSolid$mid)
    TotalCH4gen <- sum(Methanogen    * Grid$dx * porSolid$mid)
    TotalMin    <- sum(Mineralisation* Grid$dx * porSolid$mid)

    TotalNit    <- sum(Nitri         * Grid$dx * porLiquid$mid)
    TotalSox    <- sum(Sox           * Grid$dx * porLiquid$mid)
  
    return(list(c(dPOC.dt, dCO2.dt, dO2.dt, dNH3.dt, dNO3.dt, 
                  dSO4.dt, dH2S.dt, dCH4.dt),  # time-derivatives
          AeroMin        = AeroMin,            # rate profiles
          Denitri        = Denitri,
          SO4reduction   = SO4reduction,
          Methanogen     = Methanogen,
          Mineralisation = Mineralisation,
          Nitri          = Nitri,
          Sox            = Sox,
      
        # part of mineralisation due to aero, denit, SO4red, CH4genesis
          pAero          = TotalAero/TotalMin,  
          pDenit         = TotalDenit/TotalMin,
          pSO4red        = TotalSO4red/TotalMin,
          pCH4gen        = TotalCH4gen/TotalMin,

        # for creating budgets - all in [mol/m2 BULK/d]
          TotalAero      = TotalAero,           
          TotalDenit     = TotalDenit,
          TotalSO4red    = TotalSO4red,
          TotalCH4gen    = TotalCH4gen,
          TotalMin       = TotalMin,    
          TotalNit       = TotalNit,           
          TotalSox       = TotalSox,
          CO2.SWI.Flux   = tran.CO2$flux.up,    
          CO2.Deep.Flux  = tran.CO2$flux.down,  
          O2.SWI.Flux    = tran.O2$flux.up,    
          O2.Deep.Flux   = tran.O2$flux.down,  
          NH3.SWI.Flux   = tran.NH3$flux.up,    
          NH3.Deep.Flux  = tran.NH3$flux.down,  
          NO3.SWI.Flux   = tran.NO3$flux.up,    
          NO3.Deep.Flux  = tran.NO3$flux.down,  
          SO4.SWI.Flux   = tran.SO4$flux.up,    
          SO4.Deep.Flux  = tran.SO4$flux.down,  
          H2S.SWI.Flux   = tran.H2S$flux.up,    
          H2S.Deep.Flux  = tran.H2S$flux.down,  
          CH4.SWI.Flux   = tran.CH4$flux.up,    
          CH4.Deep.Flux  = tran.CH4$flux.down,  
          POC.SWI.Flux   = tran.POC$flux.up,   
          POC.Deep.Flux  = tran.POC$flux.down))
 })
}
```


# Model application

## Steady-state solution

Here, we find steady state solutions of the model for the three POC fluxes.
```{r}
std <- steady.1D (y=state.ini, func=Diamodel, parms=parms, 
                  nspec=nspec, dimens=N, names=names,
                  positive = TRUE)     # to have only positive values! 

p1 <- parms
p1["depoPOC"]  <- 20e-3 # [mol/m2/d]  POC deposition rate (flux at SWI)
std1 <- steady.1D (y=state.ini, func=Diamodel, parms=p1, 
                  nspec=nspec, dimens=N, names=names, positive = TRUE)

p2 <- parms
p2["depoPOC"]  <- 200e-3 # [mol/m2/d]  POC deposition rate (flux at SWI)
std2 <- steady.1D (y=state.ini, func=Diamodel, parms=p2, 
                  nspec=nspec, dimens=N, names=names, positive = TRUE)
```

### Plotting

First, we plot depth profiles of the concentrations of state variables, all runs in one graph.
```{r, fig.width=8, fig.height=8}
plot(std, std1, std2, 
     grid=Grid$x.mid, xyswap=TRUE, ylim=c(Length,0), lty=1, lwd=2, las=1,
     ylab="depth (m)", xlab=c("mol/m3 Solid", rep("mol/m3 Liquid",nspec-1)))
```

Next, we plot depth profiles of the process rates, all runs in one graph.
```{r, fig.width=8, fig.height=8}
plot(std, std1, std2, 
 which = c("AeroMin","Denitri","SO4reduction","Methanogen","Mineralisation","Nitri","Sox"),
 grid=Grid$x.mid, xyswap=TRUE, ylab="depth (m)", ylim=c(Length,0), lty=1, lwd=2, las=1,
 xlab=c(rep("molC/m3 Solid/d",5), "molN/m3 Liquid", "molS/m3 Liquid"))
```

Finally, we plot depth profiles of all mineralisation rates in one graph, separately for each model run. The graphs show that the organic matter is mineralised by aerobic mineralisation at the very top of the sediment (where $O_2$ is available), then by denitrification in a narrow band (where $NO_3^-$ is available but $O_2$ is depleted), then by sulphate reduction in the deeper sediment (where $SO_4^{2-}$ is available but $O_2$ and $NO_3^-$ are depleted).\footnote{Note that the latter two processes do not occur in the run with the lowest POC deposition flux.} Finally, methanogenesis occurs at depths where $O_2$, $NO_3^-$ and $SO_4^{2-}$ have been depleted and organic matter is still available (only visible in the run with the largest POC deposition flux). The plots thus illustrate the so-called *biogeochemical cascade*.

```{r, fig.width=8, fig.height=4}
# all process rates in one graph
par(mfrow=c(1,3))
s <- std
matplot(y=Grid$x.mid, x=cbind(s$AeroMin, s$Denitri, s$SO4reduction, s$Methanogen), 
        type="l", ylim=c(Length,0), col=1:4, lwd=2, lty=1,
        xlab="molC/m3 Solid/d", ylab="depth (m)")
s <- std1
matplot(y=Grid$x.mid, x=cbind(s$AeroMin, s$Denitri, s$SO4reduction, s$Methanogen), 
        type="l", ylim=c(Length,0), col=1:4, lwd=2, lty=1,
        xlab="molC/m3 Solid/d", ylab="depth (m)", main="biogeochemical cascade")
s <- std2
matplot(y=Grid$x.mid, x=cbind(s$AeroMin, s$Denitri, s$SO4reduction, s$Methanogen), 
        type="l", ylim=c(Length,0), col=1:4, lwd=2, lty=1,
        xlab="molC/m3 Solid/d", ylab="depth (m)")
legend("bottomright",legend=c("AeroMin", "Denitri", "SO4reduction", "Methanogen"),
       lwd=2, col=1:4, lty=1, bty="n")
```

## Contributions of mineralisation pathways

The following table shows that for the lowest POC deposition flux (``std``), organic matter mineralisation is primarily aerobic (almost 99% of total mineralisation). For the second largest POC deposition flux (``std1``), about 80% of organic matter is mineralised via sulphate reduction, while aerobic mineralisation and denitrification contribute by about 15% and 6%, respectively. Methanogenesis is significant (about 16%) only for the largest POC deposition flux (``std2``). Note that the depth-integrated total mineralisation (last line, values in $mol~C~m^{-2}~d^{-1}$) is equal to the POC deposition flux, showing that at the bottom of the modelled sediment column, all deposited POC was mineralised and converted to either $CO_2$ or $CH_4$.

```{r}
toselect <- c("pAero", "pDenit", "pSO4red", "pCH4gen", "TotalMin")
PATH     <- std[toselect]
PATH1    <- std1[toselect]
PATH2    <- std2[toselect]
CONTRIB  <- data.frame(std=unlist(PATH), std1=unlist(PATH1), std2=unlist(PATH2))
knitr::kable(CONTRIB, digits = 4)
```

## Oxygen budget

The following table ($O_2$ fluxes in $mol~O_2~m^{-2}~yr^{-1}$) shows that for the lowest POC deposition flux (``std``), about 2/3 of the sedimentary $O_2$ consumption occurs due to the aerobic mineralisation of the organic matter, while the remaining 1/3 is due to nitrification. In contrast, for the larger POC deposition fluxes, sedimentary $O_2$ consumption occurs primarily (50% or more) due to oxidation of the free sulphide produced by sulphate reduction, while the contributions of aerobic mineralisation and nitrification are lower.

```{r}
toselect <- c("TotalAero", "TotalNit", "TotalSox", "O2.SWI.Flux", "O2.Deep.Flux")
fac      <- 365  # to molO2/m2/yr
BUDGET   <- std[toselect]
BUDGET1  <- std1[toselect]
BUDGET2  <- std2[toselect]
O2budget <- rbind(std=unlist(BUDGET), std1=unlist(BUDGET1), std2=unlist(BUDGET2))*fac

O2BUDG   <- 
  data.frame(O2FluxIn      =  O2budget[,"O2.SWI.Flux"],
           O2FluxOut       = -O2budget[,"O2.Deep.Flux"],
           O2consByAeroMin = -O2budget[,"TotalAero"],
           O2consByNitri   = -O2budget[,"TotalNit"]*2, # 2 moles per mole of NH3 nitrified
           O2consBySox     = -O2budget[,"TotalSox"]*2) # 2 moles per mole of H2S oxidised

knitr::kable(O2BUDG, digits = 2)
```

The output below demonstrates that the difference between the $O_2$ fluxes at the domain boundaries and the sum of all depth-integrated rates of $O_2$ consumption is equal to (nearly) zero, consistent with the mass conservation law.
```{r}
rowSums(O2BUDG)
```

## Sulphur budget

The following table ($S$ fluxes in $mol~S~m^{-2}~yr^{-1}$) shows that if sulphate reduction is significant, the *total* amount of $SO_4^{2-}$ consumed in the sediment by sulphate reduction is always greater than the amount of $SO_4^{2-}$ that enters the sediment from the overlying water. This is only possible if $SO_4^{2-}$ is additionally produced within the sediment, i.e., if the free sulphide produced by sulphate reduction is *recycled* back to sulphate. The process responsible for this recycling is the aerobic oxidation of sulphide, which, as noted in the previous section, is a major contributor to the sedimentary oxygen consumption.

```{r}
toselect <- c("TotalSO4red", "TotalSox", "SO4.SWI.Flux", "SO4.Deep.Flux", 
              "H2S.SWI.Flux", "H2S.Deep.Flux")
fac <- 365  # to molS/m2/yr
BUDGET   <- std[toselect]
BUDGET1  <- std1[toselect]
BUDGET2  <- std2[toselect]
SBUDG    <- rbind(std=unlist(BUDGET), std1=unlist(BUDGET1), std2=unlist(BUDGET2))*fac
SBUDG[,"TotalSO4red"] <- SBUDG[,"TotalSO4red"]*0.5  # 0.5 moles of SO4 per mole of C
knitr::kable(SBUDG, digits = 4)
```

This conclusion is confirmed by the following output, which shows that the net flux of sulphate consumed in the sediment (which is equal to ``SO4.SWI.Flux - SO4.Deep.Flux``) together with the gross flux of sulphide production (``TotalSox``) is equal to the gross (total) rate of sulphate reduction in the sediment (``TotalSO4red``).
```{r}
Scheck <- cbind(SBUDG[,"SO4.SWI.Flux"]-SBUDG[,"SO4.Deep.Flux"] # net SO4 consumed
                + SBUDG[,"TotalSox"],                          # + H2S recycled
                SBUDG[,"TotalSO4red"])                         # gross SO4 consumed
Scheck
```
