---
title: "The Coupled Cycles of Carbon, Oxygen, Nitrogen and Sulfur in Marine Sediments"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Lubos Polerecky and Karline Soetaert, Utrecht University"
date: "July 2021"
output: 
  pdf_document:
    toc: no 
  header-includes: 
    \usepackage{float} 
    \floatplacement{figure}{ht} 
linestretch: 1.1
---

```{r child = 'diagenesis_Q.Rmd'}
```

\newpage
```{r child = 'diagenesis_A.Rmd'}
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Soetaert Karline (2009).  rootSolve: Nonlinear root finding, equilibrium and steady-state analysis of ordinary differential equations.  R-package version 1.6
  
Soetaert Karline, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential
  Equations in R: Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL
  http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09
  
Soetaert, Karline and Meysman, Filip (2012). Reactive transport in aquatic ecosystems: Rapid model
  prototyping in the open source software R Environmental Modelling & Software, 32, 49-60.

Soetaert Karline and Thomas Petzoldt (2020). marelac: Tools for Aquatic Sciences. R package
  version 2.1.10. https://CRAN.R-project.org/package=marelac
  
