---
title: "RTM course - Reversible reactions and equilibrium chemistry"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on rate expressions for reversible chemical reactions, including equilibrium formulations (pH, alkalinity)"
date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first. If you want, you can skip the first part (0:00-5:20) and jump straight to the more challenging second part (5:20-end).

### Video timeline
* 0:00 Elementary chemical reactions
* 5:20 Reversible reactions
* 8:07 Chemical equilibrium
* 9:30 Modelling processes occurring at vastly different time-scales
* 17:24 Application to ammonia dissociation and pH modelling

![](https://www.youtube.com/watch?v=loHVpLNQBZE&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=5)

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>

You can also read the following section(s) of the textbook:

* Chapter 8. Multiple time scales and Equilibrium Processes

(You can download each chapter of the textbook via the e-book University library for free!)

## Reversible reaction: rates

Assume that the following reversible chemical reaction is the only process affecting ammonia ($NH_3$), ammonium ($NH_4^+$) and protons ($H^+$).

$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

* The rate constant for the forward reaction is $k_f$.
* The rate constant for the backward reaction is $k_b$.

```{r ammonia_eq, echo=FALSE}
question("What is the correct expression for the rate of change in ammonia concentration?", type = "learnr_radio",
 answer(sprintf("$\\frac{d[NH_3]}{dt}=k_f \\cdot [NH_4^+] - k_b \\cdot [NH_3] \\cdot [H^+]$"), correct=TRUE),
 answer(sprintf("$\\frac{d[NH_3]}{dt}=k_f \\cdot [NH_4^+]$"), correct=FALSE),
 answer(sprintf("$\\frac{d[NH_3]}{dt}=k_f \\cdot [NH_4^+] - k_b \\cdot [NH_3]$"), correct=FALSE)
)
```

```{r proton_diffeq, echo=FALSE}
question("What is the correct expression for the rate of change in proton concentration?", type = "learnr_radio",
 answer(sprintf("$\\frac{d[H^+]}{dt}=k_f \\cdot [NH_4^+]$"), correct=FALSE),
 answer(sprintf("$\\frac{d[H^+]}{dt}=k_f \\cdot [NH_4^+] - k_b \\cdot [NH_3]$"), correct=FALSE),
 answer(sprintf("$\\frac{d[H^+]}{dt}=k_f \\cdot [NH_4^+] - k_b \\cdot [NH_3] \\cdot [H^+]$"), correct=TRUE)
)
```

## Reversible reaction: units

Consider the following reversible chemical reaction:

$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

* The rate constant for the forward reaction is $k_f$.
* The rate constant for the backward reaction is $k_b$.

```{r unitsKf, echo=FALSE}
msg <- sprintf("The unit of the rate is $mol\\,N\\, m^{-3}\\,s^{-1}$, and it is expressed as $R_f = k_f \\cdot [NH_4^+]$. With $[NH_4^+]$ in $mol\\,N\\,m^{-3}$, $k_f$ must be in units of $s^{-1}$.")
  
question(sprintf("What are the units of $k_f$?"), type = "learnr_radio",
 answer(sprintf("$(mol\\,N\\, m^{-3})^{-1}\\,s^{-1}$"), correct=FALSE),
 answer(sprintf("$s^{-1}$"), correct=TRUE),
 answer(sprintf("$mol\\,N\\,m^{-3}\\,s^{-1}$"), correct=FALSE),
 
 incorrect=msg, 
 correct=msg)
```

```{r unitsKb, echo=FALSE}
msg <- sprintf("The unit of the rate is $mol\\,N\\, m^{-3}\\,s^{-1}$, and it is expressed as $R_b = k_b \\cdot [NH_3]\\cdot [H^+]$. With $[NH_3]$ and $[H^+]$ in $mol\\,N\\,m^{-3}$, $k_b$ must be in units of $(mol\\,N\\,m^{-3})^{-1}\\,s^{-1}$.")
  
question(sprintf("What are the units of $k_b$?"), type = "learnr_radio",
 answer(sprintf("$(mol\\,N\\,m^{-3})^{-1}\\,s^{-1}$"), correct=TRUE),
 answer(sprintf("$s^{-1}$"), correct=FALSE),
 answer(sprintf("$mol\\,N\\,m^{-3}\\,s^{-1}$"), correct=FALSE),
 
 incorrect=msg,
 correct=msg)
```

## Equilibrium constant

Consider the following reversible chemical reaction:

$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

where the forward and backward rate constants are $k_f$ and $k_b$, respectively.

Assume that 

* an equilibrium has been reached
* the equilibrium constant for this reaction is denoted as $K_{N}$, with units of $mol~m^{-3}$
* the concentration of total ammonia is defined as $[\sum NH_x] = [NH_3] + [NH_4^+]$
* the equilibrium concentration of ammonia and ammonium is denoted as $[NH_3]^*$ and $[NH_4^+]^*$, respectively.

```{r KN, echo=FALSE}
msg <- sprintf("If you find it difficult to remember this formula and do not want to derive it, think about the units of $K_N$ $(mol\\,m^{-3})$, and the units of $k_b$ $((mol\\,m^{-3})^{-1}\\,s^{-1})$ and $k_f$ $(s^{-1})$. Units of both sides of the equation must be the same.")

question("Which of the following statements is true?", type = "learnr_radio",
 answer(sprintf("$K_{N} = \\frac{k_f}{k_b}$"), correct=TRUE),
 answer(sprintf("$K_{N} = \\frac{k_b}{k_f}$"), correct=FALSE),
 
 incorrect=msg, 
 correct=msg)
```

## Lump sums

Consider that the following reversible chemical reaction is the only reaction in the system:

$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

```{r eqStatements, echo=FALSE}
msg <- sprintf("The sums of chemical species whose combined time derivatives are equal to 0 are good candidates for lump-sums when formulating local-equilibrium reactions. In this case, the lump-sums are: $$ExcessProtons \\equiv [H^+] + [NH_4^+],$$ $$TotalAmmonia \\equiv [\\sum NH_x] = [NH_3] + [NH_4^+],$$ $$AmmoniaAlkalinity \\equiv [NH_3] - [H^+],$$ and $$ExcessNegativeCharge \\equiv - [NH_4^+] - [H^+].$$ Note that you only need two of these lump-sums to reformulate the problem.")

question("Which of the following statements is true?", type = "learnr_checkbox",
 answer(sprintf("$\\frac{d([H^+]+[NH_4^+])}{dt} =0$"), correct=TRUE),
 answer(sprintf("$\\frac{d([NH_3]+[NH_4^+]}{dt} =0$"), correct=TRUE),
 answer(sprintf("$\\frac{d([NH_3]-[NH_4^+]}{dt} =0$"), correct=FALSE),
 answer(sprintf("$\\frac{d([NH_3]+[H^+]}{dt} =0$"), correct=FALSE),
 answer(sprintf("$\\frac{d([NH_3]-[H^+]}{dt} =0$"), correct=TRUE),
 
 allow_retry = TRUE,
 incorrect=msg,
 correct=msg)
```

## Equilibrium formulation

Consider the following reversible chemical reaction:

$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

where the forward and backward rate constants are $k_f$ and $k_b$, respectively.

Assume that 

* an equilibrium has been reached
* the equilibrium constant for this reaction is denoted as $K_{N}$, with units of $mol~m^{-3}$
* the concentration of total ammonia is defined as $[\sum NH_x] = [NH_3] + [NH_4^+]$
* the equilibrium concentration of ammonia and ammonium is denoted as $[NH_3]^*$ and $[NH_4^+]^*$, respectively.

```{r eqStatements2, echo=FALSE}
msg <- sprintf("If you find it difficult to remember which formula applies for which species, a mnemonic we use is that the positively charged species has protons in the numerator.")

question("Which of the following statements is true?", type = "learnr_checkbox",
 answer(sprintf("$[NH_3]^* = \\frac{K_N}{K_N+[H^+]}$"), correct=FALSE),
 answer(sprintf("$[NH_3]^* = \\frac{K_N}{K_N+[H^+]}\\cdot [\\sum NH_x]$"), correct=TRUE),
 answer(sprintf("$[NH_3]^* = \\frac{[H^+]}{K_N+[H^+]}\\cdot [\\sum NH_x]$"), correct=FALSE),
 answer(sprintf("$[NH_4^+]^* = \\frac{K_N}{K_N+[H^+]}$"), correct=FALSE),
 answer(sprintf("$[NH_4^+]^* = \\frac{K_N}{K_N+[H^+]}\\cdot [\\sum NH_x$]"), correct=FALSE),
 answer(sprintf("$[NH_4^+]^* = \\frac{[H^+]}{K_N+[H^+]}\\cdot [\\sum NH_x$]"), correct=TRUE),
 
 incorrect=msg, 
 correct=msg)
```

## Bjerrum plot

Bjerrum plots show how the equilibrium concentration (or fraction) of the various ionic forms of a lump-sum (aka *equilibrium speciation*) changes as a function of pH. For ammonia ($NH_3$), we have $K_N = 5.380823\times 10^{-10}~mol~kg^{-1}$. Thus, we obtain

```{r, fig.width=6, fig.height=4}
KN     <- 5.380823e-10
pH     <- seq(from=2, to=12, length.out=50)
H      <- 10^{-pH}
NH3    <- KN/(KN+H)
plot(pH, NH3, col=1, type = "l", main = "Bjerrum plot", ylab="NH3 / Sum(NHx)")
abline(v=-log10(KN), lty=2)
```

Complete the R code to add the fraction of ammonium ($NH_4^+$) to the Bjerrum plot.

```{r amm, exercise=TRUE, exercise.cap = "Bjerrum plot with NH3 and NH4+", exercise.lines = 6}
plot(pH, NH3, type = "l", ylab = "fractions", main = "Bjerrum plot: NH4+ <-> NH3 + H+")

NH4 <-    ### ADD THE EQUATION HERE
  
lines(pH, NH4, col=2)
abline(v=-log10(KN), lty=2)
legend("left", legend=c("NH3","NH4+","pKN"), col=c(1,2,1), lty=c(1,1,2))
```

## Two reactions: equilibrium

Consider the following two reversible chemical reactions:

$$
Reaction~1: \qquad H_2CO_3 \leftrightarrow HCO_3^- + H^+
$$


$$
Reaction~2: \qquad HCO_3^- \leftrightarrow CO_3^{2-} + H^+
$$

Equilibrium constants for the reactions 1 and 2 are $K_1$ and $K_2$, respectively. Assume that the concentration of the total dissolved inorganic carbon ($[DIC] = [H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$) is constant and equal to [DIC].


```{r eq2Reactions, echo=FALSE}
msg <- sprintf("The equilibrium equations all depend on the DIC concentration, so the second option cannot be true. Also, the sum of $[H_2CO_3]$ (carbonic acid), $[HCO_3^-]$ (bicarbonate) and $[CO_3^{2-}]$ (carbonate) must be equal to $[DIC]$ (by definition), so the third equation cannot be true as the numerator is not part of the denominator. As for carbonate, the correct formula can be remembered as follows: carbonate is the most negatively charged ion, so it has in the numerator the least charged combination $(K_1\\cdot K_2)$. The other species, which contain one or two extra protons, will need to have one or two proton concentrations in the numerator, respectively.")

question("Which of the following equations correctly describe the relationships between the equilibrium concentrations of DIC, carbonic acid, bicarbonate, carbonate and protons?", type = "learnr_checkbox",
 answer(sprintf("$[HCO_3^-] = \\frac{K_1 \\cdot [H^+]}{[H^+] \\cdot [H^+] + K_1 \\cdot [H^+] + K_1 \\cdot K_2} \\cdot[DIC]$"), correct=TRUE),
 answer(sprintf("$[DIC] = \\frac{K_1 \\cdot [H^+]}{[H^+] \\cdot [H^+] + K_1 \\cdot [H^+] + K_1 \\cdot K_2}\\cdot [HCO_3^-]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] = \\frac{K_2 \\cdot [H^+]}{[H^+] \\cdot [H^+] + K_1 \\cdot [H^+] + K_1 \\cdot K_2}\\cdot [DIC]$"), correct=FALSE),
 answer(sprintf("$[CO_3^{2-}] = \\frac{[H^+] \\cdot [H^+]}{[H^+] \\cdot [H^+] + K_1 \\cdot [H^+] + K_1 \\cdot K_2}\\cdot [DIC]$"), correct=FALSE),
 answer(sprintf("$[CO_3^{2-}] = \\frac{K_1 \\cdot K_2}{[H^+] \\cdot [H^+] + K_1 \\cdot [H^+] + K_1 \\cdot K_2}\\cdot [DIC]$"), correct=TRUE),
 answer(sprintf("$[H_2CO_3] = \\frac{K_1 \\cdot [H^+]}{[H^+] \\cdot [H^+] + K_1 \\cdot [H^+] + K_1 \\cdot K_2}\\cdot [DIC]$")),
 answer(sprintf("$[H_2CO_3] = \\frac{[H^+] \\cdot [H^+]}{[H^+] \\cdot [H^+] + K_1 \\cdot [H^+] + K_1 \\cdot K_2}\\cdot [DIC]$"), correct=TRUE),
 
 allow_retry = TRUE,
 incorrect=msg, 
 correct=msg)
```

## Two reactions: Bjerrum plot

Similar to the total ammonia, we can display the Bjerrum plots for the equilibrium speciation of DIC as a function of pH. 

Assuming that the first and second dissociation constants for the carbonic acid are, respectively, $K_1 = 10^{-6}~mol~kg^{-1}$ and $K_2 = 10^{-9.5}~mol~kg^{-1}$ (i.e., $pK_1 = -log_{10}(K_1) = 6$ and $pK_2 = -log_{10}(K_2) = 9.5$), we can calculate the equilibrium concentrations of the carbonic acid, bicarbonate and carbonate using the following R-code. Study the code and then run it to see the result.

```{r dic, exercise=TRUE, exercise.cap = "Bjerrum plot for DIC", exercise.lines = 19, fig.width=6, fig.height=4}
pK1    <- 6
pK2    <- 9.5
pH     <- seq(from=2, to=12, length.out=50)
H      <- 10^{-pH}
K1     <- 10^{-pK1}
K2     <- 10^{-pK2}
H2CO3  <-  H*H / (H*H+K1*H+K1*K2)
HCO3   <- K1*H / (H*H+K1*H+K1*K2)
CO3    <- K1*K2/ (H*H+K1*H+K1*K2)
plot(pH, H2CO3, col=1, type = "l", main = "Bjerrum plot - DIC", ylab="fraction")
lines(pH, HCO3, col=2)
lines(pH, CO3,  col=3)
abline(v=pK1,   lty=2)
abline(v=pK2,   lty=3)
legend("left", legend=c(expression(paste("H"[2],"CO"[3])),        #H2CO3
                        expression("HCO"[3]^"-"),                 #HCO3-
                        expression("CO"[3]^"2-"),"pK1","pK2"), 
       col=c(1,2,3,1,1), lty=c(1,1,1,2,3))
```

The graph above shows that in an equilibrium at $pH=8$, which is approximately the pH of seawater, bicarbonate is the dominant DIC species. You may be familiar with this idea from your aquatic chemistry or marine chemistry class.

Experiment with the above R-code to explore how the DIC speciation shifts with changing values of $K_1$ and $K_2$. These parameters change, for instance, with changing temperature or salinity.

## Two reactions: lump-sums

Consider the following two reversible chemical reactions:

$$
Reaction~1: \qquad H_2CO_3 \leftrightarrow HCO_3^- + H^+
$$


$$
Reaction~2: \qquad HCO_3^- \leftrightarrow CO_3^{2-} + H^+
$$

To make use of the local equilibrium, we need to find combinations of species that are not affected by the above reactions.

```{r eq2Lumpsum, echo=FALSE}
msg <- sprintf("The lump sum $ALK_C=[HCO_3^-] + 2\\cdot [CO_3^{2-}] - [H^+]$ is unaffected by reaction 1 because: production of 1 $HCO_3^-$ causes production of 1 $ALK_C$, production of $H^+$: -1 $ALK_C$. It is also unaffected by reaction 2 (consumption of 1 $HCO_3^-$: -1 $ALK_C$, production of 1 $H^+$: -1 $ALK_C$, production of 1 $CO_3^{2-}$: +2 $ALK_C$). This lump sum is called the *carbonate alkalinity* or *excess negative charge*. The fifth lump sum is the *total dissolved inorganic carbon (DIC)*. The last lump sum $(2\\cdot[H_2CO_3] + [HCO_3^-] + [H^+])$ would also be suitable - it accounts for the excess protons - but it is never used in models.")

question("Which of the following lump sums would be suitable to reformulate a model that involves these equilibrium reactions?", type = "learnr_checkbox",
 answer(sprintf("$[HCO_3^-] + 2\\cdot [CO_3^{2-}] - [H^+]$"), correct=TRUE),
 answer(sprintf("$[HCO_3^-] + [CO_3^{2-}] - [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] - [CO_3^{2-}] + [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] - 2\\cdot [CO_3^{2-}] + [H^+]$"), correct=FALSE),
 answer(sprintf("$[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$"), correct=TRUE),
 answer(sprintf("$2\\cdot[H_2CO_3] + [HCO_3^-] + [H^+]$"), correct=TRUE),
 
 incorrect=msg, 
 correct=msg)
```

## Lump sums revisited

Consider the following two reversible chemical reactions:

$$
Reaction~1: \qquad H_2CO_3 \leftrightarrow HCO_3^- + H^+ \\
Reaction~2: \qquad HCO_3^- \leftrightarrow CO_3^{2-} + H^+
$$

The most straight-forward (but rather elaborate) way to determine whether a given sum of species is suitable as a lump sum (i.e., it is not affected by the reversible reactions) is to expand its time-derivative. We will demonstrate it here.

We denote by $R_{1f}$ and $R_{1b}$ the forward and backward reaction rate of reaction 1 (in $mol~m^{-3}~s^{-1}$), and by $R_{2f}$ and $R_{2b}$ the forward and backward reaction rate of reaction 2 (in $mol~m^{-3}~s^{-1}$), respectively. Thus, we can write:

$$\frac{d[H_2CO_3]}{dt}=-R_{1f}+R_{1b}$$
$$\frac{d[HCO_3^-]}{dt}=R_{1f} - R_{1b} - R_{2f} + R_{2b}$$
$$\frac{d[H^+]}{dt}=R_{1f} - R_{1b} + R_{2f} - R_{2b}$$
$$\frac{d[CO_3^{2-}]}{dt}=R_{2f} - R_{2b}$$
If we define the lump sum *carbonate alkalinity* as $$[ALK_C] = [HCO_3^-] + 2\cdot [CO_3^{2-}] - [H^+],$$ it is straight-forward to see, by direct substitution, that its time derivative is zero: $$\frac{d[ALK_C]}{dt} = \frac{d[HCO_3^-]}{dt} + 2\cdot \frac{d[CO_3^{2-}]}{dt} - \frac{d[H^+]}{dt} = 0.$$

## Combined reversible and irreversible reactions

Consider the following chemical reactions:

$$
\begin{array}{rl}
Reaction~1: & H_2CO_3 \leftrightarrow HCO_3^- + H^+ \\
Reaction~2: & HCO_3^- \leftrightarrow CO_3^{2-} + H^+\\
Reaction~3: & NH_4^+ \leftrightarrow NH_3 + H^+\\
Reaction~4: & HNO_3 \leftrightarrow NO_3^- + H^+\\
Reaction~5: & CH_2O(NH_3)_{NC} + O_2 \rightarrow H_2CO_3 + NC \cdot NH_3\\
Reaction~6: & NH_3 + 2\cdot O_2 \rightarrow NO_3^- + H^+ + H_2O
\end{array}
$$

* Assume that reactions 1–4 are fast and reversible, whereas reactions 5–6 correspond to slow irreversible processes (mediated by microbes). 
* Reaction 5, which corresponds to *aerobic mineralization* of organic matter, occurs at a rate $R_{aero}$.
* Reaction 6, which corresponds to *nitrification*, occurs at a rate $R_{nit}$.

```{r eqmultLumpsum, echo=FALSE}
msg <- sprintf("The lump sums should only include species that participate in the fast reversible reactions. This rules out answer 2 and 3.  The sum $[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$ is the total dissolved inorganic carbon (DIC) concentration comprising species that only feature in the first 2 reactions. DIC is conserved when those 2 reactions occur. Similarly, $[NH_4^+] + [NH_3]$ and $[NO_3^-] + [HNO_3]$ are the total ammonium and total nitrate concentrations, respectively, which only feature, and are conserved, in reactions 3 and 4. One more lump sum is needed to constrain the proton concentration. As protons feature in reactions 1, 2, 3, and 4, this lump sum should be conserved by all these equilibrium reactions. The sum $[HCO_3^-] + 2\\cdot [CO_3^{2-}] - [H^+]$ is not suitable, as it is not conserved in reactions 3 and 4. Only the last two lump sums are suitable, as can be verified by a similar approach (direct substitution of time derivatives) as in the previous question.")
  
question("Which of the following lump sums would be suitable to reformulate a model that involves the fast reversible reactions 1-4?", type = "learnr_checkbox",
 answer(sprintf("$[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$"), correct=TRUE),
 answer(sprintf("$[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}] + [CH_2O]$"), correct=FALSE),
 answer(sprintf("$[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}] - [CH_2O]$"), correct=FALSE),
 answer(sprintf("$[NH_4^+] + [NH_3]$"), correct=TRUE),
 answer(sprintf("$[NO_3^-] + [NH_3]$"), correct=FALSE),
 answer(sprintf("$[NO_3^-] + [HNO_3]$"), correct=TRUE),
 answer(sprintf("$[HCO_3^-] + 2\\cdot [CO_3^{2-}] - [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] + 2\\cdot [CO_3^{2-}] +[NH_3] - [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] + 2\\cdot [CO_3^{2-}] +[NO_3^-] - [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] + 2\\cdot [CO_3^{2-}] +[NO_3^-] + [NH_3] - [H^+]$"), correct=TRUE),
 answer(sprintf("$[HCO_3^-] + 2\\cdot [CO_3^{2-}] -[HNO_3] + [NH_3] - [H^+]$"), correct=TRUE),
 
 allow_retry = TRUE,
 incorrect=msg,
 correct=msg)
```

## Combined fast reversible and slow irreversible reactions: the full model
 
Consider the following reactions:

$$
\begin{array}{rl}
Reaction~1: & H_2CO_3 \leftrightarrow HCO_3^- + H^+ \\
Reaction~2: & HCO_3^- \leftrightarrow CO_3^{2-} + H^+\\
Reaction~3: & NH_4^+ \leftrightarrow NH_3 + H^+\\
Reaction~4: & HNO_3 \leftrightarrow NO_3^- + H^+\\
Reaction~5: & CH_2O(NH_3)_{NC} + O_2 \rightarrow H_2CO_3 + NC \cdot NH_3\\
Reaction~6: & NH_3 + 2\cdot O_2 \rightarrow NO_3^- + H^+ + H_2O
\end{array}
$$

* Assume that reactions 1–4 are fast and reversible, whereas reactions 5–6 correspond to slow irreversible processes (mediated by microbes). 
* Reaction 5, which corresponds to *aerobic mineralization* of organic matter, occurs at a rate $R_{aero}$.
* Reaction 6, which corresponds to *nitrification*, occurs at a rate $R_{nit}$.

Use the following lump sums (TA stands for total alkalinity):

$$
[DIC] = [H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]\\
[NH_x] = [NH_4^+] + [NH_3]\\
[H_xNO_3] = [HNO_3] + [NO_3^-]\\
[TA] = [HCO_3^−] + 2\cdot [CO_3^{2−}] + [NH_3] + [NO_3^-] − [H^+]
$$

```{r eqmultirrev, echo=FALSE}
msg <- sprintf("The slow reaction only affects the lump sums, which rules out answer 2, 4 and 5.  Answers 1, 6, and 7 describe the effects on DIC, $NH_x$, and $H_xNO_3$, and are correct. TA as defined here is affected by two processes. First, aerobic organic matter mineralization produces $NH_3$ and thus increases TA at a rate $NC\\cdot R_{aero}$. Second, nitrification leads to the consumption of $NH_3$ (-1TA), production of $NO_3^-$ (+1TA), and production of protons (-1TA, because $[H^+]$ contributes negatively to TA). Thus, nitrification yields a net TA consumption of 1 $(-1+1-1=-1)$.")
  
question("Which of the following mass balance equations correctly account for the effects of aerobic mineralization and nitrification?", type = "learnr_checkbox",
 answer(sprintf("$\\frac{d[DIC]}{dt} = R_{aero}$"), correct=TRUE),
 answer(sprintf("$\\frac{d[H_2CO_3]}{dt} = R_{aero}$"), correct=FALSE),
 answer(sprintf("$\\frac{d[CH_2O(NH_3)_{NC}]}{dt} = -R_{aero}$"), correct=TRUE),
 answer(sprintf("$\\frac{d[NH_3]}{dt} = NC \\cdot R_{aero}$"), correct=FALSE),
 answer(sprintf("$\\frac{d[NH_3]}{dt} = NC \\cdot R_{aero} - R_{nit}$"), correct=FALSE),
 answer(sprintf("$\\frac{d[NH_x]}{dt} = NC \\cdot R_{aero} - R_{nit}$"), correct=TRUE),
 answer(sprintf("$\\frac{d[H_xNO_3]}{dt} = R_{nit}$"), correct=TRUE),
 answer(sprintf("$\\frac{d[TA]}{dt} = R_{aero} + R_{nit}$"), correct=FALSE),
 answer(sprintf("$\\frac{d[TA]}{dt} = NC \\cdot R_{aero} + R_{nit}$"), correct=FALSE),
 answer(sprintf("$\\frac{d[TA]}{dt} = NC \\cdot R_{aero} - R_{nit}$"), correct=TRUE),
 answer(sprintf("$\\frac{d[TA]}{dt} = 0$"), correct=FALSE),
 
 allow_retry = TRUE,
 incorrect=msg,
 correct=msg)
```

## More exercises

If you finished this tutorial but some of your answers were incorrect, you are encouraged to do the quiz again. You can do this by clicking on <span style="color:red">**Start Over**</span> at the bottom of the menu to the left.

To learn more about modelling of chemical reactions --- and specifically of chemical reactions under the *local equilibrium* condition --- try to solve additional exercises available in the R-package RTM. To display these exercises, type one of the following commands in the R-console:

```
require(RTM)
RTMexercise("equilibriumNH3")
RTMexercise("equilibriumHCO3")
RTMexercise("equilibriumOMD")
```

## Textbook and Slides

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>