---
title: "Earth's global C cycle: the $CO_2$ Problem and Mitigation Strategies"
author: "Lubos Polerecky and Karline Soetaert"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
date: "February 2021"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: no
  word_document:
    toc: yes
---

```{r child = 'carbonCycle_Q.Rmd'}
```

```{r child = 'carbonCycle_A.Rmd'}
```


# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for
  Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential Equations in R:
  Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL
  http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09

Lubos Polerecky and Karline Soetaert (2021). Response of Systems to a Perturbation from an Equilibrium --- Part I & II. ``RTMreader("perturbation_I")`` and ``RTMreader("perturbation_II")``