---
title: "RTM course - partitioning between phases"
author: "Lubos Polerecky and Karline Soetaert"
date: "June 2021"
description: "video and exercises on rate expressions for dissolution/precipitation, degassing reactions, including mineral and gas solubility"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first.

### Video timeline
* 0:00 Mineral dissolution and precipitation
* 4:38 Gas dissolution and degassing
* 6:12 Extra information about mineral solubility
* 8:07 Extra information about gas solubility

![](https://www.youtube.com/watch?v=CPme0Qz3L7Y&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=8)

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>

There are no corresponding sections in the textbook.

## Mineral dissolution - quiz

There are $N$ spherical amorphous silica particles suspended in a well-mixed volume of water. The change over time in the concentration of dissolved silica is described by the following differential equation:

$$
\frac{dSi}{dt} = -k_p \cdot \frac {A}{V} \cdot (Si - Si_{eq})
$$
where $Si$ and $Si_{eq}$ are, respectively, the instantaneous and equilibrium concentrations of dissolved silica (both in $mol~Si~m^{-3}$).

```{r rate_constant, echo=FALSE}
msg <- sprintf("$k_p$ is the rate constant of the backward reaction, i.e., Si precipitation. It describes the amount of Si (moles) that precipitate per unit area of the mineral surface $(m^2)$, per unit time $(yr)$, and per unit concentration of dissolved silica $(mol\\,m^{-3})$. Thus, its unit is $mol/m^2/yr/(mol\\,m^{-3})$, which is equal to $m\\,yr^{-1}$.")

question(sprintf("What is the meaning and unit of the rate constant $k_p$?"), 
         type = "learnr_radio",
 answer(sprintf("$k_p$ is the dissolution rate constant, in $m~yr^{-1}$")),
 answer(sprintf("$k_p$ is the precipitation rate constant, in $mol~m^{-2}~yr^{-1}$")),
 answer(sprintf("$k_p$ is the precipitation rate constant, in $m~yr^{-1}$"), correct=TRUE),
 answer(sprintf("$k_p$ is the dissolution rate constant, in $mol~m^{-2}~yr^{-1}$")),
 
 allow_retry = TRUE,
 incorrect=msg,
 correct=msg)
```

```{r volume, echo=FALSE}
msg <- sprintf("$V$ has nothing to do with the volume of the silica particles! $V$ is the volume of the well-mixed *water* in which the silica particles are suspended.")

question(sprintf("What is the meaning and unit of $V$ in the above equation?"), 
         type = "learnr_radio",
         answer(sprintf("$V$ is the volume of an *individual* silica particle, in $m^{3}$, calculated from the radius of the individual particle as $V=\\frac{4}{3}\\pi r^3$.")),
         answer(sprintf("$V$ is the *total* volume of the silica particles, in $m^{3}$, calculated from the radius of the individual particle as $V=N\\cdot \\frac{4}{3}\\pi r^3$.")),
         answer(sprintf("$V$ is the volume of well-mixed water containing the silica particles, in $m^3$."), correct=TRUE),
 
 allow_retry = TRUE,
 incorrect=msg,
 correct=msg)
```

```{r area, echo=FALSE}
msg <- sprintf("$A$ is the *total* area of the silica particles. It corresponds to the overall surface area where the reaction (silica dissolution or precipitation) occurs.")

question(sprintf("What is the meaning and unit of $A$ in the above equation?"), 
         type = "learnr_radio",
         answer(sprintf("$A$ is the area of an *individual* silica particle, in $m^{2}$, calculated from the radius of the individual particle as $A=4\\pi r^2$.")),
         answer(sprintf("$A$ is the *total* area of all silica particles, in $m^{2}$, calculated from the radius of the individual particle as $A=N\\cdot 4\\pi r^2$."), correct=TRUE),

 allow_retry = TRUE,
 incorrect=msg,
 correct=msg)
```

```{r eq_conc, echo=FALSE}
msg <- sprintf("The answer is $k_d/k_p$. Tip: if you do not remember whether the formula should be $k_p/k_d$ or $k_d/k_p$, think about the units. Concentrations are in $mol~m^{-3}$, and this unit is obtained with $k_d/k_p$, not with $k_p/k_d$.")

question(sprintf("Assume that the rate constant describing the dissolution of a mineral in water is $k_d = 2\\,mol\\,m^{-2}\\,yr^{-1}$, and the rate constant describing the precipitation of that mineral is $k_p = 5\\,m\\,yr^{-1}$. What is the equilibrium concentration of the dissolved mineral in water? Note that there are *two* correct answers in the list! Also note that $M$ (molar) stands for $mol~L^{-1}$."), type = "learnr_checkbox",
 answer(sprintf("0.4 $mol\\,m^{-3}$"), correct=TRUE),
 answer(sprintf("2.5 $mol\\,m^{-3}$")),
 answer(sprintf("10 $mol\\,m^{-3}$")),
 answer(sprintf("40 $\\mu mol\\,L^{-1}$")),
 answer(sprintf("2.5 $mmol\\,L^{-1}$")),
 answer(sprintf("10 $mM$")),
 answer(sprintf("400 $\\mu M$"), correct=TRUE),
 
 allow_retry = TRUE,
 incorrect=msg,
 correct=msg)
```


## Mineral dissolution - exercise

Consider $N=800$ spherical amorphous silica particles suspended in a well-mixed water volume of $V=1~L$. Assume that

- initially, the concentration of the dissolved silica (silicic acid) in the water is $Si_{ini} = 100~\mu M$;
- all particles have the same *diameter* $D_{ini}=0.1~mm$;
- rate constant for silica dissolution is $k_d = 1~mol~m^{-2}~yr^{-1}$;
- rate constant for silica precipitation is $k_p = 0.8~m~yr^{-1}$;
- bulk silica density is $2200~kg~m^{-3}$, molar weight of silica is $60~g~mol^{-1}$.

```{r dSidt_pars, echo=FALSE}
N    <- 800    # number of particles, -
V    <- 1e-3   # volume of water, m3
Dini <- 0.1e-3 # diameter of an individual particle, m
kd   <- 1      # dissolution rate constant, mol m-2 yr-1
kp   <- 0.8    # precipitation rate constant, m yr-1
Cini <- 100e-3 # initial concentration, mol m-3
dens <- 2200   # silica density, kg m-3
MW   <- 60e-3  # molar weight of silica, kg mol-1
dSi.ini.dt <- -kp * N * 4*pi*(Dini/2)^2 / V * (Cini - kd/kp)
```

```{r dSidt_ini, echo=FALSE}
msg <- sprintf("Tips: 1. Use the formula for the rate of change in dissolved Si concentration, as explained earlier in this tutorial. 2. Calculate the total surface area of the silica particles (note that the *diameter*, not the *radius*, of the silica particles is provided). 3. Calculate the equilibrium concentration of dissolved Si based on the rate constants of the forward and backward processes. 4. Convert *all* quantities into SI units. In this way, you will be less likely to make an error in the final calculation step. That is, use moles, meters (and their powers), years (although seconds are the SI units for time, using years is ok in this exercise). 5. Carry out the calculation on your calculator or in R. Note that the information about the density and molar weight of silica is irrelevant in this exercise.")

question("What is the *initial* rate of increase in the silicic acid concentration (dSi/dt) in the water with the silica particles?", 
         type = "learnr_radio",
         answer(sprintf("23 $mmol~L^{-1}~yr^{-1}$")),
         answer(sprintf("23 $\\mu mol~L^{-1}~yr^{-1}$"), correct=TRUE),
         answer(sprintf("0.092 $mol~m^{-3}~yr^{-1}$")),
         answer(sprintf("55200 $mol~m^{-3}~yr^{-1}$")),
         answer(sprintf("27600 $mol~m^{-3}~yr^{-1}$")),
         answer(sprintf("69 $mol~L^{-1}~yr^{-1}$")),
         answer(sprintf("69 $mol~m^{-3}~yr^{-1}$")),
         answer(sprintf("92 $\\mu mol~L^{-1}~yr^{-1}$")),
         
 allow_retry = TRUE,
 incorrect = msg,
 correct = msg
)
```

```{r dSidt_t, echo=FALSE}
msg <- sprintf("The dissolution rate depends on the *difference* between the instantaneous and equilibrium concentration of silicic acid $(Si-Si_{eq})$. Because silica particles are dissolving, the silicic acid concentrations *increases* over time. Thus, the difference $Si-Si_{eq}$ is becoming smaller, implying that the third answer is correct.")

question("Which statement is true?", type = "learnr_radio",
 answer("The dissolution rate of silica will increase over time."),
 answer("The dissolution rate of silica will stay constant over time."),
 answer("The dissolution rate of silica will decrease over time.", correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## Air-water exchange - quiz

```{r statements2, echo=FALSE}
msg <- sprintf("Tips: 1. $O_2$ concentration in the water column increases due to net dissolution of $O_2$ across the air-water interface. 2. The area of this interface is $A$, the volume of the water column is $V=A\\cdot h$. 3. The term $A/V$ in the equation for $dO_2/dt$ is therefore equal to $1/h$. 4. The rate is therefore calculated according to $k_p/h\\cdot (O_{2,eq} - 0)$. 5. Make sure that you convert all quantities to SI units before doing the final calculation.")

question(sprintf("A well-mixed *anoxic* water column is in contact with the atmosphere. What is the initial rate of increase in the $O_2$ concentration in the water? Piston velocity is $k_p = 1\\,m\\,d^{-1}$, $O_2$ solubility in the water is $300\\,\\mu mol\\,L^{-1}$, water column height is $h=50\\,cm$."), type = "learnr_radio",
 answer(sprintf("$300\\, \\mu mol\\,L^{-1}\\,d^{-1}$")),
 answer(sprintf("$150\\, nmol\\,L^{-1}\\,d^{-1}$")),
 answer(sprintf("$600\\, nmol\\,L^{-1}\\,d^{-1}$")),
 answer(sprintf("$600\\, \\mu mol\\,L^{-1}\\,d^{-1}$"), correct=TRUE),
 
 allow_retry = TRUE,
 incorrect=msg,
 correct=msg)
```

## Air-water exchange - exercise

You plan to install in your garden a pond with goldfish. By swimming in the pond, the fish will mix the water in the pond, and by respiration they will decrease the dissolved $O_2$ concentration in the pond. This decrease in dissolved $O_2$ will be compensated by air-water exchange. 

Assume that 

- goldfish respiration is unaffected (fish are 'happy') if the dissolved $O_2$ concentration does not decrease below 80% of the air saturated value;
- piston velocity is $k_p=1~m~d^{-1}$;
- area of the pond is $4~m^2$;
- the average height of the water column is $1~m$;
- Henry’s constant for dissolved $O_2$ is $K_H=770~L~Atm~mol^{-1}$;
- molar fraction of $O_2$ in the atmosphere is 21%;
- the total atmospheric pressure is $1~Atm$;
- the average respiration rate per individual goldfish is $400~\mu mol~O_2~ind^{-1}~d^{-1}$.

```{r Fish-setup, echo=FALSE, eval=FALSE}
pSat  <- 0.8    # -
kp    <- 1      # m/d
h     <- 1      # m
KH    <- 770e-3 # m3 atm/mol
P     <- 1      # atm, total atmospheric pressure
pO2   <- 0.21   # molar fraction of O2
Resp  <- 400e-6 # mol/fish/d
Area  <- 4      # m2
# solution (one formula)
Nmax <- kp*Area * (1-pSat)* P*pO2/KH/Resp
# solution in parts
O2sat        <- pO2/KH*P
airwater     <- kp*(O2sat - pSat*O2sat)
airwaterTot  <- airwater*Area
N           <- airwaterTot/Resp
```

```{r Nfish, echo=FALSE}
msg <- sprintf("Tips: 1. $O_2$ concentration in the pond changes due to two competing processes: (a) net dissolution of atmospheric $O_2$, and (b) $O_2$ removal due to fish respiration. 2. The answer to the question is found by considering that the system reaches an *equilibrium*, i.e., $O_2$ concentration in the pond remains *constant* and equal to 80%s of the equilibrium value. 3. Use your knowledge of physical chemistry (Henry's law) to quantify the equilibrium $O_2$ concentration (also called $O_2$ solubility) at the given partial pressure of $O_2$ in the atmosphere. (Note that the unit of Henry's constant provides some clues on how to do this.) 4. Use the formula explained earlier in this tutorial to quantify the net rate of $O_2$ dissolution across the air-water interface for the given $O_2$ concentration in the water. 5. Convert this value to the rate of $O_2$ dissolution *per pond* per day. 6. Next, calculate the removal of $O_2$ in the pond due to fish respiration, based on the respiration rate of an individual fish. 7. Combine everything based on equation you derived in point 2. This will yield an equation from which you can find the number of fish in the pond. Make sure you use SI units for all quantities when carrying out the last calculation steps. This will make it less likely that you will make a calculation error.", "%")

question("What is the maximum number of 'happy' goldfish in the pond? ", 
         type = "learnr_radio",
         answer("25 fish"), 
         answer("545 fish", correct=TRUE), 
         answer("252 fish"),
         answer("1090 fish"),
         answer("100 fish"),

 allow_retry = TRUE,
 incorrect = msg,
 correct = msg
)
```

And one extra question:

```{r Fish_point_height, echo=FALSE}
msg <- "It may be somewhat counter-intuitive, but the answer is 'no'. This is because the air-water exchange rate - in terms of the *amount* of $O_2$ exchanged per unit time $(mol\\,O_2\\,d^{-1})$ - is proportional to the *area* of the air-water interface through which the exchange occurs (as long as the water column is well-mixed)."
  
question("Does the maximum number of goldfish in the pond depend on the height of the water column, assuming that the surface area of the pond remains the same?", type = "learnr_radio",
 answer("Yes"),
 answer("No", correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## More exercises

If you finished this tutorial but some of your answers were incorrect, you are encouraged to do the quiz again. You can do this by clicking on <span style="color:red">**Start Over**</span> at the bottom of the menu to the left.

To learn more about modelling of partitioning between phases, solve additional exercises available in the R-package RTM. To display these exercises, type the following command in the R-console:

```
require(RTM)
RTMexercise("dissolutionSi")
RTMexercise("evaporation", output="PDF")
RTMexercise("O18exchange", output="PDF")
```

## Textbook and Slides

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>
