---
title: "RTM course - partitioning between phases"
author: "Karline Soetaert and Lubos Polerecky"
date: "June 2021"
description: "video and exercises on rate expressions for dissolution/precipitation, degassing reactions, including mineral and gas solubility"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first.

### Video timeline
* 0:00 Mineral dissolution and precipitation
* 4:38 Gas dissolution and degassing
* 6:12 Extra information about mineral solubility
* 8:07 Extra information about gas solubility

![](https://www.youtube.com/watch?v=CPme0Qz3L7Y&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=8)

There are no corresponding sections in the textbook.

## Mineral dissolution - quiz

```{r statements, echo=FALSE}
msg <- sprintf("The answer is $k_d/k_p$. Tip: if you do not remember whether this would be $k_p/k_d$ or $k_d/k_p$, think about the units. Concentrations are in $mol~m^{-3}$, and this unit is obtained with $k_d/k_p$, not with $k_p/k_d$.")

question(sprintf("The rate constant describing the dissolution of a mineral in water is $k_d = 2~mol~m^{-2}~yr^{-1}$, and the rate constant describing the precipitation of that mineral is $k_p = 5~m~yr^{-1}$. What is the equilibrium concentration of the dissolved mineral in water?"), type = "learnr_checkbox",
 answer(sprintf("0.4 $mol~m^{-3}$"), correct=TRUE),
 answer(sprintf("2.5 $mol~m^{-3}$")),
 answer(sprintf("1 $mol~m^{-3}$")),
 answer(sprintf("10 $mol~m^{-3}$")),
 answer(sprintf("400 $\\mu M$"), correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## Mineral dissolution - exercise

There are $N$ spherical amorphous silica particles suspended in a well-mixed volume of water $V$. Initially, the silicic acid concentration is $C_{ini}$, and all particles have the same diameter $D_{ini}$. Assume the rate constant for silica dissolution of $k_d = 1~mol~m^{-2}~yr^{-1}$ and for silica precipitation of $k_p = 1.5~m~yr^{-1}$. Silica density is $2200~kg~m^{-3}$, molar weight of silica is $60~g~mol^{-1}$.

The R-code chunk below introduces the values and units of the parameters you will need. 

```{r dSidt-setup, echo=TRUE}
N    <- 400
V    <- 1    # L
Cini <- 100  # umol/L (u stands for 'micro', so 1e-6)
Dini <- 90   # um
kd   <- 1    # mol/m2/yr
kp   <- 1.5  # m/yr
dens <- 2200 # kg/m3
MW   <- 60   # g/mol
```

```{r, echo=TRUE}
N    <- 400
V    <- 1    # L
Cini <- 100  # umol/L (u stands for 'micro', so 1e-6)
Dini <- 90   # um
kd   <- 1    # mol/m2/yr
kp   <- 1.5  # m/yr
dens <- 2200 # kg/m3
MW   <- 60   # g/mol
```

Use R to solve the following question (see tips below):

```{r dSidt_ini, echo=FALSE}
question("What is the *initial* rate of increase in the silicic acid concentration (dSi/dt) in the water with the silica particles?", type = "learnr_radio",
 answer(sprintf("%.1f $\\mu mol~L^{-1}~yr^{-1}$", -kd * N * 4*pi*(Dini/2*1e-6)^2 / (V*1e-3) * (Cini - kp/kd*1e3))),
 answer(sprintf("%.1f $nmol~L^{-1}~yr^{-1}$", 1e3* -kp * 1 * 4*pi*(Dini/2*1e-6)^2 / (V*1e-3) * (Cini - kd/kp*1e3))),
 answer(sprintf("%.2f $nmol~L^{-1}~yr^{-1}$", 1e3* -kp * N * 4*pi*(Dini/2*1e-6)^2 / (V) * (Cini - kd/kp*1e3))),
 answer(sprintf("%.2f $\\mu mol~L^{-1}~yr^{-1}$", -kp * N * 4*pi*(Dini/2*1e-6)^2 / (V*1e-3) * (Cini - kd/kp*1e3)), correct=TRUE),
 answer(sprintf("%.1f $\\mu mol~L^{-1}~yr^{-1}$", -kp * N * 4*pi*(Dini*1e-6)^2 / (V*1e-3) * (Cini - kd/kp*1e3))),
 
 incorrect = sprintf("If you think your formula is correct, double check that you have not made a mistake in unit conversion."),
 correct = sprintf("Bingo!")
)
```

Tips:

* The change in silicic acid concentration over time is described by
$$
\frac{dSi}{dt} = -k_p \times \frac {A}{V} \times (Si - Si_{eq})
$$
where $Si$ and $Si_{eq}$ is the instantaneous and equilibrium silicic acid concentration, respectively.
* Think about the meaning of $A$.
* Do your calculation in several steps rather than using one (possibly large) formula.
* Be careful with unit conversions.

Type your R-code here:

```{r dSidt, exercise=TRUE, exercise.cap = "Initial rate of increase of Si", exercise.lines = 4}
# fill in the formula
dSi.dt <- 
```

```{r dSidt-hint}
# This shows the sequence of calculations - replace NA with the correct formula 
Ceq    <- NA                     # equilibrium concentration (umol/L)
A      <- NA                     # initial total active surface (m2)
A.V    <- NA                     # area density (/m)
dSi.dt <- -kp*A.V*(Cini - Ceq)   # the change in time umol/L/yr
```

```{r dSidt-solution}
Ceq    <- kd/kp*1000                # equilibrium concentration (umol/L)
A      <- 4*pi*(Dini/2*1e-6)^2 * N  # initial total active surface (m2)
A.V    <- A/(V*1e-3)                # area density (/m)
dSi.dt <- -kp*A.V*(Cini - Ceq)      # the change in time umol/L/yr
```

```{r dSidt_t, echo=FALSE}
msg <- sprintf("The dissolution rate depends on the *difference* between the instantaneous and equilibrium concentration of silicic acid ($Si-Si_{eq}$). Because silica particles are dissolving, the silicic acid concentrations *increases* over time. Thus, the difference $Si-Si_{eq}$ is becoming smaller, implying that the third answer is correct.")

question("Which statement is true?", type = "learnr_radio",
 answer("The dissolution rate of silica will increase over time."),
 answer("The dissolution rate of silica will stay constant over time."),
 answer("The dissolution rate of silica will decrease over time.", correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## Air-water exchange - quiz

```{r statements2, echo=FALSE}
msg <- sprintf("This is calculated as $k_p/h\\cdot (O_{2,eq} - 0)$.")

question(sprintf("A well-mixed anoxic water column is in contact with the atmosphere. What is the initial rate of increase in the $O_2$ concentration in the water? Piston velocity is $1~m~d^{-1}$, $O_2$ solubility in the water is $300~\\mu mol~L^{-1}$, water column height is $50~cm$."), type = "learnr_radio",
 answer(sprintf("$300~\\mu mol~L^{-1}~d^{-1}$")),
 answer(sprintf("$150~nmol~L^{-1}~d^{-1}$")),
 answer(sprintf("$600~nmol~L^{-1}~d^{-1}$")),
 answer(sprintf("$600~\\mu mol~L^{-1}~d^{-1}$"), correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## Air-water exchange - exercise

You plan to install in your garden a pond with goldfish. By swimming in the pond, the fish will mix the water in the pond, and by respiration they will decrease the dissolved $O_2$ concentration in the pond. This decrease in dissolved $O_2$ will be compensated by air-water exchange. Assume that goldfish respiration is unaffected if the dissolved $O_2$ concentration does not decrease below 80% of the air saturated value. Additionally assume the piston velocity of $1~m~d^{-1}$, the area of the pond of $4~m^2$, the average height of the water column of $1~m$, Henry’s constant for dissolved $O_2$ of $770~L~Atm~mol^{-1}$, molar fraction of $O_2$ in the atmosphere of 21%, atmospheric pressure of $1~Atm$, and the average goldfish respiration rate of $400~\mu mol~fish^{-1}~d^{-1}$.

The R-code chunk below introduces the values and units of the parameters you will need. 

```{r Fish-setup}
pSat  <- 0.8  # -
kp    <- 1    # m/d
h     <- 1    # m
K0    <- 770  # L atm/mol
P     <- 1    # atm, total atmospheric pressure
pO2   <- 0.21 # molar fraction of O2
Resp  <- 400  # umol/fish/d
Area  <- 4    # m2
```

```{r parunits, echo=TRUE}
pSat  <- 0.8  # -
kp    <- 1    # m/d
h     <- 1    # m
K0    <- 770  # L atm/mol
P     <- 1    # atm, total atmospheric pressure
pO2   <- 0.21 # molar fraction of O2
Resp  <- 400  # umol/fish/d
Area  <- 4    # m2
```

Use R to solve the following question (see hints below):

```{r Nfish, echo=FALSE}
question("What is the maximum number of goldfish you can have in the pond so that their respiration will not be affected by the decreased $O_2$ concentrations in the pond? ", type = "learnr_radio",
 answer("25 fish"),
 answer(sprintf("%.0f fish", kp*Area * (1-pSat)* P*pO2/K0*1e3/(Resp*1e-6)), correct=TRUE),
 answer("252 fish"),
 answer(sprintf("%.0f fish", kp*Area * (1-pSat)* P/K0*1e3/(Resp*1e-6))),
 answer(sprintf("%.2f fish", kp*Area * (1-pSat)* P*pO2/K0*1e3/(Resp*1e-3))),

 incorrect = sprintf("Incorrect answer. If you think your formula is correct, double check that you have not made a mistake in unit conversion."),
 correct = sprintf("Bingo!")
)
```

Use this R chunk to figure out your answer (N is the number of fish):

```{r Fish, exercise=TRUE, exercise.cap = "Goldfish in a pond"}
# This shows a possible sequence of calculations
O2sat       <-                 # saturated oxygen concentration, mmol/m3
airwater    <-                 # mmol/m2/d - air-water flux
airwaterTot <-                 # mmol/d - total amount of O2 exchanged per day
Rmmol       <-                 # mmol/fish/d - fish respiration
N           <-                 # number of fish    
```

```{r Fish-hint}
# Hints:
#
# - At the minimal oxygen concentration, the aeration rate and respiration rate will be the same.
#  - Estimate the total air-water exchange (amount of oxygen added to the pond per day) at the minimal oxygen concentration (units mmol/day). 
#  - You know the respiration rate for one fish. The total respiration for all the fish should equal the total air-water exchange.
```

```{r Fish-solution}
O2sat       <- pO2/K0*P *1e6            # umol/L = mmol/m3
airwater    <- kp*(O2sat - pSat*O2sat)  # mmol/m2/d - air-water flux
airwaterTot <- airwater*Area            # mmol/d - total amount of O2 exchanged per day
Rmmol       <- Resp*1e-3                # mmol/fish/d - fish respiration
N           <- airwaterTot/Rmmol        # number of fish
N
```

And one extra question:

```{r Fish_point_height, echo=FALSE}
msg <- "It may be somewhat counter-intuitive, but the answer is 'no'. This is because the air-water exchange rate - in terms of the *amount* of $O_2$ exchanged per unit time ($mol~O_2~d^{-1}$) - is proportional to the *area* of the air-water interface through which the exchange occurs (as long as the water column is well-mixed). This is in contrast to the air-water exchange rate in terms of the *concentration* of $O_2$ per unit time ($mol~O_2~m^{-3}~d^{-1}$), which is proportional to the *area density* ($A/V=1/h$)."
  
question("Does the maximum number of goldfish in the pond depend on the height of the water column?", type = "learnr_radio",
 answer("Yes"),
 answer("No", correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## More exercises

To learn more about modelling of partitioning between phases, solve additional exercises available in the R-package RTM. To display these exercises, type the following command in the R-console:

```
require(RTM)
RTMexercise("partitioning")
```

## References

R Core Team (2020). R: A language and environment for statistical computing. R
  Foundation for Statistical Computing, Vienna, Austria. URL
  https://www.R-project.org/.

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>