---
title: "RTM course - partitioning between phases"
author: "Lubos Polerecky and Karline Soetaert"
date: "June 2021"
description: "video and exercises on rate expressions for dissolution/precipitation, degassing reactions, including mineral and gas solubility"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first.

### Video timeline
* 0:00 Mineral dissolution and precipitation
* 4:38 Gas dissolution and degassing
* 6:12 Extra information about mineral solubility
* 8:07 Extra information about gas solubility

![](https://www.youtube.com/watch?v=CPme0Qz3L7Y&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=8)

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>

There are no corresponding sections in the textbook.

## Mineral dissolution - quiz

There are $N$ spherical amorphous silica particles suspended in a well-mixed volume of water. The change over time in the concentration of dissolved silica is described by the following differential equation:

$$
\frac{dSi}{dt} = -k_p \times \frac {A}{V} \times (Si - Si_{eq})
$$
where $Si$ and $Si_{eq}$ are, respectively, the instantaneous and equilibrium concentrations of dissolved silica (both in $mol~Si~m^{-3}$).

```{r rate_constant, echo=FALSE}
msg <- sprintf("$k_p$ is the rate constant of the backward reaction, i.e., Si precipitation. It describes the amount of Si (moles) that precipitate per unit area of the mineral surface ($m^2$), per unit time ($yr$), and per unit concentration of dissolved silica ($mol~m^{-3}$). Thus, its unit is $mol/m^2/yr/(mol~m^{-3})$, which is equal to $m~yr^{-1}$.")

question(sprintf("What is the meaning and unit of the rate constant $k_p$?"), 
         type = "learnr_radio",
 answer(sprintf("$k_p$ is the dissolution rate constant, in $m~yr^{-1}$")),
 answer(sprintf("$k_p$ is the precipitation rate constant, in $mol~m^{-2}~yr^{-1}$")),
 answer(sprintf("$k_p$ is the precipitation rate constant, in $m~yr^{-1}$"), correct=TRUE),
 answer(sprintf("$k_p$ is the dissolution rate constant, in $mol~m^{-2}~yr^{-1}$")),
 
 incorrect=msg,
 correct=msg)
```

```{r volume, echo=FALSE}
msg <- sprintf("$V$ has nothing to do with the volume of the silica particles! $V$ is the volume of the well-mixed *water* in which the silica particles are suspended.")

question(sprintf("What is the meaning and unit of $V$ in the above equation?"), 
         type = "learnr_radio",
         answer(sprintf("$V$ is the volume of an *individual* silica particle, in $m^{3}$, calculated from the radius of the individual particle as $V=(4/3)\\pi r^3$.")),
         answer(sprintf("$V$ is the *total* volume of the silica particles, in $m^{3}$, calculated from the radius of the individual particle as $V=N\\times (4/3)\\pi r^3$.")),
         answer(sprintf("$V$ is the volume of well-mixed water containing the silica particles, in $m^3$."), correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

```{r area, echo=FALSE}
msg <- sprintf("$A$ is the *total* area of the silica particles. It corresponds to the overall surface area where the reaction (silica dissolution or precipitation) occurs.")

question(sprintf("What is the meaning and unit of $A$ in the above equation?"), 
         type = "learnr_radio",
         answer(sprintf("$A$ is the area of an *individual* silica particle, in $m^{2}$, calculated from the radius of the individual particle as $A=4\\pi r^2$.")),
         answer(sprintf("$A$ is the *total* area of all silica particles, in $m^{2}$, calculated from the radius of the individual particle as $A=N\\times 4\\pi r^2$."), correct=TRUE),

 incorrect=msg,
 correct=msg)
```

```{r eq_conc, echo=FALSE}
msg <- sprintf("The answer is $k_d/k_p$. Tip: if you do not remember whether the formula should be $k_p/k_d$ or $k_d/k_p$, think about the units. Concentrations are in $mol~m^{-3}$, and this unit is obtained with $k_d/k_p$, not with $k_p/k_d$. Note that there are *two* correct answers in the list! ($M$ stands for $mol~L^{-1}$.)")

question(sprintf("Assume that the rate constant describing the dissolution of a mineral in water is $k_d = 2~mol~m^{-2}~yr^{-1}$, and the rate constant describing the precipitation of that mineral is $k_p = 5~m~yr^{-1}$. What is the equilibrium concentration of the dissolved mineral in water?"), type = "learnr_checkbox",
 answer(sprintf("0.4 $mol~m^{-3}$"), correct=TRUE),
 answer(sprintf("2.5 $mol~m^{-3}$")),
 answer(sprintf("1 $mol~m^{-3}$")),
 answer(sprintf("10 $mol~m^{-3}$")),
 answer(sprintf("400 $\\mu M$"), correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```


## Mineral dissolution - exercise

There are $N=800$ spherical amorphous silica particles suspended in a well-mixed water volume of $V=1~L$. Initially, the concentration of the dissolved silica (silicic acid) in the water is $Si_{ini} = 100~\mu M$, and all particles have the same *diameter* $D_{ini}=0.1~mm$. Assume the rate constant for silica dissolution of $k_d = 1~mol~m^{-2}~yr^{-1}$ and for silica precipitation of $k_p = 0.8~m~yr^{-1}$. Bulk silica density is $2200~kg~m^{-3}$, molar weight of silica is $60~g~mol^{-1}$.

```{r dSidt_pars, echo=FALSE}
N    <- 800    # number of particles, -
V    <- 1e-3   # volume of water, m3
Dini <- 0.1e-3 # diameter of an individual particle, m
kd   <- 1      # dissolution rate constant, mol m-2 yr-1
kp   <- 0.8    # precipitation rate constant, m yr-1
Cini <- 100e-3 # initial concentration, mol m-3
dens <- 2200   # silica density, kg m-3
MW   <- 60e-3  # molar weight of silica, kg mol-1
dSi.ini.dt <- -kp * N * 4*pi*(Dini/2)^2 / V * (Cini - kd/kp)
```

```{r dSidt_ini, echo=FALSE}
question("What is the *initial* rate of increase in the silicic acid concentration (dSi/dt) in the water with the silica particles?", 
         type = "learnr_checkbox",
         answer(sprintf("23 $mmol~L^{-1}~yr^{-1}$")),
         answer(sprintf("23 $\\mu mol~L^{-1}~yr^{-1}$"), correct=TRUE),
         answer(sprintf("0.092 $mol~m^{-3}~yr^{-1}$")),
         answer(sprintf("55200 $mol~m^{-3}~yr^{-1}$")),
         answer(sprintf("27600 $mol~m^{-3}~yr^{-1}$")),
         answer(sprintf("69 $mol~L^{-1}~yr^{-1}$")),
         answer(sprintf("69 $mol~m^{-3}~yr^{-1}$")),
         answer(sprintf("92 $\\mu mol~L^{-1}~yr^{-1}$")),
         answer(sprintf("0.023 $mol~m^{-3}~yr^{-1}$"), correct=TRUE),
 incorrect = sprintf("If you think you are using the correct formula, double check that you did not make an error during units conversion. It is recommended that you convert *all* quantities into SI units *before* calculating the final value. Note that the information about the density and molar weight of silica is irrelevant in this exercise. Also note that the *diameter*, not the *radius*, of the silica particles is provided. There are *two* correct answers!"),
 correct = sprintf("Bingo!")
)
```

```{r dSidt_t, echo=FALSE}
msg <- sprintf("The dissolution rate depends on the *difference* between the instantaneous and equilibrium concentration of silicic acid ($Si-Si_{eq}$). Because silica particles are dissolving, the silicic acid concentrations *increases* over time. Thus, the difference $Si-Si_{eq}$ is becoming smaller, implying that the third answer is correct.")

question("Which statement is true?", type = "learnr_radio",
 answer("The dissolution rate of silica will increase over time."),
 answer("The dissolution rate of silica will stay constant over time."),
 answer("The dissolution rate of silica will decrease over time.", correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## Air-water exchange - quiz

```{r statements2, echo=FALSE}
msg <- sprintf("The correct value is calculated as $k_p/h\\cdot (O_{2,eq} - 0)$.")

question(sprintf("A well-mixed *anoxic* water column is in contact with the atmosphere. What is the initial rate of increase in the $O_2$ concentration in the water? Piston velocity is $k_p = 1~m~d^{-1}$, $O_2$ solubility in the water is $300~\\mu mol~L^{-1}$, water column height is $h=50~cm$."), type = "learnr_radio",
 answer(sprintf("$300~\\mu mol~L^{-1}~d^{-1}$")),
 answer(sprintf("$150~nmol~L^{-1}~d^{-1}$")),
 answer(sprintf("$600~nmol~L^{-1}~d^{-1}$")),
 answer(sprintf("$600~\\mu mol~L^{-1}~d^{-1}$"), correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## Air-water exchange - exercise

You plan to install in your garden a pond with goldfish. By swimming in the pond, the fish will mix the water in the pond, and by respiration they will decrease the dissolved $O_2$ concentration in the pond. This decrease in dissolved $O_2$ will be compensated by air-water exchange. Assume that goldfish respiration is unaffected if the dissolved $O_2$ concentration does not decrease below 80% of the air saturated value. Additionally assume the piston velocity of $k_p=1~m~d^{-1}$, the area of the pond of $4~m^2$, the average height of the water column of $1~m$, Henry’s constant for dissolved $O_2$ of $K_H=770~L~Atm~mol^{-1}$, molar fraction of $O_2$ in the atmosphere of 21%, the total atmospheric pressure of $1~Atm$, and the average goldfish respiration rate of $400~\mu mol~O_2~fish^{-1}~d^{-1}$.

```{r Fish-setup, echo=FALSE, eval=FALSE}
pSat  <- 0.8    # -
kp    <- 1      # m/d
h     <- 1      # m
KH    <- 770e-3 # m3 atm/mol
P     <- 1      # atm, total atmospheric pressure
pO2   <- 0.21   # molar fraction of O2
Resp  <- 400e-6 # mol/fish/d
Area  <- 4      # m2
# solution (one formula)
Nmax <- kp*Area * (1-pSat)* P*pO2/KH/Resp
# solution in parts
O2sat        <- pO2/KH*P
airwater     <- kp*(O2sat - pSat*O2sat)
airwaterTot  <- airwater*Area
N           <- airwaterTot/Resp
```

```{r Nfish, echo=FALSE}
question("What is the maximum number of goldfish you can have in the pond so that their respiration will not be affected by the decreased $O_2$ concentrations in the pond? ", 
         type = "learnr_radio",
         answer("25 fish"), 
         answer("545 fish", correct=TRUE), 
         answer("252 fish"),
         answer("1090 fish"),
         answer("100 fish"),

 incorrect = sprintf("Incorrect answer. If you think your formula is correct, double check that you have not made a mistake during units conversion. If you cannot come up with the formula, consider the following hints: (1) At the minimal $O_2$ concentration where the fish respiration is still not affected, the aeration and respiration rates should be equal to each other. (2) Use your knowledge of physical chemistry (Henry's law) to quantify the $O_2$ solubility at the given partial pressure of $O_2$ in the atmosphere. (3) Estimate the total air-water exchange (amount of $O_2$ added to the pond per day, $mol~O_2~d^{-1}$) at the minimal $O_2$ concentration. (4) You know the respiration rate for one fish, so you can easily calculate the total respiration rate by all fish in the pond."),
 correct = sprintf("Bingo!")
)
```

And one extra question:

```{r Fish_point_height, echo=FALSE}
msg <- "It may be somewhat counter-intuitive, but the answer is 'no'. This is because the air-water exchange rate - in terms of the *amount* of $O_2$ exchanged per unit time ($mol~O_2~d^{-1}$) - is proportional to the *area* of the air-water interface through which the exchange occurs (as long as the water column is well-mixed)."
  
question("Does the maximum number of goldfish in the pond depend on the height of the water column?", type = "learnr_radio",
 answer("Yes"),
 answer("No", correct=TRUE),
 
 incorrect=msg,
 correct=msg)
```

## More exercises

To learn more about modelling of partitioning between phases, solve additional exercises available in the R-package RTM. To display these exercises, type the following command in the R-console:

```
require(RTM)
RTMexercise("dissolutionSi")
```

## Textbook and Slides

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>
