---
title: |
  | Local Equilibrium Chemistry in R --- part III:
  | Four Equilibrium Reactions
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Karline Soetaert and Lubos Polerecky, Utrecht University"
date: "February 2021"
output:
  pdf_document:
    number_sections: false
  html_document: default
linestretch: 1.15
---

# Answer

## Task 1: Mass balance equations of the carbonate system

Assuming that the forward and backward reactions are elementary, we can use the first-order kinetics to describe the rate laws for each reaction. Considering the stoichiometries of reactions 1--4, we obtain the following mass balance equations:

$$
\begin{array}{rclr}
\displaystyle
\frac{d[H_2CO_3]}{dt} & = & -R_1 + \color{red}{R_C} & \qquad (1a)\\[5mm]
\displaystyle
\frac{d[HCO_3^-]}{dt} & = & R_1 - R_2 & \qquad (1b) \\[5mm]
\displaystyle
\frac{d[CO_3^{2-}]}{dt} & = & R_2  & \qquad (1c)\\[5mm]
\displaystyle
\frac{d[NH_4^+]}{dt} & = & -R_3 & \qquad (1d) \\[5mm]
\displaystyle
\frac{d[NH_3]}{dt} & = & R_3 + \color{red}{R_N} & \qquad (1e) \\[5mm]
\displaystyle
\frac{d[H_2O]}{dt} & = & -R_4 & \qquad (1f) \\[5mm]
\displaystyle
\frac{d[OH^-]}{dt} & = & R_4 & \qquad (1g) \\[5mm]
\displaystyle
\frac{d[H^+]}{dt} & = & R_1 + R_2 + R_3 + R_4 & \qquad (1h) \\[5mm]
\end{array}
$$

where the individual rates ($R_1$, $R_2$, $R_3$ and $R_4$) are defined as the difference between the forward and backward reaction rates:

$$
\begin{array}{rclr}
R_1 &=& k_{1f} \cdot [H_2CO_3] - k_{1b} \cdot [HCO_3^-] \cdot [H^+] &
\qquad\hspace{2cm} (2a) \\[4mm]
R_2 &=& k_{2f} \cdot [HCO_3^{-}] - k_{2b} \cdot [CO_3^{2-}] \cdot [H^+] &
\qquad\hspace{2cm} (2b) \\[4mm]
R_3 &=& k_{3f} \cdot [NH_4^{+}] - k_{3b} \cdot [NH_3] \cdot [H^+]  &
\qquad\hspace{2cm} (2c) \\[4mm]
R_4 &=& k_{4f} \cdot [H_2O] - k_{4b} \cdot [OH^-] \cdot [H^+] &
\qquad\hspace{2cm} (2d)
\end{array}
$$

Note that the slow processes, $R_C$ and $R_N$, only affect $H_2CO_3$ and $NH_3$.

## Task 2: Equilibrium conditions

In an equilibrium, the time derivative of each species in the system is zero. Based on equations 1a and 2a, we obtain $R_1=0$, which yields the relationship

$$
K_{1} = \frac{ [HCO_3^-] \cdot [H^+] }{ [H_2CO_3] }. \qquad\hspace{2cm} (3a)
$$

Similarly, based on Eq. 1c and 2b, we obtain $R_2=0$, which yields the relationship

$$
K_{2}  = \frac{ [CO_3^{2-}] \cdot [H^+] }{[HCO_3^-]}. 
\qquad\hspace{2cm} (3b)
$$
Based on Eq. 1d and 2c, we obtain $R_3=0$, which yields the relationship

$$
K_{n}  = \frac{ [NH_3] \cdot [H^+] }{[NH_4^+]}. 
\qquad\hspace{2cm} (3c)
$$

Finally, based on Eq. 1f and 2d, we obtain $R_4=0$, which yields the relationship

$$
K_{w}'  = \frac{ [OH^-] \cdot [H^+] }{[H_2O]}. 
\qquad\hspace{2cm} (3d)
$$

Here, we defined the dissociation constants $K_1$, $K_2$, $K_n$, $K_w'$ according to $K_1 = k_{1f}/k_{1b}$, $K_2 = k_{2f}/k_{2b}$, $K_n = k_{3f}/k_{3b}$, and $K_w' = k_{4f}/k_{4b}$. If these relationships are valid, then the time derivatives in equations 1b, 1e, 1g and 1h are zero. This means that these four equations do *not* provide any *extra* information for the equilibrium concentrations.

## Task 3: Lump-sum species

Here we combine equations 1a--1h and see what they imply for the suggested lump-sum species. By summing equations 1a, 1b and 1c, we obtain 

$$
\frac{d[DIC]}{dt} = {\color{red}{R_C}} \qquad\hspace{2cm} (4a)
$$

Similarly, by summing equations 1d and 1e, we obtain

$$
\frac{d[NH_x]}{dt} = {\color{red}{R_N}} \qquad\hspace{2cm} (4b)
$$

Similarly, by summing equations 1f and 1g, we obtain

$$
\frac{d[H_xO]}{dt} = 0 \qquad\hspace{2cm} (4c)
$$

Finally, by summing equations according to (1b) + $2\times$(1c) + (1g) $-$ (1d) $-$ (1h), we obtain
$$
\frac{d[ALK]}{dt} = 0 \qquad\hspace{2cm} (4d)
$$
These equations show that DIC is *added* at a rate of the *slow process*, $R_C$, $NH_x$ is *added* at a rate of the *slow process*, $R_N$, while $H_xO$ and $ALK$ are *conserved* (i.e., not changing over time). Specifically, *none* of the lump-sum species $DIC$, $NH_x$, $H_xO$ and $ALK$ is affected by the fast reversible reactions.

## Task 4: From lump-sum species to the original species

This task has been solved in parts I and II of this reader series. Here we only summarize the results:

$$
\begin{array}{rclr}
[HCO_3^-] &=& \displaystyle \frac{K_{1}[H^+]}{[H^+][H^+] + K_{1}[H^+] +  K_{1} K_{2}}\cdot [DIC] &
\qquad\hspace{2cm} (5a) \\[5mm]
[H_2CO_3] &=& \displaystyle \frac{[H^+][H^+]}{[H^+][H^+] + K_{1}[H^+] +  K_{1} K_{2}}\cdot [DIC] &
\qquad\hspace{2cm} (5b) \\[5mm]
[CO_3^{2-}] &=& \displaystyle \frac{K_1 K_2}{[H^+][H^+] + K_{1}[H^+] +  K_{1} K_{2}}\cdot [DIC] &
\qquad\hspace{2cm} (5c) \\[5mm]
[NH_3] &=& \displaystyle \frac{K_n}{K_n + [H^+]}\cdot [NH_x] & \qquad\hspace{2cm} (5d) \\[5mm]
[NH_4^+] &=& \displaystyle \frac{[H^+]}{K_n + [H^+]} \cdot [NH_x] & \qquad\hspace{2cm} (5e) \\[5mm]
[OH^-] &=& \displaystyle \frac{K_w'}{K_w' + [H^+]}\cdot [H_xO] & \qquad\hspace{2cm} (5f) \\[5mm]
[H_2O] &=&  \displaystyle \frac{[H^+]}{K_w' + [H^+]}\cdot [H_xO] & \qquad\hspace{2cm} (5g)
\end{array}
$$

Once the concentrations $[HCO_3^-]$, $[CO_3^{2-}]$, $[NH_4^+]$ and $[OH^-]$ are calculated from equations 5a, 5c, 5e and 5f, respectively, the alkalinity is calculated from the definition:
$$
[ALK] = [HCO_3^-] + 2\cdot [CO_3^{2-}] + [OH^-] - [NH_4^+] - [H^+]
\qquad\hspace{2cm} (5h)
$$
Because $[HCO_3^-]$ and $[CO_3^{2-}]$ depend on $[DIC]$ and $[H^+]$ (Eqs. 5a and 5c), $[NH_4^+]$ depends on $[NH_x]$ and $[H^+]$ (Eq. 5d), and $[OH^-]$ depends on $[H_xO]$ and $[H^+]$ (Eq. 5f), **equation 5h provides an intimate relationship between $[ALK]$, $[DIC]$, $[NH_x]$, $[H_xO]$ and $[H^+]$** when the carbonate, ammonium and water species are in equilibrium. Using this relationship, one can calculate any of the five species provided that the other \emph{four} species are known. Because equation 5h is highly non-linear, these calculations are done numerically.

## Task 5: Application to organic matter mineralization

### R implementation

First, we expand the function that calculates alkalinity by including the concentration of $NH_4^+$ (Eq. $5e$).

```{r}
solveALK <- function(K1=4.44e-07, K2=4.67e-11, Kn=5.68e-10, Kw.prime=1.79e-16,
                     HxO=55.3, DIC=0, NHx=0, pH=7){
  
  H      <- 10^(-pH)
  HCO3   <- K1*H /  (H^2 + K1*H + K1*K2) * DIC    # eq 5a
  CO3    <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC    # eq 5c
  NH4    <- H / (H + Kn) * NHx                    # eq 5e
  OH     <- Kw.prime / (Kw.prime + H) * HxO       # eq 5f
  ALK    <- HCO3 + 2*CO3 + OH - NH4 - H           # eq 5h
  
  return(ALK)
}
```

Next, we update the function that calculates $[H^+]$ from given concentrations of $ALK$, $DIC$, $NH_x$ and $H_xO$ (Eq. $5h$). This is done by finding a root of a function defined as the left-hand-side minus the right-hand-side of Eq. $5h$.

```{r}
solveH <- function(K1=4.44e-07, K2=4.67e-11, Kn=5.68e-10, Kw.prime=1.79e-16,
                   HxO=55.3, DIC=0, NHx=0, ALK=0){
  
  # function whose root has to be sought
  rootFun <- function(H) {
    ALK.est <- solveALK(K1=K1, K2=K2, Kn=Kn, Kw.prime=Kw.prime,
                        HxO=HxO, DIC=DIC, NHx=NHx, pH=-log10(H))
    return(ALK.est - ALK)
  }
  
  # uniroot will find the root    
  h <- uniroot(f = rootFun, lower = 0, upper = 1, tol = 1e-30)
  return(h$root)
}
```

It is always good to test the functions. We see that the results are consistent with expectations.

```{r}
pH.new     <- -log10(solveH(DIC=2e-3, NHx=1e-5, ALK=2e-3))
ALK        <-  solveALK(DIC=2e-3, NHx=1e-5, pH = pH.new)
pH.neutral <- -log10(solveH(DIC=0, NHx=0, ALK=0))
c(pH.new=pH.new, ALK=ALK, pH.neutral=pH.neutral)
```

Next, we implement the model function based on the local equilibrium assumption. Note that 

1. The equilibrium constants (dissociation constants) are in $mol~L^{-1}$, so all species must be in this unit, too! 

2. $OM$ is added to the model as a state variable removed by the slow process.

3. Because we will additionally want to use the function for modelling ammonia degassing (Task 6), the effect of this process is added, too. Its rate constant ($\lambda$) will be set to 0 by default.

```{r}
OMMpHmodel <- function(t, state, parms) {
  with (as.list(c(state, parms)), {

    # variables needed as output and for regulation of process rates
    H       <- solveH(K1=K1, K2=K2, Kn=Kn, Kw.prime=Kw.prime, 
                      HxO=HxO, DIC=DIC, NHx=NHx, ALK=ALK)
    HCO3    <- K1*H /  (H^2 + K1*H + K1*K2) * DIC   # eq 5a
    CO3     <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC   # eq 5c
    NH4     <- H / (H + Kn) * NHx                   # eq 5e
    OH      <- Kw.prime / (Kw.prime + H) * HxO      # eq 5f
    NH3     <- NHx - NH4          # from definition of NHx
    H2CO3   <- DIC - HCO3 - CO3   # from definition of DIC
    H2O     <- HxO - OH           # from definition of HxO
    
    # process rates
    RC      <- rC*OM         # OM degradation
    Rdegas  <- lambda * NH3  # ammonia degassing (specifically removes NH3)
    
    # mass balance equations (assuming local equilibrium)
    dOM     <- -RC
    dDIC    <-  RC              # Eq. 4a
    dNHx    <-  RC*NC - Rdegas  # Eq. 4b - ammonia degassing
    dHxO    <-  0               # Eq. 4c
    dALK    <-  0               # Eq. 4d
    
    return(list(c(dOM, dDIC, dNHx, dHxO, dALK), 
                pH = -log10(H), 
                H = H, OH = OH, H2O = H2O,
                H2CO3 = H2CO3, HCO3 = HCO3, CO3  = CO3,
                NH4 = NH4, NH3 = NH3))
  })
}
```

We run the model for 10 days, using the initial conditions given in the task.

```{r, message=FALSE}
require(deSolve)

pars <- c(rC = 0.1,            # [1/day] OM degradation
          lambda = 0,          # [1/day] ammonia degasing
          NC = 16/106,         # molN/molC
          K1 = 4.44e-07,       # mol/L
          K2 = 4.67e-11,       # mol/L
          Kn = 5.68e-10,       # mol/L
          Kw.prime = 1.79e-16) # mol/L

pH.ini  <- 7.9                 # H.ini based on pH
OM.ini  <- 100*1e-6            # mol/L
DIC.ini <- 2000*1e-6           # mol/L
NHx.ini <- 10*1e-6             # mol/L
HxO.ini <- 55.3                # mol/L
ALK.ini <- solveALK(HxO = HxO.ini, DIC = DIC.ini, NHx = NHx.ini, pH = pH.ini)

state.ini <- c(OM=OM.ini, DIC=DIC.ini, NHx=NHx.ini, HxO=HxO.ini, ALK=ALK.ini)
times     <- seq(from=0, to=10, length.out=100)
out       <- ode(y=state.ini, times=times, func=OMMpHmodel, parms=pars)
```

```{r, fig.height=10, fig.width=10}
plot(out, mfrow = c(4,4), las = 1)
```

We see that mineralization of about $60~\mu mol~L^{-1}$ of organic matter over the period of 10 days leads to a decrease in $pH$ from 7.9 to about 7.63. As expected, $ALK$ and $H_xO$ are conserved and changes in the concentration of $H_2O$ are negligible due to vastly larger concentrations of water compared to all the other species in the system.

## Task 6: Does the formulation of slow reversible processes matter?

If the stoichiometry of OM mineralization is represented by the reaction
$$
\quad CH_2O(NH_3)_{NC} + O_2 \rightarrow HCO_3^- + H^- + {NC} \cdot NH_3,
$$
the mass balance equations 1c--g do not change while equations 1a, 1b, and 1h change in the following way:

$$
\begin{array}{rclr}
\displaystyle
\frac{d[H_2CO_3]}{dt} & = & R_1  & \qquad (1a')\\[5mm]
\displaystyle
\frac{d[HCO_3^-]}{dt} & = & -R_1 + R_2 + \color{red}{R_C}& \qquad (1b') \\[5mm]
\displaystyle
\frac{d[H^+]}{dt} & = & -R_1 - R_2 - R_3 - R_4 + \color{red}{R_C} & \qquad (1h') \\[5mm]
\end{array}
$$
This change, however, has *no* effect on the differential equations for the lump-sum variables $DIC$, $NH_x$, $H_xO$ and $ALK$, as can easily be verified by following the same derivation steps as those leading to equations 4a--d. This can also be shown more easily: although $R_C$ adds $HCO_3^-$ instead of $H_2CO_3$, these two species are both included in the $DIC$ pool, thus summing them up yields the same result for the time-derivative of $DIC$. Similarly, although the process *increases* $ALK$ at a rate $R_C$ due to the addition of $HCO_3^-$, it also *decreases* $ALK$ at the same rate $R_C$ due to the addition of $H^+$, as follows from the *negative* sign in front of $[H^+]$ in the definition of $ALK$.

This result has important implications: the specific "chemical" formulation of the slow reversible process has *no effect* on the dynamics of the lump-sum variables, as long as the chemical reactions are correctly balanced in terms of the elements and charges and the local equilibrium conditions can be assumed. 

## Task 7: Application to ammonia degassing

### R implementation

First, we reproduce here the ammonia degassing model developed in Part I:

```{r}
solveB <- function(Keq, Atot, Btot){
  rootFun <- function(B) B + B/(Keq+B)*Atot - Btot
  return(uniroot(f = rootFun, lower = 0, upper = Btot, tol = 1e-20)$root)
}

AmmoniaDegassing <- function(t, state, parms) {
  with (as.list(c(state, parms)), {
    H   <- solveB(Keq = Kn, Atot = NHx, Btot = Hx) 
    NH3 <- Kn/(Kn+H) * NHx
    NH4 <- NHx - NH3
    # mass balance equations for the lump-sum variables:
    dNHx <- -lambda * NH3
    dHx  <- 0
    return(list(c(dNHx, dHx), 
                H = H, NH3 = NH3, NH4 = NH4, # NH4 corresponds to [NH4+]
                pH = -log10(H),     # return also pH
                Q = NH3*H/NH4 ))    # return also Q
  })
}
```

We define the initial conditions and run the model. In this step, we reuse the parameters defined for the ``OMMpHmodel`` model with the value of the degassing rate constant, $\lambda$, changed to $1~d^{-1}$.

```{r, message=FALSE}
# update model parameters
pars1           <- pars
pars1["lambda"] <- 1

# initial conditions
NHx.ini <- 1e-3         # mol/L
pH.ini  <- 8
H.ini   <- 10^(-pH.ini) # mol/L
Hx.ini  <- H.ini + H.ini/(pars1[["Kn"]] + H.ini) * NHx.ini
state.ini1 <- c(NHx = NHx.ini, Hx = Hx.ini)

# run the model
out1 <- ode(y=state.ini1, times=times, func=AmmoniaDegassing, parms=pars1)
```

Next, we run the ``OMMpHmodel`` model with the same initial conditions. By setting the OM degradation rate constant to 0, and further setting the concentrations of $DIC$ and $H_xO$ to 0, we expect that the model becomes equivalent to the ``AmmoniaDegassing`` model.

```{r}
# update model parameters
pars2       <- pars1
pars2["rC"] <- 0

# initial conditions
ALK.ini    <- solveALK(HxO=0, DIC=0, NHx=NHx.ini, pH=pH.ini)
state.ini2 <- c(OM=OM.ini, DIC=0, NHx=NHx.ini, HxO=0, ALK=ALK.ini)

# run the model
out2 <- ode(y=state.ini2, times=times, func=OMMpHmodel, parms=pars2)
```

Thus, we expect to obtain the same results, at least for the state variables that are common for both models. This is indeed the case when we plot the results.

```{r, fig.width=10, fig.height=2.5}
plot(out2, obs=out1, which=c("NHx", "NH4", "NH3", "pH"), 
     mfrow = c(1,4), lwd=2, lty=1, col=1,
     obspar = list(lty=2,lwd=2,col=2,type="l"))
```

Now that we have verified that the model ``OMMpHmodel`` can be used for solving the ammonia degassing problem, too, we can illustrate the $pH$ buffering role of $DIC$ in natural waters. We will use the same initial conditions and the ammonia degassing rate constant as above and perform the calculations for $DIC$ concentrations of 0, 1, and $2~mmol~L^{-1}$.

```{r}
# without DIC (but with ammonia and water!)
ALK.ini3   <- solveALK(DIC=0, HxO=HxO.ini, NHx=NHx.ini, pH=pH.ini)
state.ini3 <- c(OM=OM.ini, DIC=0, NHx=NHx.ini, HxO=HxO.ini, ALK=ALK.ini3)
out3       <- ode(y=state.ini3, times=times, func=OMMpHmodel, parms=pars2)

# model with DIC=1 mmol/L
ALK.ini4   <- solveALK(DIC=1e-3, HxO=HxO.ini, NHx=NHx.ini, pH=pH.ini)
state.ini4 <- c(OM=OM.ini, DIC=1e-3, NHx=NHx.ini, HxO=HxO.ini, ALK=ALK.ini4)
out4       <- ode(y=state.ini4, times=times, func=OMMpHmodel, parms=pars2)

# model with DIC=2 mmol/L
ALK.ini5   <- solveALK(DIC=2e-3, HxO=HxO.ini, NHx=NHx.ini, pH=pH.ini)
state.ini5 <- c(OM=OM.ini, DIC=2e-3, NHx=NHx.ini, HxO=HxO.ini, ALK=ALK.ini5)
out5       <- ode(y=state.ini5, times=times, func=OMMpHmodel, parms=pars2)
```

```{r, fig.height=5.5, fig.width=10, echo=FALSE}
# plot results
plot(out3, out4, out5, 
     which=c("pH", "H", "OH", "HxO", "DIC", "ALK"),
     mfrow = c(2,3), las = 1, col=1:3, lty=1:3, lwd=2)
legend("topright", legend = c("DIC=0", "DIC=1 mmol/L", "DIC=2 mmol/L"), bty="n",
       col = 1:3, lwd = 2, lty = 1:3)
```

```{r, fig.height=5.5, fig.width=10, echo=FALSE}
# plot results
plot(out3, out4, out5, 
     which=c("NHx", "NH4", "NH3", "H2CO3", "HCO3", "CO3"),
     mfrow = c(2,3), las = 1, col=1:3, lty=1:3, lwd=2)
legend("topright", legend = c("DIC=0", "DIC=1 mmol/L", "DIC=2 mmol/L"), bty="n",
       col = 1:3, lwd = 2, lty = 1:3)
```

The results show that when $DIC$ is present, the protons released due to ammonia degassing are scavenged by both $CO_3^{2-}$ and $HCO_3^-$ in the solution, leading to an increase in $H_2CO_3$. Consequently, the increase in the concentration of protons, and thus the decrease in $pH$, is smaller in the presence of $DIC$, illustrating the $pH$ buffering effect by the carbonate system. The magnitude of the effect does not, however, scale linearly with the $DIC$ concentration. 
