---
title: |
  | Local Equilibrium Chemistry in R --- part III:
  | Mixed Equilibrium-Irreversible Reactions
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Karline Soetaert and Lubos Polerecky, Utrecht University"
date: "February 2021"
output:
  pdf_document:
    number_sections: false
  html_document: default
---

# Answers

## Task 1: Mass balance equations of the carbonate system

Assuming that the forward and backward reactions are elementary, we can use the first-order kinetics to describe the rate laws for each reaction. Considering the stoichiometries of reactions 1, 2 and 3, we obtain the following mass balance equations:

$$
\begin{array}{rclr}
\displaystyle
\frac{d[H_2CO_3]}{dt} & = & R_1 + R_C & \qquad (1a)\\[5mm]
\displaystyle
\frac{d[HCO_3^-]}{dt} & = & -R_1 + R_2 & \qquad (1b) \\[5mm]
\displaystyle
\frac{d[CO_3^{2-}]}{dt} & = & -R_2  & \qquad (1c)\\[5mm]
\displaystyle
\frac{d[NH_4^+]}{dt} & = & R_3 & \qquad (1d) \\[5mm]
\displaystyle
\frac{d[NH_3]}{dt} & = & -R_3 + R_N & \qquad (1e) \\[5mm]
\displaystyle
\frac{d[H^+]}{dt} & = & -R_1 - R_2 - R_3 & \qquad (1f)
\end{array}
$$

where the individual rates are defined as the sum of the rates of the forward and backward reactions:

$$
R_1 = - k_{1f} \cdot [H_2CO_3] + k_{1b} \cdot [HCO_3^-] \cdot [H^+] 
\qquad\hspace{2cm} (2a)
$$

$$
R_2 = - k_{2f} \cdot [HCO_3^{-}] + k_{2b} \cdot [CO_3^{2-}] \cdot [H^+] 
\qquad\hspace{2cm} (2b)
$$

$$
R_3 = - k_{3f} \cdot [NH_4^{+}] + k_{3b} \cdot [NH_3] \cdot [H^+] 
\qquad\hspace{2cm} (2c)
$$

## Task 2: Equilibrium conditions

In an equilibrium, the time derivative of each species in the system is zero. Based on equations 1a and 2a, we obtain $R_1=0$, which yields the relationship

$$
K_{1} = \frac{ [HCO_3^-] \cdot [H^+] }{ [H_2CO_3] }. \qquad\hspace{2cm} (3a)
$$

Similarly, based on equation 1c and 2b, we obtain $R_2=0$, which yields the relationship

$$
K_{2}  = \frac{ [CO_3^{2-}] \cdot [H^+] }{[HCO_3^-]}. 
\qquad\hspace{2cm} (3b)
$$

Finally, based on equation 1d and 2c, we obtain $R_3=0$, which yields the relationship

$$
K_{n}  = \frac{ [NH_3] \cdot [H^+] }{[NH_4^+]}. 
\qquad\hspace{2cm} (3c)
$$

Here, we defined the dissociation constants $K_1$, $K_2$ and $K_n$ as $K_1 = k_{1f}/k_{1b}$, $K_2 = k_{2f}/k_{2b}$, and $K_n = k_{3f}/k_{3b}$. If these relationships are true, then the time derivatives in equations 1b, 1e and 1f are zero. This means that these three equations do *not* provide any *extra* information for the equilibrium concentrations.

## Task 3: Lump-sum species

By summing equations 1a, 1b and 1c, we obtain 

$$
\frac{d[DIC]}{dt} = R_C \qquad\hspace{2cm} (4a)
$$

Similarly, by summing equation 1d and 1e, we obtain

$$
\frac{d[NH_x]}{dt} = R_N \qquad\hspace{2cm} (4b)
$$

Finally, by summing previous equations as (1b) + $2\times$(1c) $-$ (1d) $-$ (1f), we obtain
$$
\frac{d[ALK]}{dt} = 0 \qquad\hspace{2cm} (4c)
$$
These equations show that DIC is *added* at a rate of the *slow process* with rate $R_C$, $NH_x$ is *added* at a rate of the *slow process* $R_N$, while ALK is *conserved* (not changing in time). Specifically, *none* of the lump-sum species DIC, $NH_x$ and ALK is affected by the fast equilibration reactions.

## Task 4: From lump-sum species to the original species

This task has been solved in Parts I and II of the Reader. Here we only summarize the results.

$$
[HCO_3^-]= \frac{K_{1}[H^+]}{[H^+][H^+] + K_{1}[H^+] +  K_{1} K_{2}} [DIC]
\qquad\hspace{2cm} (5a)
$$

$$
[H_2CO_3] = \frac{[H^+][H^+]}{[H^+][H^+] + K_{1}[H^+] +  K_{1} K_{2}} [DIC]
\qquad\hspace{2cm} (5b)
$$

$$
[CO_3^{2-}] = \frac{K_1 K_2}{[H^+][H^+] + K_{1}[H^+] +  K_{1} K_{2}}[DIC]
\qquad\hspace{2cm} (5c)
$$

$$
[NH_3] = \frac{K_n}{K_n + [H^+]} [NH_x] \qquad\hspace{2cm} (5d)
$$

$$
[NH_4^+] = \frac{[H^+]}{K_n + [H^+]} [NH_x] \qquad\hspace{2cm} (5e)
$$

Once the concentrations $[HCO_3^-]$, $[CO_3^{2-}]$, $[NH_4^+]$, and $[H^+]$ are known, the alkalinity can be calculated from the definition,
$$
[ALK] = [HCO_3^-] + 2\cdot [CO_3^{2-}] - [NH_4^+] - [H^+]
\qquad\hspace{2cm} (5f)
$$
Because $[HCO_3^-]$ and $[CO_3^{2-}]$ depend on $[DIC]$ and $[H^+]$ (see 5a and 5c), and $[NH_4^+]$ depends on $[NH_x]$ and $[H^+]$, \bf equation 5f provides an intimate relationship between $[ALK]$, $[DIC]$, $[NH_x]$, and $[H^+]$ \rm when the system is in an equilibrium. Using this relationship, one can calculate any of the species provided that the other three species are known. Because equation 5f is highly non-linear, these calculations are done numerically.

## Task 5: Application to organic matter mineralization

### R implementation

First, we expand the function that calculates alkalinity to include ammonium as the additionally input (equation 5f).

```{r}
solveALK <- function(K1=1.422e-06, K2=1.082e-09, Kn=5.381e-10,
                     DIC, NHx, pH){
  
  H      <- 10^(-pH)
  HCO3   <- K1*H /  (H^2 + K1*H + K1*K2) * DIC    # eq 5a
  CO3    <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC    # eq 5c
  NH4    <- H / (H + Kn) * NHx                    # eq 5e
  ALK    <- HCO3 + 2*CO3 - NH4 - H                # eq 5f
  
  return(ALK)
}
```

Then, we update the function that calculates $[H^+]$ from the known alkalinity, DIC and NHx concentration (equation 5f). This is done by finding a root of a function defined as the left-hand-side minus the right-hand-side of equation 5f.

```{r}
solveH <- function(K1 = 1.422e-06, K2 = 1.082e-09, Kn = 5.381e-10,
                   DIC, NHx, ALK){
  
  # function whose root has to be sought
  rootFun <- function(H) {
    
    HCO3    <- K1*H /  (H^2 + K1*H + K1*K2) * DIC   # eq 5a
    CO3     <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC   # eq 5c
    NH4     <- H / (H + Kn) * NHx                   # eq 5e
    ALK.est <- HCO3 + 2*CO3 - NH4 - H # right-hand-side of eq. 5d
    
    return(ALK.est - ALK)
  }
  
  # uniroot will find the root    
  h <- uniroot(f = rootFun, lower = 0, upper = 1e-3, tol = 1e-30)
  return(h$root)
}
```

It is always good to test the functions.

```{r}
(pH  <- -log10(solveH(DIC = 2e-3, NHx = 1e-5, ALK = 2.14e-3)))
(ALK <- solveALK(DIC = 2e-3, NHx = 1e-5, pH = 7.923034))
```

The model function that implements the model based on the local equilibrium assumption is defined below. Note that the equilibrium constants (dissociation constants) are in $mol~L^{-1}$, so all species should be in this unit!

```{r}
OMMpHmodel <- function(t, state, parms) {
  with (as.list(c(state, parms)), {

    # OMM rate
    RC   <- rC*OM
    
    # mass balance equations
    dOM  <- -RC
    dDIC <-  RC
    dNHx <-  RC*NC
    dALK <- 0

    # for output
    H <- solveH(K1 = K1, K2 = K2, Kn = Kn, DIC = DIC, NHx = NHx, ALK = ALK)
    HCO3    <- K1*H /  (H^2 + K1*H + K1*K2) * DIC   # eq 5a
    CO3     <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC   # eq 5c
    NH4     <- H / (H + Kn) * NHx                   # eq 5e
    
    return(list(c(dOM, dDIC, dNHx, dALK), 
                pH = -log10(H), H = H, 
                H2CO3 = DIC - HCO3 - CO3,
                HCO3 = HCO3,
                CO3  = CO3,
                NH4  = NH4,
                NH3  = NHx - NH4))
  })
}
```

Now we run the model for 10 days, using the initial conditions given in the task.

```{r, fig.height=8, fig.width=10}
require(deSolve)

pars <- c(rC = 0.1,          # [1/day]
          NC = 16/106,       # molN/molC
          K1 = 1.422e-06, # mol/L
          K2 = 1.082e-09, # mol/L
          Kn = 5.381e-10) # mol/L

pH.ini  <- 7.9                   # Hini based on pH
DIC.ini <- 2000*1e-6             # mol/L
NHx.ini <- 10*1e-6               # mol/L
ALK.ini <- solveALK(DIC = DIC.ini, NHx = NHx.ini, pH = pH.ini)
OM.ini  <- 100*1e-6              # mol/L

state.ini <- c(OM=OM.ini, DIC=DIC.ini, NHx=NHx.ini, ALK=ALK.ini)
times     <- seq(from=0, to=10, length.out=100)

out   <- ode(y=state.ini, times=times, func=OMMpHmodel, parms=pars)

plot(out, mfrow = c(3,4), las = 1)
```

We see that as a result of the mineralization of about 60 $\mu mol~L^{-1}$ of organic matter over the period of 10 days, pH decreased from 7.9 to about about 7.73.

## Task 6: Application to ammonia degassing

### R implementation

The model reuses many elements from the model developed in Task 5. We reuse the functions *solveALK* and *solveH*. Compared to the model function *OMMpHmodel*, the present model function implements different mass balance equations (only NHx is removed, at a rate $\lambda\cdot[NH_3]$, while DIC and ALK are conserved), and does not use the state variable OM. The dissociation constants are the same as before.

```{r}
AmmoniaDegassing2 <- function(t, state, parms) {
  with (as.list(c(state, parms)), {

    # for output
    H <- solveH(K1 = K1, K2 = K2, Kn = Kn, DIC = DIC, NHx = NHx, ALK = ALK)
    HCO3    <- K1*H /  (H^2 + K1*H + K1*K2) * DIC   # eq 5a
    CO3     <- K1*K2 / (H^2 + K1*H + K1*K2) * DIC   # eq 5c
    NH4     <- H / (H + Kn) * NHx                   # eq 5e
    NH3     <- NHx - NH4

    # mass balance equations
    dDIC <-  0
    dNHx <- -lambda * NH3 # ammonia degassing, first-order kinetics NH3
    dALK <-  0

    return(list(c(dDIC, dNHx, dALK), 
                pH = -log10(H), H = H, 
                H2CO3 = DIC - HCO3 - CO3,
                HCO3 = HCO3,
                CO3  = CO3,
                NH4  = NH4,
                NH3  = NHx - NH4))
  })
}
```

We run the model for 20 days using the initial conditions given in the task. We also run the model with $[DIC]=0$ to check whether we get the same result as in Part I.

```{r, fig.height=8, fig.width=10}
require(deSolve)

pars <- c(lambda = 1,        # [1/day]
          K1 = 1.422e-06, # mol/L
          K2 = 1.082e-09, # mol/L
          Kn = 5.381e-10) # mol/L

pH.ini  <- 8.0                   # Hini based on pH
DIC.ini <- 2000*1e-6             # mol/L
NHx.ini <- 1000*1e-6             # mol/L
ALK.ini <- solveALK(DIC = DIC.ini, NHx = NHx.ini, pH = pH.ini)

times <- seq(from=0, to=20, length.out=100)

state.ini <- c(DIC=DIC.ini, NHx=NHx.ini, ALK = ALK.ini)
out       <- ode(y=state.ini, times=times, func=AmmoniaDegassing2, parms=pars)

# model without DIC, should give the same result as in Part I
ALK2.ini   <- solveALK(DIC = 0, NHx = NHx.ini, pH = pH.ini)
state2.ini <- c(DIC=0, NHx=NHx.ini, ALK = ALK2.ini)
out2   <- ode(y=state2.ini, times=times, func=AmmoniaDegassing2, parms=pars)

plot(out, out2, mfrow = c(3,4), las = 1, col=1:2, lty=1, lwd=2)
plot.new()
legend("top", legend = c("DIC=2000 umol/L", "DIC=0"), 
       col = 1:2, lwd = 2, lty = 1)

```

Without any DIC in the solution, the decrease in pH is the same as that obtained in Part I of the Reader (red line). When DIC is present, the protons released due to ammonia degassing are scavenged by $CO_3^{2-}$ in the solution, leading to an increase in $HCO_3^-$ and $H_2CO_3$. As a result, there is only a small increase in the proton concentration and a small decrease in pH. This illustrates the pH buffering effect by the carbonate system.

