---
title: "RTM course - reaction and transport in porous media"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on formulating reaction transport models in (saturated) porous media"
date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first.

### Video timeline
* 0:00 Introduction
* 1:34 Simplified reaction-transport equation
* 2:50 Porosity vs. solid volume fraction
* 4:35 Role of porosity on molecular diffusion
* 5:53 Role of porosity on measured & reported values of concentrations
* 7:07 Role of porosity on reaction-transport equation
* 8:29 Conversions between liquid and bulk
* 9:32 Reaction-transport equation reformulated for porous media
* 10:33 Conversions between solid and bulk
* 11:07 Reaction-transport equation in porous media - summary
* 12:14 Differential equations for multi-component models
* 13:02 Rate expressions for processes involving solutes and solids
* 15:30 Conclusion

![](https://www.youtube.com/watch?v=ANdrxhanosk&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=12)

You can also read the following section(s) of the textbook:

* Chapter 3.4.6. One-Dimensional Diffusion in Porous Media (Sediments)

(You can download each chapter of the textbook via the e-book University library for free!)

## General Reaction-Transport Equation

Consider the following (simplified) differential equation describing a one-dimensional reaction-transport model:

$$
\frac{\partial C}{\partial t} = \frac{\partial}{\partial x}\left(D\cdot \frac{ \partial C}{\partial x}\right) -\frac{\partial}{\partial x}(v \cdot C)  + R
$$

```{r GTE, echo=FALSE}
question("In the differential equation above, the term on the left hand side describes:", type = "learnr_checkbox",
 answer(sprintf("the substance concentration")),
 answer(sprintf("the substance concentration gradient (in space)")),
 answer(sprintf("the rate of change (in time) in the substance concentration"), correct = TRUE)
)  
```

```{r GTE2, echo=FALSE}
msg <- sprintf("Because $J=-D\\cdot (\\partial C/\\partial x)$ is the transport flux due to dispersion or diffusion, the first term on the right hand side describes the flux gradient due to dispersive or diffusive transport.")

question("In the differential equation above, the first term on the right hand side describes:", type = "learnr_checkbox",
 answer(sprintf("the substance concentration")),
 answer(sprintf("the substance concentration gradient")),
 answer(sprintf("the flux gradient due to dispersive transport"), correct = TRUE),
 answer(sprintf("the flux gradient due to advective transport")),
 answer(sprintf("the flux gradient due to diffusive transport"), correct = TRUE),
 
 correct=msg,
 incorrect=msg)
```

```{r GTE3, echo=FALSE}
msg <- sprintf("Because $J=v\\cdot C$ is the advective transport flux, the second term on the right hand side describes the flux gradient due to advective transport.")

question("In the differential equation above, the second term on the right hand side describes:", type = "learnr_checkbox",
 answer(sprintf("the substance concentration")),
 answer(sprintf("the substance concentration gradient")),
 answer(sprintf("the flux gradient due to dispersive transport")),
 answer(sprintf("the flux gradient due to advective transport"), correct = TRUE),

 correct=msg,
 incorrect=msg)
```

## Porosity

```{r por, echo=FALSE}
question("In water-saturated porous media such as sediments, porosity is defined as", type = "learnr_checkbox",
 answer("volume of liquid divided by volume of solid"),
 answer("volume of liquid divided by bulk volume", correct = TRUE),
 answer("volume of solid divided by volume of liquid"),
 answer("volume of solid divided by bulk volume")
)  
```

## Solid volume fraction

```{r svf, echo=FALSE}
question("In water-saturated porous media such as sediments, solid volume fraction is defined as", type = "learnr_checkbox",
 answer("volume of liquid divided by volume of solid"),
 answer("volume of liquid divided by bulk volume"),
 answer("volume of solid divided by volume of liquid"),
 answer("volume of solid divided by bulk volume", correct = TRUE),
 answer("one minus porosity", correct = TRUE)  
)  
```

## Effective diffusion coefficient

Complete the sentence to make a correct statement:

```{r effdif, echo=FALSE}
msg <- sprintf("The effective diffusion coefficient is defined as the molecular diffusion coefficient *divided* by tortuosity. Because tortuosity represents the path-length *prolongation*, it is always $>1$. Thus, the effective diffusion coefficient is always smaller than the molecular diffusion coefficient.")

question("In porous media, the effective diffusion coefficient of solutes is [. . .] than the molecular diffusion coefficient.", type = "learnr_radio",
 answer("smaller", correct=TRUE),
 answer("greater"),
  
 incorrect=msg,
 correct=msg)
```

## Reaction-transport in porous media I

```{r rteqpor, echo=FALSE}
msg <- sprintf("The mass balance equation is derived for the *bulk* concentration, $\\phi C$ (on the left hand side). However, the diffusive transport is driven by the concentration gradient in the *liquid* phase, hence $-D \\frac{ \\partial C}{\\partial x}$ is the diffusive flux in the liquid phase. To convert the diffusive flux from the liquid phase to the bulk phase, we need to multiply it with the porosity: $J=-\\phi D \\frac{ \\partial C}{\\partial x}$. Substituting this expression to the flux gradient term in the general mass balance equation yields the diffusive term in the form $\\frac{\\partial }{\\partial x}\\left(\\phi D \\frac{ \\partial C}{\\partial x}\\right)$. Because the reaction rate is expressed per volume of porewater, it also needs to be multiplied with porosity to express it per volume of the bulk.")
  
question(sprintf("Which of the following equations correctly describes the reaction-diffusion equation of a dissolved substance S with a porewater concentration $C$ (in $mol~S~m^{-3}_{L}$) in a water-saturated porous medium with porosity $\\phi$? The reaction rate $R$ is in the units of $mol~S~m^{-3}_{L}~s^{-1}$."), type = "learnr_checkbox",
 answer(sprintf("$\\frac{\\partial \\phi C}{\\partial t} =  \\frac{\\partial}{\\partial x}(\\phi D \\frac{ \\partial C}{\\partial x}) + \\phi R$"), correct=TRUE),
 answer(sprintf("$\\frac{\\partial \\phi C}{\\partial t} =  \\frac{\\partial}{\\partial x}(D \\frac{ \\partial (\\phi C)}{\\partial x}) + \\phi R$")),
 answer(sprintf("$\\frac{\\partial C}{\\partial t} =  \\frac{\\partial}{\\partial x}(\\phi D \\frac{ \\partial C}{\\partial x}) + \\phi R$")),
 answer(sprintf("$\\frac{\\partial \\phi C}{\\partial t} =  \\frac{\\partial}{\\partial x}(D \\frac{ \\partial C}{\\partial x}) + R$")),
 
 incorrect=msg,  
 correct=msg)  
```

## Reaction-transport in porous media II

```{r rteqpor2, echo=FALSE}
question(sprintf("The form of the reaction-diffusion equation in porous media is often written as $$\\frac{\\partial C}{\\partial t} = \\frac{1}{\\phi_x}  \\frac{\\partial}{\\partial x}\\left(\\phi_x D \\frac{ \\partial C}{\\partial x}\\right) + R,$$ where $C$ denotes the porewater concentration of a dissolved subtance S (in $mol~S~m^{-3}_{L}$). What does this formulation assume?"), type = "learnr_checkbox",
 answer("That the porosity is constant in time but variable in space", correct=TRUE),
 answer("That the porosity is constant in space AND in time"),
 answer("That the porosity is variable in space AND in time"),
 answer("That the reaction rate $R$ is expressed in the units of $mol~S~m^{-3}_{bulk}~s^{-1}$."),
 answer("That the reaction rate $R$ is expressed in the units of $mol~S~m^{-3}_{L}~s^{-1}$.", correct=TRUE)
)  
```

## True or False

```{r rteqpor3, echo=FALSE}
msg <- sprintf("To calculate the flux across the sediment-water interface in the units of mol per $m^2$ of *total* surface per time, we definitely need porosity: $J=-\\phi D(\\partial C/\\partial x)$.")

question(sprintf("When sediment porosity is constant in time AND in space, the following equation fully determines the diffusion model in a water-saturated sediment: $$\\frac{\\partial C}{\\partial t} = \\frac{\\partial}{\\partial x}\\left(D \\frac{ \\partial C}{\\partial x}\\right),$$ where $D$ is the effective (i.e., corrected for tortuosity) diffusion coeffient. Since there is no porosity in the equation, we do not need to know porosity at all."), type = "learnr_checkbox",
 answer("This is true as porosity does not play a role at all.", correct=FALSE),
 answer("This is not true if we want to quantify the flux across the sediment-water interface.", correct=TRUE),
 
 incorrect=msg,
 correct=msg)  
```

## Porosity conversions

```{r porconv1, echo = FALSE}
msg <- sprintf("The unit of $\\phi \\cdot C_L$ is $(m^3_L/m^3_b) \\cdot mol~m^{-3}_L$. This is equal to $mol~m^{-3}_b$, the units of $C_b$. Similarly for $(1-\\phi)\\cdot C_S$.")

question(sprintf("Select correct formulas for converting between concentrations per bulk volume ($C_b$), per volume of liquid ($C_L$), and per volume of solid ($C_S$). Porosity is denoted by $\\phi$."),
 answer(sprintf("$C_b = \\phi \\cdot C_L$"), correct=TRUE),
 answer(sprintf("$C_L = \\phi \\cdot C_b$")),
 answer(sprintf("$C_b = (1-\\phi) \\cdot C_L$")),
 answer(sprintf("$C_L = (1-\\phi) \\cdot C_b$")),
 answer(sprintf("$C_b = \\phi \\cdot C_S$")),
 answer(sprintf("$C_S = \\phi \\cdot C_b$")),
 answer(sprintf("$C_b = (1-\\phi) \\cdot C_S$"), correct=TRUE),
 answer(sprintf("$C_S = (1-\\phi) \\cdot C_b$")),

 incorrect=msg,
 correct=msg)
```

## Early diagenesis

Consider a two-component model in a porous media with porosity $\phi$. The two components are:

* dissolved $SO_4^{2-}$, which is a solute whose porewater concentration is in $mol~S~m^{-3}_L$ ($S$ refers to sulphur, $m^{-3}_L$ refers to the volume of the liquid phase).
* particulate organic matter, POM, which is a solid whose concentration is in $mol~C~m^{-3}_S$ ($C$ refers to carbon, $m^{-3}_S$ refers to the volume of the solid phase). 

The two components are involved in the process called "sulphate reduction", whose stoichiometry is as follows:

$$
POM + 0.5\cdot SO_4^{2-} \rightarrow CO_2 + 0.5\cdot S^{2-} + \cdots
$$
The rate of sulphate reduction, $R_{sr}$ is expressed in $mol~C~m^{-3}_S~d^{-1}$.

```{r porconv2, echo = FALSE}
msg <- sprintf("First, you need to convert from $m^{-3}_S$ to $m^{-3}_L$, which is done by the factor $(1-\\phi)/\\phi$. Additionally, you need to convert from $mol~C$ to $mol~S$, where you need to consider that only 0.5 moles of $SO_4^{-2}$ are consumed per mole of $C$ respired.")

question(sprintf("Which is the correct expression for the rate of $SO_4^{2-}$ removal by sulphate reduction?"),
 answer(sprintf("$R_{SO_4^{2-}} = R_{sr} \\times (1-\\phi) / \\phi$")),
 answer(sprintf("$R_{SO_4^{2-}} = 0.5 \\times R_{sr} \\times (1-\\phi) / \\phi$"), correct=TRUE),
 answer(sprintf("$R_{SO_4^{2-}} = 2 \\times R_{sr} \\times (1-\\phi) / \\phi$"), correct=FALSE),
 answer(sprintf("$R_{SO_4^{2-}} = R_{sr} \\times \\phi/(1-\\phi)$"), correct=FALSE),
 answer(sprintf("$R_{SO_4^{2-}} = 0.5 \\times R_{sr} \\times \\phi/(1-\\phi)$"), correct=FALSE),
 answer(sprintf("$R_{SO_4^{2-}} = 2 \\times R_{sr} \\times \\phi/(1-\\phi)$"), correct=FALSE),
 answer(sprintf("$R_{SO_4^{2-}} = R_{sr}$")),
 answer(sprintf("$R_{SO_4^{2-}} = 0.5 \\times R_{sr}$")),
 answer(sprintf("$R_{SO_4^{2-}} = 2 \\times R_{sr}$")),
 
 incorrect=msg,
 correct=msg)
```

## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

## References

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.
