---
title: "A Bioeconomic model of Scallop Cultures"
subtitle: "Exercises Accompanying the Course Environmental Modelling"
author: "Karline Soetaert and Dries Bonte, Ghent University"
date: "2021"
output: 
  pdf_document:
    toc: no 
  header-includes: 
    \usepackage{float} 
    \floatplacement{figure}{H} 
---

# Answers

The equations describing the ecological part of this problem are given:
$$\frac{dL}{dt}=k\cdot (L_\infty-L)$$
$$\frac{dN}{dt}=-z \cdot N$$
where the density of the scallops continuously declines due to mortality. The initial conditions are:
$L_{(t=0)} = 20~mm$ and $N_{(t=0)} = 5\times 10^4$.


For the economical part, we take the *cost* as a state variable that increases with the daily cost of culturing:
$$\frac{dCost}{dt} = Culturecost$$
We assume $Cost_{(t=0)}=0$.

At each time step, we can estimate the price the manager would get if she sold the Scallops. Substracting the current costs from this revenue then gives the current profit or loss that would be obtained if they were sold.

To test whether it is worthwhile to invest in new trays, the cost is augmented, at the beginning with the investment cost, i.e., the initial condition is changed $Cost_{(t=0)}=InvestCost$. Also the initial condition for the density is doubled: $N_{(t=0)} = 10\times 10^4$.

# R implementation

## Model implementation

```{r}

require(deSolve)  # package with solution methods

# state variables
state <- c(L = 20,        # Scallop length, [mm]
           N = 5*10^4,    # Scallop number  [ind]   
           Cost = 0)      # Integrated cost [dollars]

# parameters
parms <- c(
  k    = 0.8,            # growth rate constant       [/yr] 
  Linf = 100,            # asymptotic length          [mm]
  z    = 0.5,            # mortality rate constant    [/yr]
  a    = 8E-6, 
  b    = 3.4,            # allometric coefficients to estimate weight
  numtrays    = 100,     # number of trays
  minWeight   = 30,      # Minimal weight at which Scallops can be sold
  culturecost = 150,     # cost for culturing          [/yr]
  price_kg    = 1.43,    # price per/KG 
  price_a     = 0.1      # increased price per gram over minW
)

#The derivative function
Scallops <- function(t, state, params){
  with (as.list(c(state, params)), {
  
    dL    <- k*(Linf-L)
    dN    <- -z*N
    dCost <- culturecost          # daily cost for culturing adds to total

    weight <- a * L^b             # [g ind-1] 
    biom   <- weight*N/1000       # biomass, [kg]
    price  <- (price_kg + price_a*(weight-minWeight))*(weight > minWeight)
    profit <- (biom*price - Cost)* numtrays # total profit = price minus integrated cost
    
    return (list(c(dL, dN, dCost),
                 weight = weight,
                 biomass = biom,
                 price = price,
                 profit = profit))
  })
}

``` 

## Model runs

```{r, fig.height = 10, fig.width = 8}

# output times           
outtimes <- seq(from = 0, to = 3, length.out = 100)

#solve the equation
out <- ode(y = state, parms = parms, func = Scallops, times = outtimes)
head(out)
max(out[,"profit"] )

# ---------------------------------------------------
# second run - a warmer climate, but more expensive
# ---------------------------------------------------
pars <- parms
pars["k"] <- 1
pars["culturecost"] <- pars["culturecost"]*2
out2 <- ode(y = state, parms = pars, func = Scallops, times = outtimes)
plot(out, out2, xlab = "year")
max(out[,"profit"]) 

# ... and without the extra profit for larger scallops
p2 <- parms
p2["price_a"] <- 0
outb <- ode(y = state, parms = p2, func = Scallops, times = outtimes)
plot(out, outb, xlab = "year")


# ---------------------------------------------------
# third run - increase the size of the farm? 
# ---------------------------------------------------
# first year
stateb <- state
stateb["Cost"] <- 250   # initial investment cost

parsb <- parms
parsb["culturecost"] <- parms["culturecost"]*1.2   # maintenance is more expensive
parsb["numtrays"] <- 2*parms["numtrays"]   # number of trays * 2

out3 <- ode(y = stateb, parms = parsb, func = Scallops, times = outtimes)

# second year: no investment cost 
statec <- stateb; statec["Cost"] <- 0
out4 <- ode(y = statec, parms = parsb, func = Scallops, times = outtimes)
```

```{r, fig.height = 5, fig.width = 10}
par(mfrow = c(1, 2))
plot(out, out2, which = "profit", lwd = 2, xlab = "year", ylab = "dollar", mfrow = NULL)
legend("bottomright", col = 1:2, lwd = 2, lty = 1:2, c("original", "warmer"))


plot(out, out3, out4, which = "profit", lwd = 2, xlab = "year", ylab = "dollar", 
     mfrow = NULL)
legend("topleft", col = 1:3, lwd = 2, lty = 1:3, 
       c("original", "doubling, 1st year", "doubling, 2nd year"))

```
