---
title: "RTM course - Enzymatic Rate Laws"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on enzymatic reactions (mainly Michaelis-Menten)"
date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first.

### Video timeline
* 0:00 Michaelis-Menten kinetics
* 5:50 Application to organic matter mineralization

![](https://www.youtube.com/watch?v=z2p_qPvRuTg&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=6)

You can also read the following section(s) of the textbook:

* Chapter 2.4. Enzymatic reactions
* Chapter 2.5.3. One Rate-Limiting Resource, 3 Types of Functional Responses
* Chapter 2.5.4. More than One Limiting Resource

(You can download each chapter of the textbook via the e-book University library for free!)

## Enzymatic rate

```{r enz_rate, echo=FALSE}
question("Which of the following expressions describes the regulation of an enzymatic process by the concentration of the rate-limiting substrate A?", type = "learnr_radio",
 answer(sprintf("$\\frac{[A]}{[A]-K_A}$"), correct=FALSE),
 answer(sprintf("$\\frac{[A]}{[A]+K_A}$"), correct=TRUE),
 answer(sprintf("$\\frac{K_A}{[A]-K_A}$"), correct=FALSE),
 answer(sprintf("$\\frac{K_A}{[A]+K_A}$"), correct=FALSE)
)
```

## Type of reactions

```{r enz_truefalse, echo=FALSE}
question("Reactions whose rate is described by Michaelis-Menten kinetics are", type = "learnr_radio",
 answer("reversible", correct=FALSE),
 answer("irreversible", correct=TRUE)
)
```

## Enzymatic parameter

```{r enzPar, echo=FALSE}

question(sprintf("What is the name of the parameter $K_A$ in the Michaelis-Menten rate expression $\\frac{[A]}{[A]+K_A}$?"), type = "learnr_checkbox",
 answer("Enzyme affinity to the rate-limiting substrate A", correct = TRUE),
 answer("Equilibrium concentration", correct = FALSE),
 answer("Half-saturation concentration", correct = TRUE),
 answer("Rate constant", correct = FALSE),
 answer("Solubility concentration"),
 answer("Affinity concentration",  correct = TRUE)
)
```

## Michaelis-Menten shape

```{r enzShape, echo=FALSE}
question(sprintf("What is the shape of the curve in the Michaelis-Menten rate expression $\\frac{[A]}{[A]+K_A}$?"), type = "learnr_radio",
 answer("Hyperbola",  correct = TRUE),
 answer("Parabola"),
 answer("Exponential")
)
```

## Michaelis-Menten use

```{r enzUse, echo=FALSE}
question("Michaelis-Menten kinetics is approximated by first-order kinetics when the substrate concentration is", type = "learnr_radio",
 answer("around the affinity concentration"),
 answer("much lower than the affinity concentration", correct = TRUE),
 answer("much greater than the affinity concentration")
)
```

```{r enzUse2, echo=FALSE}
question("Michaelis-Menten kinetics is approximated by zero-order kinetics when the substrate concentration is", type = "learnr_radio",
 answer("around the affinity concentration"),
 answer("much lower than the affinity concentration"),
 answer("much greater than the affinity concentration", correct = TRUE)
)
```

## Recognizing rate-limitation

```{r enzRateLim, echo=FALSE}
question(sprintf("How would you characterize a process whose rate is described by the following rate expression (where $r$, $K_A$ and $K_B$ are constants): $R = r \\times \\frac{[A]}{[A]+K_A} \\times \\frac{K_B}{[B] + K_B} \\times [C]$?"), type = "learnr_checkbox",
 answer("The process is stimulated by substrate A",  correct = TRUE),
 answer("The process is inhibited by substrate A"),
 answer("The process is stimulated by substrate B"),
 answer("The process is inhibited by substrate B",  correct = TRUE),
 answer("The process is first-order with respect to substrate C",  correct = TRUE),
 answer("The process is zero-order with respect to substrate C")
)
```

## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

## References

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.
