---
title: "RTM course - Rate Laws for Enzymatic Reactions"
author: "Karline Soetaert and Lubos Polerecky"
description: "video and exercises on enzymatic reactions (mainly Michaelis-Menten)"
date: "June 2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first.

### Video timeline
* 0:00 Michaelis-Menten kinetics
* 5:50 Application to organic matter mineralization

![](https://www.youtube.com/watch?v=z2p_qPvRuTg&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=6)

Slides available at <a href="https://drive.google.com/drive/folders/1Bys0-xzXLCpFpWhD-HrH6jDKOyl8iJMd" target="_blank">google-drive</a>

You can also read the following section(s) of the textbook:

* Chapter 2.4. Enzymatic reactions
* Chapter 2.5.3. One Rate-Limiting Resource, 3 Types of Functional Responses
* Chapter 2.5.4. More than One Limiting Resource

(You can download each chapter of the textbook via the e-book University library for free!)

## Enzymatic rate

```{r enz_rate, echo=FALSE}
question("Which of the following expressions describes the regulation of an enzymatic process by the concentration of the rate-limiting substrate A?", type = "learnr_radio",
 answer(sprintf("$\\frac{[A]}{[A]-K_A}$"), correct=FALSE),
 answer(sprintf("$\\frac{[A]}{[A]+K_A}$"), correct=TRUE),
 answer(sprintf("$\\frac{K_A}{[A]-K_A}$"), correct=FALSE),
 answer(sprintf("$\\frac{K_A}{[A]+K_A}$"), correct=FALSE)
)
```

## Type of reactions

```{r enz_truefalse, echo=FALSE}
question("Reactions whose rate is described by Michaelis-Menten kinetics are", type = "learnr_radio",
 answer("reversible", correct=FALSE),
 answer("irreversible", correct=TRUE)
)
```

## Enzymatic parameter

```{r enzPar, echo=FALSE}

question(sprintf("What is the name of the parameter $K_A$ in the Michaelis-Menten rate expression $\\frac{[A]}{[A]+K_A}$?"), type = "learnr_checkbox",
 answer("Enzyme affinity to the rate-limiting substrate A", correct = TRUE),
 answer("Equilibrium concentration", correct = FALSE),
 answer("Half-saturation concentration", correct = TRUE),
 answer("Rate constant", correct = FALSE),
 answer("Solubility concentration"),
 answer("Affinity concentration",  correct = TRUE)
)
```

## Michaelis-Menten shape

```{r enzShape, echo=FALSE}
question(sprintf("What is the shape of the curve in the Michaelis-Menten rate expression $\\frac{[A]}{[A]+K_A}$?"), type = "learnr_radio",
 answer("Square-root"),
 answer("1 minus Exponential"),
 answer("Hyperbola",  correct = TRUE),
 answer("Parabola"),
 answer("Exponential"),
 answer("Arc-tan")
)
```

```{r, echo=FALSE, fig.width=6, fig.height=4}
A <- 0:100
par(mfrow=c(2,3))
plot(x=A,y=sqrt(A)/10, main="Square-root", ylab="", xlab="[A]", type="l", lwd=2)
plot(x=A,y=1-exp(-A/30), main="1 minus exponential", ylab="", xlab="[A]", type="l", lwd=2)
plot(x=A,y=A/(A+10), main="Hyperbola", ylab="", xlab="[A]", type="l", lwd=2)
plot(x=A,y=1-(A-50)^2/2500, main="Parabola", ylab="", xlab="[A]", type="l", lwd=2)
plot(x=A,y=0.001^(-A/200), main="exponential", ylab="", xlab="[A]", type="l", lwd=2)
plot(x=A,y=atan(A/20)/(pi/2), main="Arc-tan", ylab="", xlab="[A]", type="l", lwd=2)
```

## Michaelis-Menten use

```{r enzUse, echo=FALSE}
msg <- "When [A] is very small ($[A]\\ll K_A$), the Michaelis-Menten rate expression can be approximated as $\\frac{[A]}{[A]+K_A}\\approx \\frac{[A]}{K_A} = \\frac{1}{K_A}\\cdot [A] \\equiv r\\cdot [A]$, which is first-order in [A]."
question("Michaelis-Menten kinetics is approximated by first-order kinetics when the substrate concentration is", type = "learnr_radio",
 answer("around the affinity concentration"),
 answer("much lower than the affinity concentration", correct = TRUE),
 answer("much greater than the affinity concentration"),
 incorrect = paste("Incorrect. ", msg),
 correct = paste("Correct. ", msg)
 
)
```

```{r enzUse2, echo=FALSE}
msg <- "When [A] is very large ($[A]\\gg K_A$), the Michaelis-Menten rate expression can be approximated as $\\frac{[A]}{[A]+K_A} \\approx \\frac{[A]}{[A]} \\equiv 1$, which is zero-order in [A]."

question("Michaelis-Menten kinetics is approximated by zero-order kinetics when the substrate concentration is", type = "learnr_radio",
 answer("around the affinity concentration"),
 answer("much lower than the affinity concentration"),
 answer("much greater than the affinity concentration", correct = TRUE),
 incorrect = paste("Incorrect. ", msg),
 correct = paste("Correct. ", msg)
)
```

```{r, echo=FALSE, fig.width=8, fig.height=4}
A <- 0:100
A2 <- 0:300
ks <- 10
par(mfrow=c(1,3))
plot(x=A,y=A/(A+ks), main="Michaelis-Menten (KA=10)", ylab="", xlab="[A]", type="l", lwd=2)
plot(x=A,y=A/(A+ks), main="first-order approximation", ylab="", xlab="[A]", type="l", lwd=2, lty=1, 
     xlim=c(0,40), ylim=c(0,1))
lines(x=A,y=A/ks, ylab="", xlab="[A]", type="l", lwd=2, lty=2, col=2)
plot(x=A2,y=A2/(A2+ks), main="zero-order approximation", ylab="", xlab="[A]", type="l", lwd=2, lty=1,
     ylim=c(0,1))
lines(x=A2,y=rep(1, times=301), ylab="", xlab="[A]", type="l", 
      lwd=2, lty=2, col=2)
```

## Recognizing rate-limitation

```{r enzRateLim, echo=FALSE}
msg <- sprintf("The dependency of the process on A is given by the Michaelis-Menten kinetics $\\frac{[A]}{[A]+K_A}$; the dependency on B is given by $\\frac{K_B}{[B] + K_B}$ which is equal to $1-\\frac{[B]}{[B] + K_B}$, and this function will decrease with increasing $[B]$ (inhibition); the dependency on C is given by $r \\times [C]$, which is first-order with respect to [C].")

question(sprintf("How would you characterize a process whose rate is described by the following rate expression (where $r$, $K_A$ and $K_B$ are constants): $R = r \\times \\frac{[A]}{[A]+K_A} \\times \\frac{K_B}{[B] + K_B} \\times [C]$?"), type = "learnr_checkbox",
 answer("The process is stimulated by substrate A",  correct = TRUE),
 answer("The process is inhibited by substrate A"),
 answer("The process is stimulated by substrate B"),
 answer("The process is inhibited by substrate B",  correct = TRUE),
 answer("The process is first-order with respect to substrate C",  correct = TRUE),
 answer("The process is zero-order with respect to substrate C"),
 incorrect = paste("Incorrect. ", msg),
 correct = paste("Correct. ", msg)
 )
```

## Textbook

Karline Soetaert, Peter M.J. Herman (2009). A Practical Guide to Ecological Modelling. Springer Netherlands. DOI: 10.1007/978-1-4020-8624-3.

